\documentclass[8pt,letterpaper,extrafontsizes]{memoir}

% change paper size to novel.
\usepackage[a5paper,twoside]{geometry}

% enable graphics, needed to display the pictures.
\usepackage[pdftex]{graphicx}

% enable appendix
\usepackage[toc,page]{appendix}

% print pages with header, footer and page number.
\pagestyle{headings}

% supplies in-line enumerations.
\usepackage{paralist}

% increase the size of the mini page width for epigraphs.
% \setlength{\epigraphwidth}{.5\textwidth}

% for the 20th, 2nd etc.
\usepackage[super]{nth}

% \setlength{\parskip}{0.2ex plus 0.0ex minus 0.0ex}
% \setlength\parindent{0pt}

\usepackage{lettrine}

% Use the built-in division styling
\headstyles{memman}

% ToC down to subsections
\settocdepth{subsection}

% Numbering down to subsections as well
\setsecnumdepth{subsection}

% extra index for first lines
\makeindex[lines]

\begin{document}
\tightlists
\midsloppy
\raggedbottom
\chapterstyle{demo3}

% \frontmatter
\pagestyle{empty}
\vspace*{\fill}
\begin{center}
\HUGE\textsf{The Right To Development}\par
\end{center}

\vspace*{\fill}
\begin{center}
\Large\textsf{Shahinara Begum}\par
\end{center}

\clearpage
\pagestyle{headings}
\tableofcontents

\chapter*{Table of Cases}
\addcontentsline{toc}{chapter}{\protect\numberline{}Table of Cases}

\begin{itemize}
\item International Court of Justice
\item Anglo-Norwegian Fisheries Case: United Kingdom v Norway (1951) ICJ Rep 116.
\item Asylum Case: Columbia v Peru (1950) ICJ Rep 266
\item Legality of the Threat or Use of Nuclear Weapons (Request for
  Advisory Opinion by the World Health Organisation) (1996) ICJ Rep 90
\item Military and Paramilitary Activities in and Against Nicaragua:
  Nicaragua v United States of America, (Merits), (1986) ICJ Rep 14
\item North Sea Continental  Shelf Cases: Federal Republic of Germany
  v Denmark; Federal Republic of Germany v Netherlands (1969) ICJ Rep
  3.
\item African Commission on Human and Peoples' Rights
\item Malawi African Association etc v Mauritania (2000) AHRLR 149
  (ACHPR 2000).
\item Social and Economic Rights Action Center and the Center for
  Economic and Social Rights v Nigeria Communication 155/96, 27
  October 2001.

\end{itemize}

\chapter*{Table of Treaties}
\addcontentsline{toc}{chapter}{\protect\numberline{}Table of Treaties}

\begin{itemize}
\item Including Covenants, Treaties, Conventions, ICJ Statute and UN
  Charter etc.
\item Additional Protocol to the American Convention on Human Rights
  in the Area of Economic, social and cultural rights Protocol of San
  Salvador 1988
\item Additional Protocol to the European Social Charter Providing for
  a System of Collective Complaints 1995
\item African Charter on Human and People's Rights 1981
\item African Charter on the Rights and Welfare of the Child 1990
\item American Convention on Human Rights 1969
\item American Declaration of Human Rights and Duties of  Man 1948
\item Arab Charter on Human Rights 1994
\item Charter of the Organisation of African Unity 1963
\item Convention on the Elimination of All Forms of Discrimination
  Against Women 1979
\item Convention on the Elimination of All Forms of Racial
  Discrimination 1965
\item Convention on the Rights of the Child 1989
\item Draft Protocol to the African Charter on Human and Peoples'
  Rights on the Rights of Women
\item European Convention for the Protection of Fundamental Human
  Rights and Fundamental Freedoms 1950
\item European Social Charter 1961
\item European Social Charter 1961
\item Inter-American Convention on Human Rights 1969
\item International Convention on the Protection of the Rights of All
  Migrant Workers and Members of their Families 1990
\item International Covenant on Civil and Political Rights 1966
\item International Covenant on Economic, Social and Cultural Rights
  1966
\item Revised European Social Charter 1996
\item Statute of the International Court of Justice 1945
\item Statute of the Permanent Court of International Justice 1920
\item United Nations Charter 1945

\end{itemize}

\chapter*{Table of Other Documents}
\addcontentsline{toc}{chapter}{\protect\numberline{}Table of Other Documents}

\section*{Declarations, Resolutions etc.}
\addcontentsline{toc}{section}{\protect\numberline{}Declarations,
  Resolutions etc.}

\begin{itemize}
\item International Labour Organisation Philadelphia Declaration 1944
\item General Assembly of the United Nations Resolutions and and
  Declarations.
\item Charter of Economic Rights and Duties of States 1974, Resolution
  3281 (XXIX).
\item Declaration on Principles of International Law Concerning
  Friendly Relations and Co-operation among States in Accordance with
  the Charter of the United Nations 1970, Resolution 2625 (XXV).
\item Declaration on Social Progress and Development 1969, Resolution
  2542 (XXIV)
\item Declaration on the Permanent Sovereignty over Natural Resources
  1962, Resolution 1803 (XVII)
\item Declaration on the Preparation of Societies for Life in Peace
  1978, Resolution 33/73 (1978)
\item Declaration on the Right to Development 1986, Resolution 41/128
  (1986).
\item Definition of Aggression 1974, Resolution 3314 (XXIX).
\item General Assembly Resolutions on the Right to development:
\item Resolution 50/184 (1996) Right to development
\item Resolution 51/99 (1997) Right to development
\item Resolution 52/136 (1998) Right to development
\item Resolution 53/155 (1999) Right to development
\item Resolution 54/175 (2000) The right to development
\item Resolution 55/108 (2001) The right to development
\item Resolution 56/150 (2002) The right to development.
\item Millennium Declaration 2000, Resolution 55/2 (2000)
\item Universal Declaration of Human Rights 1948, Resolution 217A (III).
\end{itemize}

\section*{International Conferences}
\addcontentsline{toc}{section}{\protect\numberline{}International Conferences}

\begin{itemize}
\item International Conference on Population and Development, Cairo 1994
\item International Conference on Human Rights, Tehran 1969
\item Second UN Conference on Human Settlements (Habitat II)1996
\item The Platform for Action 1995 Fourth World Conference on Women
\item The World Food Summit 1996
\item World Conference against Racism, Racial Discrimination,
  Xenophobia and Related Intolerance 2001
\item World Conference on Human Rights, Vienna, Vienna Declaration and
  Programme of Action 1993
\item World Summit for Social Development, Copenhagen 1995
\item World Summit on Sustainable Development 2002
\end{itemize}

\chapter*{Introduction}
\addcontentsline{toc}{chapter}{\protect\numberline{}Introduction}

\lettrine{T}{raditionally}, international law regulated the relations
between states, based on the principle of sovereign equality of each
state and consent. The \nth{20} century saw the birth of international
human rights law. In contrast to international law, international
human rights law regulates state behaviour toward its citizens and the
citizens of other states on their territory. International human
rights law is also based on the principle of state sovereignty and
consent and, as such, each state determines to what extent it wishes
to be bound by it. Under the principles of non-intervention and state
sovereignty, states do not have to ensure the protection of human
rights in any other territory but their own.

Whilst international human rights law emerged, in a parallel setting
international economic law also materialised as a separate branch of
international law, dealing with development issues. A states
development was measured by its economic growth in terms of its Gross
National Product (GNP) or Gross Domestic Product (GDP) and focused on
policies such as 'basic needs' and 'structural adjustment'. This
branch of law predominately dealt with issues regarding poverty,
primarily seen as an economic concern. However, it became apparent
that the definition of development within this branch of law was
essentially concerned with development as economic growth and material
wealth and omitted concerns for human rights. With the need to
recognise fundamental human rights such as the rights to food,
housing, basic health care and education, as concerns which affect the
development discourse, the human right to development (TRTD) emerged,
forging a link between the human rights discourse and the development
discourse.

The concept of the right to development is fairly new within
international law. It was endorsed as a human right by the General
Assembly Declaration on the Right to Development (TDRTD) by a
majority, with only the USA casting a dissenting vote. The link
between human rights and economic and social development was
recognised in the United Nations Charter. The Universal
Declaration of Human Rights, declared that human rights
encompassed both civil and political rights (CPR's) and Economic,
social and cultural rights (ESCR's). During the cold war,
disagreement occurred regarding the unity of such rights, which
resulted in the adoption of two separate covenants, the International
Covenant on Civil and Political Rights (ICCPR) and the International
Covenant on Economic, Social and Cultural rights (ICESCR). This
created a hierarchy between rights which was not envisaged by the
Universal Declaration and as such, CPR's were given priority over
ESCR's, by many states. Although neither covenant specifically
address TRTD, TDRTD re-integrates CPR's and ESCR's as indivisible and
interdependent.

This paper focuses upon the TRTD as an emerging concept in
international law and attempts to determine if there is a legal
obligation for the international community to realise the right. TRTD
has been framed as a human right on numerous occasions. This suggests
that it falls within international human rights law. I will first
clarify the status of TRTD within the human rights framework. This is
of vital significance as the international community does not have a
legal obligation to realise TRTD as a human right -- this is the duty
of individual states. TRTD is then analysed in the context of
international law, which serves to clarify the rules that denote legal
obligations on states.

This research uses several methodologies. An historical methodology is
used in numerous sections to identify the background of the issues
which are discussed and to place TRTD into context. Stemming from
this, a descriptive methodology is used to provide information on the
key issues, this serves to set the scene for a critical analysis. In
many of the sections, correlational and explanatory methodologies are
employed to identify, discuss and explain the relationships between
TRTD and the issues which have been identified.

TRTD remains contentious and as such, this research serves to provide
a synthesis on the current debate of the subject. While undertaking
this investigation, it was discovered that, although there are primary
materials specifically addressing TRTD, there is a limited range of
secondary and tertiary material. Journal articles are scarce to find
and appear sporadically, with only a few academic scholars commenting
on the right. In addition, many of the articles have been written
specifically by United Nations experts and other governmental
organisations; there is very little Non-Governmental Organisations
(NGO) feedback on the RTD and thus there is a risk of bias in this
research. Furthermore, this chosen area of study can encompass a wide
variety of issues and can become too dense and complicated. It is for
this reason that this paper excludes any in depth analysis on economic
policies, economic law and international environmental law, making it
more suitable both in terms of the time given to undertake it and also
the strict word limit.

Chapter 1 lays the foundations and introduces the emergence of
TRTD. It provides a brief historical account of international human
rights law and identifies the implicit references to TRTD by the
international community, to place TRTD into the framework of human
rights and international law as a whole.  Chapter 1 also introduces
the political factors which affect TRTD and identifies the explicit
references to TRTD made by the international community.

Chapter 2, summarises the non-controversial aspects of TRTD, which
have been agreed by the international community. It then focuses on
the controversies surrounding TRTD in order to establish whether or
not these can clarify the nature of the right, its relationship with
different aspects of human rights and where TRTD belongs in
international law.

Chapter 3 examines the legal status of TRTD in international treaties
and in customary international law, in order to establish whether or
not there is a legal obligation for the international community to
realise TRTD. This chapter provides information on the background of
specific legal and political instruments.

\chapter{Introducing the Right to Development}

\lettrine{T}{his} chapter provides a brief historical account of
international human rights law and identifies the implicit references
made to the right. It provides an insight into the political factors
surrounding the emergence of international human rights law. This
chapter then focuses on the background of TRTD which essentially deals
with the explicit references to the right and the political backdrop
in which it emerged.

\section{The Emergence of the Right to Development}

The end of the First World War saw the creation of the League of
Nations with its subsidiary organs, the International Labour
Organisation (ILO) and the Permanent Court of International Justice
(PCIJ). The ILO was essentially created to improve people's working
conditions and promote freedom of association. It was the first
international organisation to address social inequality and social
justice, a fundamental maneuver for the recognition of social rights
as human rights.

The 1944 ILO Philadelphia Declaration contained implicit references to
the TRTD, affirming that every individual has ``the right to pursue
both their material well-being and their spiritual development in
conditions of freedom and dignity, of economic security and equal
opportunity.''

As a result of the atrocities witnessed during the Second World War,
the distinct branch of international human rights law evolved
significantly within international public law. The abuses perpetrated
by Nazi Germany stimulated the international community to create a
framework in which fundamental rights could be protected without
infringing state sovereignty and thus emanating the foundations of
what is now modern international human rights law. Although the notion
of human rights developed from \nth{19} century western thought and
philosophy, the foundations of social justice, liberty and freedom can
be traced back over centuries, in many different parts of the world;
numerous religions; cultures; philosophies, to name but a few.

The United Nations (UN) was established in the aftermath of the Second
World War. Created to maintain international peace and security
through cooperation, the UN charter underlined the necessity to
promote human rights. The UN declared one of its main aims was ``to
reaffirm faith in fundamental human rights, in the dignity and worth
of the human person, in the equal rights of men and women and of
nations large and small''. Article 55 crucially links international
stability alongside social and economic development and human rights,
suggesting their interdependence. Article 56 pledges all members to
take both joint and individual action to achieve the purposes set
forth in Article 55. Furthermore, Article 68 stipulated that the
Economic and Social Council (ECOSOC) shall set up a Commission on
Human Rights (CHR). These articles could be used to indirectly affirm
the existence of TRTD.

The General Assembly of the United Nations (GA) adopted the Universal
Declaration of Human Rights (UDHR) in 1948. The UDHR was hailed as a
``a declaration of the basic principles to serve as a common standard
for all nations''. The Declaration begins with declaring that ``all
human beings are born free and equal in dignity and rights'' and that
everyone is entitled to the rights set forth in the
declaration. Articles 3-20 concern the variety of CPR's contained in
the UDHR while ESCR's were set forth in Articles 23-28.

The legal status of the UDHR has been questioned as the Declaration
was not intended to be a binding legal document instead, it was
envisaged at the time that the Conventions which were to follow the
UDHR, would be of a binding legal character. Although the UDHR was not
legally binding, the impact it had on the international law of human
rights was phenomenal. The Declaration was adopted into the written
constitutions of many states, it was also referred to in a number of
other instruments including, other specific human rights agreements,
regional agreements created to protect human rights and GA
resolutions.  Many commentators have suggested that the principles
contained in the UDHR have become customary international law and has
been raised to the status of \textit{sui generis}.

The importance of the UDHR cannot be underestimated. It has provided
the world with a common standard of rights afforded to all people,
taking into account the diversity of opinions and thoughts throughout
the globe. It placed international human rights law firmly on the
agenda, alongside the specific rights and principles which TRTD would
later encompass.

The GA indirectly made reference to the contents of what would later
become TRTD. In 1957 it issued a resolution which endorseded the
notion that integrated economic and social development contributed to
the promotion of peace, security, social progress, better standards of
living throughout the world and respect for human rights. A crucial
moment to the integration of development objectives and human rights
occurred when a link between the two was identified in a UN report.

The GA declared the 1960's as the \textit{United Nations Development
  Decade }and in 1965 adopted a resolution which recognised the need
to promote human rights within the context of the development
decade. These recognitions made by the GA are of significant
importance as they lay the foundations for TRTD.

The different ideologies of States were enhanced during the cold war
and evidenced in discussions held by the CHR regarding the content of
a covenant, which would have legally binding obligations and
monitoring mechanisms for the human rights already established in the
UDHR. During the discussions it became apparent that the ideologies of
different nations would compromise the outcomes for the human rights
discourse. Many of the western delegates argued that the inclusion of
economic and social rights in one single legal covenant would lead to
many countries restricting CPR's in the pursuit of economic
stability. Whilst on the other side, delegates argued that a covenant
only addressing CPR's completely disregarded economic exploitation and
extreme poverty. Even the juridical nature of ESCR's was questioned,
leading to arguments regarding justiciability and enforcement. Some
delegates believed that ESCR's were not justiciable and were not
capable of immediate implementation as they were a set of 'positive'
obligations, requiring the state to 'to do something' in order to
realise such rights. Whereas CPR's were seen as 'negative'
obligations, only requiring the state to abstain from 'doing
something' to realise CPR's. To solve such differences the original
idea of one covenant was divided into two, one concerning CPR's and
the other on ESCR's, a division not envisaged by the UDHR as all
rights were seen as indivisible and interdependent. Both the ICCPR and
ICESCR are afforded the status of an international treaty and thus all
states who ratify the covenants are legally bound by their obligations
under international law.

The ICCPR expands on the rights provided for in the UDHR and provides
mechanisms for enforcement. Article 28 established the Human Rights
Committee, the body responsible for the enforcement of the
covenant. Furthermore, Article 40 requires states to submit a report
after one year of ratification and then every five years, regarding
the progress made on realising the CPR's contained in the
covenant. Article 1 of the first Optional Protocol 1966 provided
individual victims with measures to seek redress for violations of
CPR's from the HRC. Moreover, the HRC can also seek information from
other sources including Non- Governmental Organisations (NGO's).

The ICESCR provides numerous provisions for the advancement of ESCR's
broadening the scope of the UDHR. Articles 16 and 17 provide
obligations for states to submit reports which will be considered by
ECOSOC, reports can be obtained by other specialised agencies and can
also be examined by the Commission on Human Rights.  However, the
ICESCR has been severely criticised for a number of reasons including
criticisms regarding the 'vague' language used throughout the
document; implementation differs from one covenant to the other thus
seeming to afford more protection to one set of rights over another in
contradiction with the aims of the human rights discourse; state
obligations differ in the covenants; progressive implementation of
provisions makes it difficult to determine who has fulfilled their
obligations and the covenants provisions are too 'aspirational' and
can never be achieved in full . Nevertheless, the ICESCR provides the
only framework in the international arena in which ESCR's are declared
and protected human rights. Furthermore it provides indirect
references to the contents of TRTD.

The consequences of dividing CPR's and ESCR's was seen throughout the
subsequent decades and in other human rights instruments. Three
generations of rights emerged. First generation rights consisted of
CPR's, 'freedom' rights; second generation rights were ESR's,
'equality' rights and a further generation emerged as third generation
rights, rights of 'solidarity' encompassing both CPR's and ESCR's,
which included amongst others the right to a healthy environment, the
right to peace and the TRTD.

\subsection{Towards Integrating Development and Human Rights}

Many factors were accountable to the increase of attention to rights
within the development discourse. The end of both the world wars gave
rise to an increase of States, either newly established or the result
of decolonisation, who's primary concerns included economic
development in the international sphere. Developing countries
contended that it was necessary to afford more attention to economic
and social rights and furthermore, that colonialism and neocolonialism
constituted gross violations of international law and as such
development cooperation was required between the developed and
developing countries. Moreover, the developing countries suggested
that developed countries had a legal obligation to make reparations,
either by transfers of technology, capital or other forms of goods and
services. They believed they were entitled to such transfers and
cooperation and they should not be perceived as charity, aid or
welfare. The Arab oil embargo of 1973 was another factor which
increased attention as it further intensified North-South
relations. Such events led to the search for a 'New International
Economic Order' (NIEO), which according to the developing nations was
a search for equality and justice for all states and its peoples.

It also became evident that the international community was
dissatisfied with the separation of human rights with the adoption of
two covenants. In 1968 the Proclamation of Tehran expressed that
``since all human rights and fundamental freedoms are indivisible, the
full realisation of civil and political rights without the enjoyment
of ESCR's is impossible.'' The Declaration on Social Progress and
Development 1969 emphasised the interdependence of all human rights
and by the 1970's, the notion of a 'right to development' surfaced.

The first direct references to TRTD can be attributed to the
developing countries in the context of the NIEO. In 1966 the
Senegalese Foreign Minister at a GA meeting expressed

\begin{quote}
Not only must we affirm our right to development, but we must also
take the steps which will enable this right to become a reality. We
must build a new system, based not only on the theoretical affirmation
of the sacred rights of people and nations but on the actual enjoyment
of these rights.
\end{quote}

In 1972 at the UN Conference for Trade and Development (UNCTAD III),
in Chile, TRTD was claimed by many of the developing nations. This
notion was elaborated by the Sengalese jurist Keba M'baye who
advocated extensively on the existence of the right and Karel Vasak
who advocated on behalf of third generation solidarity rights. By 1977
the CHR issued a resolution calling upon the Secretary General to
examine and prepare a study on the human right to development and its
relationship with other human rights and the NIEO. It was evident that
the study was not to see whether the right existed, this was
implicitly recognised in the resolution. The Declaration on Race and
Racial Prejudice adopted in 1978 by UNESCO referred to ``the right of
every human being and group to full development''. The GA Declaration
on the Preparation of Societies for Life in Peace 1978 also stated
that everyone has the right ``to determine the road of their
development''. By 1979, having considered the suggestions made by the
Secretary General, the CHR issued a further resolution recognising
TRTD. Whilst a working group was established to prepare a draft
declaration on TRTD, The Banjul Chartergave TRTD legal recognition for
the first time. In 1986, TDRTD was adopted by the GA.

\subsubsection{From the Declaration on the Right to Development and Beyond}

While TDRTD framed TRTD as a human right, this did not bring a closure
to the controversies. TDRTD was adopted after several years of
negotiations between the North and the South - or developed and
developing countries - and as such the text of the declaration was
constructed with heavy compromises. TDRTD was unable to escape the
uneasy political atmosphere which was present during its negotiations
and, although it was adopted by consensus, with 146 votes for the
declaration, there was a notable negative vote made by the United
States. There were also eight abstentions made by Denmark, Finland,
The Federal Republic of Germany, Iceland, Israel, Japan, Sweden and
the United Kingdom. The underlying reasons for the vote against TDRTD
and the abstentions reveals the discontentment surrounding the nature
of TRTD. Objections to TDRTD included: the possibility that the
declaration would lead to an erosion of individual rights; TRTD could
be used to legitimise violations of the rights of citizens; TDRTD did
not make a distinction between individual rights and group rights; the
human rights discourse tended to become diluted and confused with the
addition of TRTD; the declaration provided a mistaken link between
human rights and a NIEO, and it did not take into account the complex
relationship between development, security and disarmament - but
rather provided a very simplistic view of the relationship.

As a result of such objections, the definition and the content of TRTD
have been the crux of debates surrounding the right. The vague
language adopted by TDRTD reflects the complexity of TRTD and the
demands for political compromises.

A Working Group of Governmental Experts on TRTD failed to provide any
concrete recommendations with regards to the nature of the right and
its implementation. Thus in 1989 the UN Commission on Human Rights
asked the Secretary-General to organise a 'Global Consultation'
involving representatives from all arenas including the UN, its
specialised agencies, governments and non-governmental organisations,
to ``focus on the fundamental problems faced by the implementation of
the Declaration, the criteria which might be used to identify progress
and the possible mechanisms for evaluating such progress''.

The affirmation of TRTD in the 1993 Vienna Declaration and Programme
of action as ``a universal an inalienable human right and an integral
part of fundamental human rights'' was of vital significance. Although
the international community had not fully reached consensus on the
actual definition of the right and its content, it nonetheless
affirmed its existence as a human right, with the USA in favour. But
the problem remained and the need to elaborate on TDRTD proved to be
the catalyst for the creation of further working groups on the matter.

In 1997, the GA endorseded that the inclusion of TRTD in the
International Bill of Rights would be an appropriate means of
celebrating the \nth{50} anniversary of the UDHR. The CHR later
recognised that the celebrations were an important opportunity for all
human rights, and in particular TRTD, to be placed at the top of the
agenda for the whole international community. After many years of
discussions and debate, it was evident that, even at this stage,
consensus had not been completely reached. Many States still had their
reservations while others purported TRTD to be something which it was
not. Thus, the Commission recommended the establishment of a further
working group, the Open Ended Working Group (OEWG) and an Independent
Expert (IE). The OEWG was to monitor and review the progress made by
the IE and report its conclusions and findings to the commission. The
mandate of the IE, Arjun Sengupta, was to ``present to the working
group at each of its sessions a study on the current state of progress
in the implementation of the TRTD as a basis for a focused discussion,
taking into account, inter alia, the deliberations and suggestions of
the working group''. One of the first tasks faced by the IE was to
clarify the definition and content of TRTD, a highly controversial
area in which many different nations had their own interpretation.

The precise nature of the right has proved elusive. Many issues have
been brought forward such as; what kind of right it is; who are the
right holders and duty bearers; and what forms of obligations are
imposed and to whom. These issues are the crux of contention
surrounding TRTD and once resolved, if they can be resolved, will
identify the true nature of TRTD and contribute to the identification
of whether there exists a legal obligation for the international
community to realise TRTD.

\chapter{The Definition and Content of TRTD}

\lettrine{T}{he} non-controversial aspects of TRTDare easily
established from TDRTD and these are generally agreed upon by the
international community. This chapter will first summarise the
non-controversial aspects of TRTD. The chapter will then address each
of the controversies surrounding TRTD. This chapter aims to determine
the contours of TRTD and what it entails.

\section{The Non-Controversial Aspects of the Right to Development}

This section will summarise the non-controversial aspects of TRTD
including the principles of universality and indivisibility;
non-discrimination and equality of opportunity; TRTD as a right to
participate in development; accountability and transparency and
finally equity and justice.

\subsection{Universality and \textit{Indivisibility}}

The principles of universality and indivisibility are of crucial
significance to TRTD. Universality is a key principle since, if TRTD
is a human right, it must be applicable to everyone. The notion of
universality has been questioned in a much larger debate concerning
universalism and cultural relativism. Nonetheless, it is evident that
TRTD can be applicable to everyone regardless of cultural background,
gender etc. Article 1(1) defines the right as a human right belonging
to every person and peoples, who are all entitled to participate in,
contribute to, and enjoy TRTD. It is not just a right of individuals
and peoples in developing countries; it can easily be applied to
people in developed countries as well.

Indivisibility is also crucial as TDRTD emphasises that all rights are
interrelated since one right cannot be enjoyed in isolation; its
enjoyment is dependent on other rights and thus all rights should be
given equal attention. This principle is concisely articulated in
Article 6(2) TDRTD and leaves no room for interpretation.

The IE clearly articulates the principles of universality and
indivisibility when he states:

\begin{quote}
[H]uman rights must conform to the principles of \textit{universality}
and \textit{indivisibility. }Universality implies that every
individual is endowed with human rights, by virtue of being human,
irrespective of their cultural background or citizenship.[...]
Indivisibility is associated with the principle of
interdependence. Two rights are indivisible if one cannot be enjoyed
if the other is violated. Two rights are interdependent if the level
of enjoyment of one is dependent on the level of enjoyment of the
other.
\end{quote}

The IE's articulation of the principles of universality and
indivisibility is highly commendable as it's direct and precise,
clearly identifying what the principles are and how they apply to
TRTD. However, Sengupta may be accused of inferring far more depth
into the principles than those agreed within TDRTD and perhaps
exaggerates their meanings as he continues:

\begin{quote}
The obligations related to such rights are also universal, to be
implemented to the best of their possibility by all agents who are in
a position to help, whether they are State authorities or others
belonging to the same country or other States and international
organisations.
\end{quote}

Sengupta suggests that, the implications of the principle of
universality make the obligations related to such rights also
universal. As will be discussed below, the issue of duty-bearers is
still very controversial. Thus Sengupta's notion of ``all agents in a
position to help'' has not been agreed upon by the international
community. Likewise, the obligations of international organisations
and individuals relating to TRTD are also still unclear, and thus the
IE has perhaps interpreted the notion too widely. Having noted this,
the IE's mandate was to clarify the content of TRTD and in this light
the concepts identified by him provide a useful background to what the
right aims towards.

\subsection{Non-Discrimination and Equality of Opportunity}

A common feature of all human rights documents is the emphasis placed
on the principle of non-discrimination. This is a fairly
straightforward principle, intended to protect people from various
forms of discrimination. TDRTD is no exception and the principle of
non-discrimination is clearly conceptualised in Article
6. Nevertheless, the formulation of the principle in TDRTD could have
carried greater force, as Article 5 of TDRTD uses the term ``shall''
implying that states must act, the term ``should'' used in Article 6,
by comparison, could now be construed as less powerful and less
obligatory. Furthermore, both the ICCPR and the ICESCR convey the
principle in more obligatory forms.

TDRTD affirms the principle of equality for all its beneficiaries by
stating that development policies in line with TRTD are aimed ``at the
constant improvement of the well-being of'' not only ``individuals''
but also the ``entire population''. Article 8(1) underscores the need
for states to ensure equal opportunity for everyone to access ``basic
resources, education, health services, food, housing, employment, and
the fair distribution of income''. Women have been highlighted
specifically and they should be able to actively participate in the
development process. The role of women has been emphasised in
subsequent GA resolutions as ``the empowerment of women and their full
participation on a basis of equality in all spheres of society is
fundamental for development''.

\subsection{The Right to Development as a Right to Participate in Development}

One of the main themes of TRTD as formulated in TDRTD is that of
`participation'. There is a distinct emphasis on the principle of
participation throughout TDRTD and it is mentioned in Articles 1,
2(1), 2(3) and Article 8. The interpretation of TRTD as a right to
participate in development has been described by the Global
Consultation as ``the right through which all other rights in the
declaration on the right to development are exercised and
protected''.

The principle of participation has been described by Bunn as:

\begin{quote}
an on-going process at the local, regional, national, and
international levels. It is the primary mechanism for setting goals
toward the realisation of the right to development, and for assuring
the compatibility of development activities with human rights and
cultural values. Participation is also important in evaluating
progress toward the realisation of the right.
\end{quote}

As noted by Bunn, participation is of crucial importance to the
realisation of TRTD. People who are affected by any development
processes must be able to participate in the decisions which affect
their lives. As the individual is the central subject of development,
all policies concerning such development should be accessible to that
individual. The type of participation is also of relevance. It is not
sufficient inform people of the policies being implemented;
participation, in TRTD perspective, requires that people must have
control over the process. This is clearly articulated by the Human
Rights Council of Australia, who extrapolate the human rights meaning
of participation as:

\begin{quote}
Participation understood as control cannot easily be confused with
`involved', `consulted', `empowered' or even `ownership'. To ask who
has control, authority, direction over a particular aspect of the
development program is a much tougher question than to ask who is
involved or empowered by it. It also leads to significantly more
meaningful answers.
\end{quote}

In contrast, the IE view of participation is that it empowers people:

\begin{quote}
Participation is concerned with access to decision-making and the
exercise of power in the execution of projects which lead to the
programme of development. This means that citizens have to be
empowered and have ownership of the programme.
\end{quote}

At first sight it seems as though both interpretations of the meaning
of participation are similar. Rather, on close analysis, the Human
Rights Council of Australia provide a more complex view which in fact
develops that of the IE's. Perhaps the view of the Human Rights
Council is more appropriate for TRTD empowerment of people to be
involved - to have their say - is relevant, but to actually have
control over the decision making brings a new dimension to the
notion. This gives people the \textit{authority} to act according to
their judgment on a given situation. For the purpose of illustration
lets say a development project has aspects which are incompatible with
their views; in this case they will be able to act to remedy the
situation with relative ease. Thus participation is vital to the
realisation of TRTD, particularly if it is seen as a form of control.

\subsection{Accountability and Transparency}

The concepts of accountability and transparency are not specifically
addressed by TDRTD. However, the General Secretary has identified a
connection between accountability and the assessment of activities
with respect to promoting human rights within development as early as
1980.

The concept of accountability infers notions of responsibility and
answerability and is thus quite important for the realisation of
TRTD. The duty holder of the right will be responsible to realise
their obligations; if they do not fulfill their duties they can be
identified and held accountable. The IE's descriptions of
accountability and transparency convey the vital role they play toward
the realisation of TRTD:

\begin{quote}
Accountability involves establishing responsibility for the
non-fulfillment of a right and applying corrective measures. Any
development programme must identify duty-holders and their
obligations.  Transparency is essential for ensuring accountability
because programmes must be designed in such a manner as to bring out
openly all the interrelations and linkages between different actions
and actors.
\end{quote}

Unfortunately, identifying the right holders and duty bearers of TRTD
remains controversial as will be seen below.

\subsection{Equity and Justice}

The principles of equity and justice are quite complex and lie beyond
the scope of this paper, but a brief summary of the principles which
are contained in TDRTD will now be presented. TDRTD stipulates that
all individuals are entitled to ``the fair distribution of the
benefits'' resulting from development. According to Article 8(1)
states must also ensure equal access to basic resources, education and
health services, amongst others. The IE suggests that:

\begin{quote}
Equity is reflected in policies that are aimed at ensuring the
equitable distribution of benefits and that target the most vulnerable
and marginalised segments of society.
\end{quote}

Article 8(1) specifically addresses the notion of ``social
injustices'', focusing on their eradication, and stipulates that
states should carry out both economic and social reforms. This Article
is of considerable importance as some of the key values of TRTD can be
extrapolated from it. In the most concise terms, Bunn describes:

\begin{quote}
Although there is only one reference to ``injustice'' within the text
of the UNDR[T]D (Article 8.1), it is nevertheless clear that much of
the Declaration's underlying intention is aimed precisely at
eliminating it. While international social and economic justice is an
entire topic in itself, the right to development can help provide a
vision for the international order. The right seeks to address complex
issues of equality, fairness, distribution of benefits and burdens
within and among societies. The right focuses on solving problems of
exploitation and oppression, fulfilling basis human needs, and
maintaining respect for all persons. It serves as a reminder of our
responsibility to future generations. Finally, on a global scale, it
raises the classic question of justice: how to render to each person
what is due.
\end{quote}

Thus, TDRTD encompasses an array of different principles, each one as
important as the next, and all are crucial to the effective
realisation of TRTD, which seeks to address a variety of concerns at
the international and national levels and, ultimately, eradicate
injustices of different kinds.

\section{The Controversies Surrounding the Right to Development}

The legal obligation for the international community to realise TRTD
is greatly affected by its clarity. The controversies surrounding TRTD
and the ambiguous language used in TDRTD obscure the substance and
meaning of the right. In order to evaluate if TRTD legally obligates
the international community to realise TRTD, such controversies are
addressed in this section to examine if these can be reconciled to
denote specific legal obligations on the international community. This
chapter addresses the different views and interpretations regarding
the TRTD as a human right; collective rights and TRTD; TRTD as a
synthesis of all rights; TRTD as a right to a process of development
and the duty bearers of TRTD.  Finally, a brief introduction of the
link between security, peace and development is provided at the end of
the chapter.

\subsection{The Right to Development as a Human Right}

Of crucial importance to the understanding of TRTD is the emphasis
placed on the principle of respect for human rights. The word 'right'
or 'rights' has been used in TDRTD at least forty-five times, denoting
several different aspects of TRTD. Firstly, TDRTD proclaimed it as an
``inalienable human right''. This is of fundamental importance as it
is the additional value of claiming a human right which has
significant consequences for many people. Secondly, TRTD is formulated
as a right in which ``economic, social, cultural and political
development'' will accommodate the realisation of all human rights and
fundamental freedoms.

All human beings have the responsibility for development, including
the promotion and protection of ``an appropriate political, social and
economic order for development''. However, a limitation on this is
provided from the principle of respect for human rights which states
that they must take into account ``the need for full respect for their
human rights and fundamental freedoms as well as their duties to the
community''. Such a limitation is provided in order to protect human
rights against those willing to repress them in the pursuit of
development. Furthermore, states should fulfill their duties in such a
manner as to ``encourage the observance and realisation of human
rights'' and they are required to take steps to eliminate the massive
and flagrant violations of human rights. Article 5 expressly uses the
term ``shall'' whereas in other articles ``should'' is used
instead. This implies that states must act in order to eliminate
massive and flagrant violations of human rights. In comparison, the
term ``should'' , used in other Articles, could now be construed as
less powerful and less obligatory. For example, states ``should'' take
steps or make considerations, but it could imply that it is of no
great consequence if they are unable to.

TDRTD reaffirms the principle contained in previous human rights
documents, that all states should co-operate to strengthen the
observance of all human rights and fundamental freedoms without any
distinction as to race, sex, language or religion etc. It further
emphasises that equal attention must be given to all rights, be they
civil, political, economic, social or cultural. States should also
take ``take steps to eliminate obstacles to development resulting from
failure to observe civil and political rights, as well as economic
social and cultural rights''.

Thus it is made evidently clear that the concept of human rights is
intrinsically linked with development. Nevertheless, the proclamation
of TRTD as a human right has by no means settled the debate on whether
TRTD is a human right, a moral claim, a synthesis of all human rights,
a collective right or a right to a process of development which could
also be framed as a human right. Neither has it been determined which
theory of rights it stems from, whether the naturalist or positivist.

Sengupta recognises that the traditional notion of positive rights
assumes that, in order to recognise a right, not only must the
right-holder be identified, but also the duty-bearer with certain
precision and clarity. He asserts that according to this view,

\begin{quote}
to have a right means to have a claim to something of value on other
people, institutions, a state, or the international community, who in
turn have the obligation of providing or helping to provide something
of value.
\end{quote}

Sen articulates positivist human rights with correlated duties, he
states that:

\begin{quote}
Rights are entitlements that require, in this view, correlated
duties. If person A has a right to some x, then there is some agency,
say B, that has a duty to provide A with x.
\end{quote}

Within the human rights regime, each person within the state is deemed
as the right holder and the duty bearer corresponds to that of the
State, thus the correlated entitlements and duties are easily
established. Unfortunately, as TDRTD identifies several possible
right-holders and also several duty bearers, it has thus been argued
by the positivist school that these entities are too vague, making it
impossible to accept TRTD as a human right.

Moreover, according to the positivist school, rights which are not
legally enforceable, cannot be regarded as human rights. Thus any
entitlements which are derived from a human right must be sanctioned
by a legal authority and appropriate legislative
instruments. Disagreeing with this, Sengupta states that;

``This view, however, confuses human rights with legal rights. Human
rights proceed law and are derived not from law but from the concept
of human dignity. There is nothing in principle to prevent a right
being an internationally recognised human right even if it is not
justiciable.''

Naturalist doctrines identify human rights as inherent to the
individual. But as TRTD has been formulated as a right which also
belongs to groups of people it is incompatible with this notion of
human rights. This means that the IE has not identified TRTD as
belonging to either of school of thought, as something within each,
can conflict with the notion of TRTD as a human right. Consequently,
TRTD can not be easily ``pigeon-holed, analogized or catalogued in
terms of one or another theory of moral, neutral or legal
rights''. The above statements lead one to question what exactly human
rights \textit{are}, in order to clarify what kind of right TRTD
\textit{is}. However, the debate surrounding the nature of human
rights, including its foundations and sources, is still ongoing. As
Alston clearly notes:

\begin{quote}
[I]t must be acknowledged that there is substantial disagreement among
philosophers as to the precise nature of human rights, their sources
or foundations and their practical implications.
\end{quote}

He then quotes ``Philosophers disagree on almost every aspect of the
theory of human rights\ldots''. From this it can be concluded that an
in depth analysis on the precise nature of human rights, the
theoretical and philosophical background of these and the issue of
justiciability, are beyond the scope of this paper.

\subsection{Collective Rights}

During the last few decades, increased attention has been drawn to the
subject of 'collective rights', and TRTD has played a prominent role
throughout the discussions, alongside the right to self-determination,
the rights of minorities and the right to a healthy environment, to
name a few.

The introduction of these rights has led to some confusion surrounding
human rights and their collective nature. The confusion begins from
the very terminology used to refer to these rights as not only the
term 'collective rights' is used, but also a whole array of
expressions including, 'solidarity rights', 'communitarian rights',
'group rights' and 'third generation rights'. Collective rights have
been advanced as third generation, solidarity rights, distinguishing
them from other rights. The concept of generations of human rights
first emerged in the 1970's by Karel Vasak. Vasak proposed third
generation rights as rights of solidarity, and this generation of
human rights was said to include TRTD. According to him, first
generation rights are CPR's, which impose negative obligations on the
state to refrain from infringing a citizen's liberty. Second
generation rights are ESCR's which require positive obligations for
the state to act in some way to enforce these rights. Third generation
rights are rights which have emerged from global interdependence and
problems of a global nature requiring international solidarity.

There have been many arguments put forward which have opposed the
notion of collective rights or third generation rights and also many
which have been strongly in favour of their emergence. These arguments
will be described in turn, starting from the point of view of the
critics and then from the proponents of collective rights. I will then
seek to analyse these, to see if such arguments contribute to the
clarification of TRTD or result in further confusion.

The predominately western view is that the notion of collective
rights, and the tendency to proclaim them and afford them the status
of 'human rights', is very dangerous ground for a number of reasons
and should be approached with extreme caution.

Firstly, with regards to the definition and concept of human rights,
it has been argued that human rights in their very nature are rights
which have been afforded to human beings by virtue of being human,
and, thus, by definition, cannot be afforded to a
collective. Consequently, it has been argued that specific conditions
must be met in order for a right to be classified as a human right and
such conditions have not been met by collective rights including TRTD.

Secondly, the nature of collective rights has been criticised as it is
unclear who the right holders and duty-bearers are and also who would
represent the collective. Thus it has been argued that the traditional
notion of human rights could be eroded if collective rights which are
vague and ambiguous are afforded the status of human rights.

Thirdly, there is concern that the existence and proclamation of
collective rights may be politically abused by states and groups in
order to justify violations of individual rights. Overall, the critics
of collective rights argue that because of their very nature they are
fundamentally incompatible with the human rights discourse.  Finally,
the notion of positive and negative rights in the first and second
generation of right, as postulated in Vasak's argument, have been
heavily criticised. It has now been widely established that both
generations require positive and negative actions by States.

In defending collective rights, the proponents have argued that,
firstly, those who believe that by definition collective rights cannot
be human rights are mostly influenced by the natural law doctrines,
which are in itself incompatible with the collective rights
approach. But, if human rights are universal and based on universal
values, then the natural law doctrine should not be the only one
consulted when establishing human rights. Thus accordingly, human
rights are dynamic, have universal values and are not solely based on
natural law theories and therefore collective rights can be human
rights.

In rebutting the criticism regarding the vagueness of collective
rights, it has been argued that some individual rights are also vague
and thus such a reason cannot be used to deny the emergence of
collective rights. Additionally, some collective rights have been
evolving for many decades, including the rights of minorities, self
determination and people who are protected from genocide. It has been
claimed that some collective rights are \textit{de lege lata} or at
least \textit{de lege ferenda}.

Tensions between collective rights and individual rights cannot be
used as an argument against collective rights as such tensions exist
between rights anyhow and collective rights are not exceptional in
this regard. Furthermore, all rights are liable to be politically
abused and thus to reject collective rights from this angle cannot be
justifiable.

With regards to the beneficiaries of collective rights, at least four
can be identified;


\begin{enumerate}
\item the global community, e.g. the right to peace
\item the state, e.g. TRTD, the right to self determination
\item a group within a state, e.g. TRTD,  indigenous rights
\item the individual who belongs to a group, e.g. TRTD, minority
  rights including cultural identity
\end{enumerate}

As human rights have traditionally been regarded as belonging to each
individual of the human race, every person, individually, is the
beneficiary of those rights. However, with the emergence of the
concept of collective rights, the beneficiary of the right can be any
of the entities described above and thus conflicts between the
entities claiming such a right can emerge. It is in such a situation
that the rights of individuals can be curtailed for the benefit of one
of the other beneficiaries within the collective rights domain. Not
only is it possible for an individual's rights to be curtailed but it
is also possible that there could be conflicts between each of the
beneficiaries within the collective right in question. For example,
the right to retain one's own cultural identity can conflict with
TRTD, as Prottclearly articulates:

\begin{quote}
[T]he assertion of the right to development may run directly counter
to the right to preserve or develop a culture. Economic development
may obliterate or mutilate important cultural sites and destroy social
structures which are essential for the survival of traditional arts
and other cultural activities. Pressures to exploit tourism as a
source of foreign revenue often lead to degradation of traditional
crafts to cater with the increased demand, to the provision of
services for tourists inimical to the local environment, and to damage
to static life-styles from the constant stamp of tourist
feet. Economic development may mean cultural stagnation: cultural
development may mean economic stagnation.
\end{quote}

It seems evident that, although Prott's criticism on the conflicting
nature between cultural rights and TRTD has some ground, he has failed
to note that TRTD is not only an economic right but is also a social
and cultural right as well as reinforcing CPR's. Thus in the scenarios
proposed, economic factors will not be the sole indicators of whether
TRTD has been fulfilled or not; if economic factors are the only
indicator of fulfilment, then TRTD itself has not been
realised. Furthermore, the Vienna Declaration underlined the
indivisibility of all human rights.

From this analysis of collective rights as a whole, they seem to be
completely irreconcilable with the concept of human rights as
individual rights. Sieghart's description of the process of first
claiming to be acting on behalf of the 'people' and then leading to
the violation and abuse of exactly those 'peoples' rights, cannot be
framed any clearer. The danger is evident. It is for this precise
reason that the nature of the beneficiaries of collective rights have
clearly caused controversy within the human rights
domain. Nonetheless, the concept of 'collective rights' cannot be
disregarded; each right which emerges as a collective right needs to
be analysed on its own.

TRTD as a collective right, as both Prott and Sieghart have suggested,
may conflict with not only individual rights but also with other
collective rights. TDRTD recognises that 'every human person and all
peoples' are the beneficiaries of TRTD, which in itself is
ambiguous. The concept of 'every human person' is already firmly
established in human rights as meaning every individual belonging to
the human race, but the concept of 'all peoples' is clearly uncertain;
it could mean all peoples in the world community, all peoples
belonging to a State, all peoples belonging to a group within a state
and also all peoples as individuals within a group, within a
state. The reasons for criticising TRTD become evident.

The collective dimension of TRTD has been claimed by some states as a
right of states. The basis for this claim comes from the preamble of
TDRTD:

\begin{quote}
the right to development is an inalienable human right and that
equality of opportunity for development is a prerogative both of
nations and of individuals who make up nations.
\end{quote}

The content of this article confirms TRTD as a human right, and then
states that equality of opportunity for development is a prerogative
of nations. The exact substance of this article and what it denotes is
unclear. Two very different concepts are merged in one sentence and
thus its meaning becomes obscured. It has been argued by many states
that this article identifies TRTD as a right of states alongside other
articles.

Article 3(3), in conjunction with the above preambular sentence, can
be interpreted to give substance to such as a claim as it stated that
``States should realise their rights and fulfil their duties in such a
manner as to promote a new international economic order''. The words
\textit{States should realise their rights} infers the possibility
that states hold TRTD but is far too ambiguous and fails to identify
what kind of rights, belonging to the state, are to be realised in the
promotion of a new economic order. Although TDRTD is vague, in
absolute terms, it does not explicitly or implicitly recognise TRTD as
a right of states, and the formulation of the right as a right of
states has been heavily criticised. Ghai has suggested that some Asian
governments have advocated on behalf of TRTD for the purposes of
seeking ``to promote the ideology of developmentalism which justifies
repression at home and the evasion of responsibility abroad''. He
further suggests that governments are supporting TRTD as part of a
wider goal of establishing ``the primacy of economic development over
human rights.'' Furthermore, Donnelly argues that if TRTD can be
claimed as a right of states, ``[t]he danger here is that the state is
thereby placed in a position to use its human rights to deny the human
rights of individuals, while still plausibly claiming to be pursuing
human rights.''

Nonetheless, from the offset, TDRTD declared that all aspects of the
declaration ``are indivisible and interdependent and each of them
should be considered in the context of the whole''. Thus although TRTD
undoubtedly has a collective dimension, this is to be construed in the
light that ``The human person is the central subject of development
and should be the active participant and beneficiary of the right to
development'', and this aspect was confirmed by the Vienna
Declaration.

Article 2(3) sheds light on how TRTD can be conceived as a right of
states. This article identifies the state as a holder of TRTD, but as
Orford suggests the state is acting

\begin{quote}
as the agent 'of the entire population and of all individuals'. By
Implication, the right is exercisable by the state against those with
the power to deny or constrain the capacity of the state to formulate
national development policies that benefit the entire population and
all individuals.
\end{quote}

Consequently, Alston suggests that, in this context, TRTD is a right
of states insofar as the state is a ``medium through which the rights
of individuals are effectively asserted vis-a-vis the international
\textit{community}''. Such a formulation of TRTD is the most
comprehensible and one which poses little threat to the human rights
field. Thus, although there are valid criticisms regarding such a
controversial aspect of TRTD, these comments are unnecessary if TRTD
is seen as a right of states acting as agents for the well-being of
the whole population and all individuals. It then becomes apparent
that what was meant by TRTD as a right of states is that the state can
act as agent on behalf of its people but it cannot claim rights for
itself.

TDRTD affirms that every human person and all peoples are the right
holders of TRTD in Article 1(1). To date, this aspect of TRTD is still
very contentious. Not only is there disagreement amongst scholars
surrounding this dimension, but, more importantly, consensus has still
not been reached amongst states. Thus, without the resolution or a
compromise as to the meaning and definition of the collective
dimension of TRTD, uncertainty regarding it will remain.

Having noted the dangers posed by the emergence of the concept of
collective rights, it would be a gross misconception to identify the
whole concept as a fallacy. The right-holders, albeit sometimes
uncertain are equally in some other cases quite determinable. For
example, the right-holder of a right to peace, which has also been
suggested as belonging to the third generation of human rights, is
every person within the world community, and the duty-bearer is thus
evidently the people in power~--- ordinarily the governments of each
and every state in the world. Thus the conflict of rights as described
by Sieghart will not occur. With regards to TRTD, such a problem would
not be posed if the right holders were to be all peoples belonging to
the world community at large. The problem lies in the fact that TRTD
has also been claimed as a right of states, and it is this claim which
can be seen as irreconcilable with the whole human rights
framework. As Galenkemp states, commenting on collective rights as a
whole:

\begin{quote}
it seems fair to conclude that the debate on collective rights to
date has been conducted in rather extreme terms, resulting in
polarisation. The ordinary reader is left with the impression of two
irreconcilable camps.
\end{quote}

This denotes that because of the collective dimension of TRTD, the
legal obligation for the international community to realise TRTD is
highly questionable and further clarification of this dimension of the
right, is necessary before any legal obligation can be assumed.
Nevertheless, TRTD is not only seen as a collective right, it is also
a right of every person and, even though it is a right of all peoples,
the human person is the central subject of TRTD. Thus the
individualistic component of TRTD cannot be disregarded, but this
still does not denote that the international community is legally
obligated to realise TRTD.

\subsection{The Right to Development as a Synthesis of All Rights}

TDRTD clearly articulates that all human rights are indivisible and
confirms that priority must not be given to TRTD over other human
rights. This was reiterated with a stronger sense of affirmation in
the Vienna Declaration. TRTD as a synthesis of human rights has been
proclaimed on a number of instances and suggested by scholars such as
Alston. The concept was evident throughout the early working group
discussions and many of the experts considered TRTD as embodying all
rights and as the ``synthesis of already defined rights''. This led to
the Secretary-General to affirm ``[t]he approach by which the right to
development is viewed as a synthesis of a large number of human rights
has found favor with a number of commentators''. Reiterating this
concept, it was stated that ``in the view of most experts, this right
is a combination of existing rights recognised by the international
community and contributing to the development of peoples and states''.
Thus, the relationship between TRTD and other human rights can be
deemed as crucial to the understanding of TRTD and what it
constitutes.

Having noted this, it has been expressed by a number of commentators
that perceiving TRTD as synthesis of all other rights adds nothing new
to the human rights discourse and ``takes attention away from the
specific rights, for example, speech, assembly, social welfare, to an
ambiguous portmanteau right to development''. Donnelly expresses the
same sentiment but further adds:

\begin{quote}
Discussion of the right to development as a \textit{synthesis of all
  human rights} shifts the focus from particular rights to the package
as whole. This is most likely to lead not to a broader and more
comprehensive view but rather an increasing detachment from the
realities of implementation in particular cases; that is, it is likely
to obscure the central fact that progress in realising the whole must
be achieved by hard work in implementing the \textit{parts}, the
separate civil, political, Economic, social and cultural rights
already recognised internationally.
\end{quote}

While Donnelly and Ghai make some valid points regarding the need to
implement recognised human rights and questioning the value of TRTD as
a synthesis of all rights, both detract from the notion of
indivisibility of all human rights which has been unequivocally
affirmed in a variety of international human rights
instruments. Orford argues ``that the focus on the indivisibility of
rights is an important aspect'' of TRTD.

While Orford clearly identifies the role of the concept of
indivisibility of all human rights and does so in a very concise
manner, she fails to address the notion of TRTD as a synthesis of all
rights. Stating that all human rights are indivisible is a very valid
point - but it falls short of arguing that TRTD entails all these
rights, and Orford fails to grasp it. It is clear that Donelly is not
arguing the value of the concept of indivisibility of all rights but a
new notion that TRTD encompasses all those specific rights. There is
no attempt within his text to undermine the principle of
indivisibility of all rights when he states that attention should be
given ``to implementing the ``parts'', the separate political, civil,
political, Economic, social and cultural rights already recognised
internationally.'' He does not infer in any way that such ``parts''
are to be implemented unequally or in some form of hierarchical
order. Having noted this, Donnelly's view that TRTD as a synthesis of
all rights may obscure individual rights is a very restrictive
one. TRTD aims at complementing all rights and strengthening their
implementation through a rights-based approach to development.

However, the IE has notably disagreed with the concept that TRTD is
just a ``synthesis of all human rights''. Rather, he describes:

\begin{quote}
The right to development is a composite right to a process of
development; it is not just an ``umbrella'' right, or the sum of a set
of rights. The integrity of these rights implies that if any of them
is violated, the whole composite right to development is also
violated. The independent expert describes this in terms of a
``vector'' of human rights composed of various elements that represent
the various economic, social and cultural as we as the civil and
political rights. The realisation of the right to development requires
an improvement of some, or at least one of those rights without
violating any other.
\end{quote}

TRTD as a ``composite right to a process of development'' is much more
than TRTD as a ``synthesis of all human rights'', and this has been
clearly articulated by the IE. The ``composite right'' element,
identified as a vector, adds substance and value to the right - which
the synthesis approach did not - and clearly identifies the purpose of
TRTD as a human right. It would now be beneficial to understand what
the process of development entails in the context of TRTD, since one
of its key features is the wide definition of development.

\subsection{The Right to Development as a Right to a Process of Development}

The task to define development is, \textit{per se}, not the purpose of
this paper, and many will acknowledge that it has been the subject of
much scholarly attention. This notwithstanding, since development is a
core element of TRTD, its definition is of crucial significance. TDRTD
defines development in the preamble of TDRTD. This is a wide
interpretation, comprised not just of economic aspects but also
social, cultural and politic elements, and thus it varies
fundamentally from traditional notions of economic development. What
development \textit{is}, is crucial to understanding what people are
claiming when they assert they have the right to
\textit{development}. It is unfortunate that this aspect of TRTD has
gone relatively unnoticed in the literature available. It can be noted
that whilst writers have acknowledged that the definition of
development in TRTD is uncertain and vague, any in-depth analysis or
even suggestion as to what it might constitute have been notably
absent. Thus, in the context of TRTD, the question \textit{what is
  development?} remains at large open.

Development in its purely economic sense is ``seen as simultaneously
the vision of a better life, a life materially richer, institutionally
more modern and technologically more efficient and an array of means
to achieve that vision''. This definition of development is typically
associated with the capitalist model and seems to imply that material
wealth is of key importance in order to have a better life, but fails
to take into account that the individual may be able to develop
himself through social, cultural and political means, and, in doing
so, contribute to the development of society or the state as a
whole. In addition it does not address cases where such a vision may
not be possible as a reality due to the circumstances surrounding that
individual, as for example in cases of extreme poverty.

Having noted this, Rodney, dedicating a whole book on \textit{How
  Europe Underdeveloped Africa}, suggests what development could mean,
in the context of identifying what underdevelopment constitutes. While
offering different definitions of development and noting that
``development in human society is a many-sided process'', Rodney
states that development, ``[a]t the level of the individual, implies
increased skill and capacity, greater freedom, creativity,
self-discipline, responsibility, and material well-being''. With
regards to groups, Rodney states that; ``development implies an
increasing capacity to regulate both internal and external
relationships''. Hence it can be easily established that development
in TRTD is something more than merely economic development. This is
not to say that economic development is not important; quite on the
contrary. Development in the social, cultural and political sense will
be curtailed without the economic component. Economic development is
an indivisible and interdependent aspect of the `whole' of
development, in a similar way that all human rights are interdependent
and indivisible: one human right cannot be fulfilled at the expense of
another and all are interconnected, as are the different facets of
development.

From this perspective it becomes clear why traditional notions of
economic development are not adequate to determine whether or not the
people within a state have developed. The social, cultural and
political aspects of development have emerged to bridge this gap, and
take into account not just the state as a whole but also the
development of the people~--- as opposed to the traditional notion
where the individual is not of paramount significance, as evidenced by
the indices of pure economic development such as GDP and GNP, which
focus only on economic growth and ignore the individual. The need for
a human rights approach to development, or at least a more
people-centred approach, became apparent.

Many institutions already incorporated human rights or a
people-centred approach into their policies, but it has been argued
that all such incorporations were merely rhetoric, and were making
very little difference for people at the grass-roots level. Uvin felt
that:

\begin{quote}
During the 1990s, bilateral and multilateral aid agencies published a
slew of policy statements, guidelines and documents on the
incorporation of human rights in their mandate. An enormous amount of
this work was little more than thinly disguised repackaging of old
wine in new bottles.
\end{quote}

Similarly, Frankovits is also highly critical of the incorporation of
human rights terminology by donors and also the World Bank.

Taking all this into account, Allot's General Felicity Index (GFI) was
one of the first indices to recognise that, to measure development,
one must measure ``not just the increase in the number of factories or
expansion of services, but basically whether life is happier and more
fruitful and enjoyable for the individual. In doing this, one has to
balance one factor against another''. Such an approach can be seen in
the UNDP Human Development Reports, where a wide range of indices are
used to measure development.

TRTD has not escaped being criticised for its economic bias, as
eloquently argued by Charlesworth. Her criticism is highly valuable
because it points out that the ``international law of development''
has assumed development to mean industrialisation and westernisation,
and, although TRTD contains social, cultural and political elements,
the foundations on which these have been erected were markedly
economic. Underdevelopment has been defined by the model of a
capitalist economy and thus any state attempting to recover from such
`underdevelopment' will, by proxy, be basing the evaluation of that
recovery with what development is, as dictated by the model of a
capitalist economy. This undermines a human rights approach to
development and to TRTD. Assuming that development means
industrialisation and westernisation, the suggestion made by
Charlesworth could imply that in asserting one's TRTD, one is
asserting the right to industrialisation and westernisation. This is a
bit far fetched since there are many states and peoples that support
TRTD~--- or at any rate, their understanding of it - and have no
desire to be westernised or industrialised. The reality is in fact
that the world community is culturally diverse and one model of
development cannot possibly fit all nations.  Furthermore, such an
approach to development would not be acceptable to many developing
countries, considering the fear of being controlled by developed
states, in the form of neo-colonialism.  Charlesworth fails to take
into account that in fact many developing countries do not envisage
TRTD as a right to westernisation or industrialisation; although they
may assert they are entitled to some economic assistance, it has never
been suggested that this assistance is for westernisation.

With the approach to development expanding, the notion of TRTD as a
right to a process of development emerged. Article 1(1) of TDRTD
provides a definition of TRTD, while this does not explicitly mention
the word 'process', this is implied in the preamble of TDRTD which
provides a definition of development and specifically addresses the
notion of development as a process. Sieghart, expressed this notion as
early as 1986. He stated that;

``If one defines development as a process designed, progressively, to
create conditions in which \textit{every} individual can enjoy,
exercise, and utilise, under the rule of law, \textit{all} his human
rights ... then the individual's right to development can be simply
defined as his right to participate in, benefit from, that process
without any of the prohibited forms of discrimination''.

However, it is clear that the IE expands on the notion of a process to
development, from that described by Sieghart above, as the IE contends
that the actual process is a human right in itself. Distinguishing the
outcomes of a process of development from the means of how that
process is implemented. This approach to development is very similar
to that advocated by Sen in his work \textit{Development} \textit{as}
\textit{Freedom}. The justification of claiming the actual process of
development as a human right and not only the outcomes of that process
was put forward in terms of the concept of a \textit{metaright} by
Sen.  A clear conceptualisation of the \textit{metaright} is contained
in the following quote, which states that;

``a \textit{metaright} to something \textit{x} can be defined as the
right to have policies \textit{p(x)} that genuinely pursue the
objective of making the right to \textit{x} realisable ... So a right
to \textit{x}~--- such as the right to food or the right to
livelihood~--- may be an abstract, background right, but to give a
person the right to demand that policy be directed towards securing
objectives of making the right to food or the right to livelihood a
realisable right is a \textit{right to p(x)}, which is a
\textit{metaright to x} a real right. So, even if the right to
\textit{x} remains unfulfilled or immediately unrealisable, the
metaright to x, p(x) can be a fully valid right if all the obligations
associated with p(x) can be clearly specified. Therefore, if the
programme of actions and measures that is associated with realising
TRTD are designed in a way that all the obligations ... have been
clearly specified, that programme implemented as a process of
development can be regarded as a \textit{metaright}. That programme
can be claimed as a right entailing enforceable obligations on the
duty-bearers and can be realisable, with a high probability of leading
to the realisation of the background rights.''

This extension of the right to a process of development cannot be
articulated better in any other form. The concept of a process of
development entailing specified obligations on different agencies, as
a \textit{metaright}, to realise TRTD, is of significant value as it
clarifies the contours of the right and adds substance to TRTD as a
whole.

Overall, the definition of TRTD as a right to a process of development
elaborated by the IE is certainly the most appealing. It identifies
the contours of TRTD, although some may be critical in accepting the
\textit{metaright} as a human right. Piron suggests that the IE has
defined this notion of development as a compromise, acceptable to
developing and developed states. Whilst this definition is clear and
concise, it can be noted that disagreement still continues, which
overall undermines TRTD and the prospects of its implementation and
creates uncertainty regarding any legal obligation to realise TRTD.

\subsection{International Co-operation v National Implementation}

Traditionally, international law regulates the relations between
states and international human rights law does not envisage a state
having a duty to realise human rights in any other state but its own.
Increasingly, however, with the emergence of globalisation,
international law is attempting to go beyond the realms of regulating
state actions and into those of International Organisations, including
international financial institutions such as the World Bank,
International Monetary Fund (IMF) and the Word Trade Organisation
(WTO), as the actions of these actors are increasingly affecting human
rights across the globe. TDRTD identifies several actors as having the
responsibility for realising TRTD, including all human beings, States
and the international community. There is a suggestion that
International Organisations also have the responsibility to cooperate
in the realisation of TRTD. However, as has been noted, the concept of
all these actors as the duty-bearers of TRTD in essence could dilute
the right.

Although many states have acknowledged the principle of respect for
human rights, they do not have positive obligations to provide
assistance to another state's citizens' to realise those citizens
human rights. Under international human rights law they are only
obliged to respect, promote and protect the rights of their own
citizens. Thus the legal obligation to realise TRTD by the
international community is questionable. Nevertheless, the duty to
cooperate with a state who is attempting to realise the rights of its
citizens is another issue, but this is also complicated by the fact
that some developing countries have interpreted TRTD as a right of
states.

The two key political sides view the provisions on State
implementation and international cooperation in two different
lights. Developing states argue that there is a duty for international
cooperation for the realisation of TRTD, while developed countries
argue that the focus of TDRTD is on state implementation and TDRTD in
no way articulates there is a legal duty to cooperate. Thus the extent
of which the international community has a duty to cooperate in
realising TRTD is still unresolved. This issue will be addressed
briefly in this section and analysed further in the next
chapter.

\subsubsection{All Human Beings as the Duty-Bearers of the Right to Development}

TDRTD identifies all human beings, individually and collectively, as
having the responsibility for development. Such responsibility entails
that they take into account respect for their human rights and
fundamental freedoms and also their duties to the community. Such
consideration will ensure their fulfilment as a human being and, for
this reason, all human beings are responsible ``to promote and protect
an \textit{appropriate political, social and economic order for
  development}''. This article fundamentally recognises that although
all human beings are entitled to enjoy all human rights, their
responsibilities extend to the whole community and thus, whilst
enjoying their human rights, they must respect those belonging to
others. According to the analysis of the IE, the obligations of all
human beings are then those considered as \textit{imperfect
  obligations} and all human beings in a position to help are
obligated to realise the right.

The obligation of all human beings to respect and promote human rights
was further recognised by the Commission on Global Governance, which,
in 1995, called for recognition that ``governments are only one source
of threats to human rights'' and that ``all citizens \ldots should
accept that obligation to recognise and help promote the rights of
others''. However, the legal obligation for all people to realise TRTD
is questionable as international law does not regulate the behaviour
of people towards each other.


\subsubsection{States as the Duty-Bearers of the Right to Development}

International human rights law does not envisage a state having a duty
to realise human rights in any other state but its own.

TDRTD emphasises that states have the primary duty to create national
and international conditions to realise TRTD. Although this can be
construed as meaning that states have the primary duty to realise
TRTD, it does however cast doubts to this element as it could imply
that states have the duty only to create favorable international and
national conditions for the realisation of TRTD, but they can stop
short there and not actually implement anything which would realise
the right. Furthermore, the term `favorable conditions' is very
ambiguous as it could either be interpreted subjectively, in which
case each state determines what the favorable conditions
are. Alternatively, `favorable conditions' could imply that they have
to be `favorable' with a certain model of development in mind. For
example, the capitalist model could be advanced by many of the
developed states as the appropriate method to achieve `favorable
conditions' for the realisation of TRTD, which may not be acceptable
to some of the developing countries.

However, Article 2(3) clearly stipulates that states have the right
and duty to formulate national policies, implying that the state
itself decides what is appropriate or not~--- although, the policies
must be aimed at the constant improvement of the entire population and
all individuals~--- for the realisation of TRTD. The Working Group on
TRTD firmly stated that;

``Sates have the primary responsibility to ensure the conditions
necessary for the enjoyment of the right to development, as both an
individual and a collective right. Development cannot be seen as an
imported phenomenon or one that is based on the charity of developed
countries.''

This reinforces that TRTD is to be realised by states for their
people, and implemented through appropriate national
policies. Udumbana further suggests that developing countries must
adopt an appropriate definition of development which suits their own
needs. This in accordance with the view that states themselves should
ultimately decide what appropriate national policies they will need to
implement to create conditions which are favourable for the
realisation of TRTD. Furthermore, states should not attempt to import
foreign notions of development which may not work for their own state
and should re-evaluate what is meant by development in TRTD in the
context of their own circumstances.

The IE describes TRTD as a right to a particular process of
development. This particular process~--- and thus TRTD~--- would
expand the basic capabilities of the individual to enjoy their human
rights and also enables the realisation of all human rights and
fundamental freedoms.  The IE suggests that states adopt a development
plan which incorporates the principles of equity, non-discrimination,
transparency, accountability and participation. He considers a
step-by-step approach appropriate in realising TRTD as a concrete
programme. From this, the IE identifies that immediate action in
realising the rights to food, primary education and health is required
as part of the process for the realisation of TRTD. Although the IE
expresses that this does not imply that any right takes precedence
over another, it seems as though such an approach to the realisation
of TRTD is controversial as the international community had finally,
with the adoption of TDRTD accepted that all human rights were
indivisible and interdependent.

The duties of states to realise human rights have been enumerated by
Steiner and Alston to include several obligations. The Maastricht
Guidelines on Economic, social and cultural rights identify the
obligations of states to respect, protect and fulfill those
rights. Marks further suggests that states have at least four
obligations to realise TRTD, two 'perfect' and two 'imperfect'. The
two perfect obligations, include those to respect and to protect,
which are justiciable and capable of being enforced by a judicial
process. While the obligations to provide and promote are ``general
commitments to pursue a certain policy or achieve certain
results''. Marks considers these obligations as non-justiciable but
nevertheless as legal obligations, as states are required to take the
necessary steps ``in the direction of sound progressive realisation of
the right''.

However, the difficulty still remains as to who decides what the
\textit{international} `favourable conditions' for the realisation of
TRTD are, as it is already established that each state has the duty to
create such conditions but, as the whole international community is
involved, the notion can differ from state to state. This is primarily
the reason why the provisions on international cooperation are highly
controversial.


\subsubsection{The International Community as the Duty-Bearer of the Right to Development}

TDRTD declares that States have a duty to co-operate with each
other. This has caused some contention in the international community
as under traditional international law, there is no such duty.  It has
been argued that the legal obligation for international cooperation
stems from the UN Charter and the International Bill of Rights. This
issue will be analysed in the following chapter to see if there is any
foundation to such a claim and whether or not there is a legal
obligation for international cooperation to fulfill TRTD.

TDRTD stipulates that it is the primary duty of the state to realise
TRTD, however in some circumstances it may not be possible for the
state to implement the right and for this reason the DRTD calls upon
the international community to cooperate to enable states to fulfill
their obligations on TRTD. TRTD does not explicitly recognise a legal
obligation for the international community to realise TRTD.

The notions of a duty to cooperate contained in TDRTD stipulate that
this duty exists in ensuring development and eliminating obstacles to
development. Furthermore, based on principles such as sovereign
equality, interdependence, mutual interest and co-operation, States
should realise their rights and duties in such manner which promotes a
new international order. Moreover, States should encourage observance
to and realisation of, human rights. Article 4 stipulates that states
have a duty to formulate international policies, both individually and
collectively. It also underlines that effective international
cooperation is essential to provide developing countries with the
appropriate means and facilities to foster their comprehensive
development. The IE suggests that in order to fully appreciate the
emphasis TDRTD places on international cooperation, Article 4 should
be read alongside the preamble~--- which refers to Article 1 of the UN
Charter . Subsequently, when this is all read in conjunction with
Articles 55 and 56 of the Charter, the commitment of international
cooperation to realise TRTD is further strengthened.

The IE has expressed the view that TRTD should be realised through
coordinated efforts, including developing states, developed states,
international organisations, international financial institutions and
NGO's. He advocates on behalf of a `\textit{Development Compact}' (DC)
between developing countries and donor countries or international
financial institutions. This underscores the importance of
international cooperation in realising TRTD. Although the DC would be
specific to each state, the DC envisages that developing countries
undertake to fulfill the realisation of human rights within their
state while the international community provides certain resources and
shares the cost burden. Furthermore, at the international level,
states should act on measures which would allow;


\begin{itemize}

\item trade and access to markets
\item debt adjustment for the poorest countries
\item transfer of resources and technology
\item protection of migrants and labour standards
\item restructuring of the international financial system to give the
  developing countries a greater share in power and decision-making
  and to increase the flow of private capital to their economies.''
\end{itemize}

Although TDRTD declares that states have a duty to cooperate with each
other, it is difficult to establish to what extent international
cooperation is required and what kind of international
cooperation. The above suggestions serve to clarify the notion but it
is still clear that this aspect has not been resolved by the
international community.  Nevertheless, it is clear that TDRTD denotes
some obligations on the international community to realise TRTD but
whether or not these are legal, is highly questionable.

Donnelly has heavily criticised the notion that obligations of
realising TRTD can extend to actors other than states. According to
him, human rights have traditionally been conceived as held only
against a state by its citizens and ``are essentially instruments to
protect the individual against the state or to assure that the state
guarantees to each individual certain minimum gods, services and
opportunities''. Donnelly argues that the focus of TRTD on violators
other than states functions as a tool to avoid state responsibility
for human rights violations and thus the emergence of TRTD has allowed
repressive governments ``to shift attention from particular rights to
general issues, and from the primary role of the state as a violator
of human rights to external forces that also contribute to human
rights violations''. Donnelly envisages TRTD as a tool ``to use as an
excuse not to act on human rights now'' and because of this, he urges
that people must:

\begin{quote}
not lose sight of the fact that most human rights violations are
directly perpetrated on people by the governments of their own
countries. Discussions of the right to development, however, seem to
have the effect, and perhaps even the intent, of obscuring this
central point.
\end{quote}

Ghai takes a similar stance as Donnelly and believes that the
obligation to fulfill TRTD by actors other than states will ``provide
an alternative framework for the international discourse on human
rights'' and it ``shifts the focus from domestic arenas (where most
violations of rights take place) to the international''.

Some of the arguments put forward by both Ghai and Donnelly seem very
persuasive at first glance. Orford acknowledges that some governments
have attempted to claim that the realisation of TRTD takes precedence
over other human rights and have restricted human rights because of
this. Nevertheless, Orford suggests that arguments put forward by such
critics fundamentally ignore the influence and power which
international institutions have and their affects on human rights
globally. Orford argues that states no longer have the sovereign power
to determine the ``economic, social and cultural conditions in which
people live'' or power over ``policy decisions that shape access to
resources and services and determine the nature of constitutional and
governmental systems''. Agreeing with Alston that globalisation has
had serious effects to state sovereignty, Orford states that:

\begin{quote}
Economic globalisation has made the fictitious nature of state
sovereignty apparent to all but the most myopic observer of
international relations and international law.
\end{quote}

Moreover, Orford argues that;

\begin{quote}
The arguments of critics like Donnelly fail to address situations
where individuals or peoples do not need protection only or primarily
against the state, but also from other powerful states, transnational
corporations (TNC's) or international institutions. Nor do such
arguments address situations where actors other than states make
decisions about the provision of goods, services and opportunities,
and thus about the protection of Economic, social and cultural
rights. In many cases in the globalised economy, other states,
international organisations or foreign investors will be in a position
to deny or effect those protections or guarantees.
\end{quote}

As international law concerns relations between states and human
rights law concerns state behavior toward its citizens, it is
difficult to apprehend how non-state actors can be held accountable
for direct and indirect violations of human rights. States are
responsible for the activities of non-state actors on their
territories but many actions of organisations which are likely to
violate human rights may not happen on the territory of a state - and
thus that state is not accountable to prevent any violations by that
particular organisation.  It is for these reasons that states are not
the only violators of human rights and the notion of international
cooperation contained in TDRTD extends to the behavior of states in
their international relations. This concludes that the element of
international cooperation is significant in realising TRTD by both
states and non-state actors in an increasingly globalised world.

\subsubsection{International Organisations as the Duty-Bearers of the Right to Development}

The concept that international organisations have a duty to realise
TRTD is not without some controversy. Nevertheless, an an increasingly
globalised world, international actors other than states, play a
pivotal role in international relations. As institutions like the WB,
IMF and WTD consist of states, it has been argued that the obligations
of states~--- as part of these organisations~--- include the need to
make the activities of these institutions, compatible with human
rights and TRTD. For example, international cooperation from this
perspective includes the removal of trade barriers.

\subsection{Security, Peace and the Right to Development}

The link between security, peace and development is made in Article 7
of the DRTD. Along the lines of the UN charter, Article 7 stipulated
that states ``should promote the establishment, maintenance and
strengthening of international peace''. Moreover, to achieve this end
states are required to do their ``utmost to achieve general and
complete disarmament''. These measures will benefit the development
process as, states, in achieving general and complete disarmament
should ensure that the resources released by such measures are used
for comprehensive development. Unfortunately this is not without
controversy as Article 7 could be seen to imply differentiated
responsibilities for developing and developed countries: one can
interpret it to mean that the latter should allocate part of these
revenues to help the development process in developing countries.

\chapter{The Legal Status of the Right to Development in International Law}

\lettrine{T}{his} chapter attempts to clarify the current status of
TRTD in international law. In order to establish whether TRTD has any
binding legal force in international law, it must be established if it
falls into any of the categories of the sources of international law
as stipulated by the International Court of Justice (ICJ). Article 38
of the Statute of the Court reveals five distinct sources;

\begin{enumerate}

\item International conventions (or treaties)
\item International custom
\item General principles of law
\item Judicial decisions
\item Writings of publicists
\end{enumerate}

This chapter provides a brief introduction on international treaties
and focuses on the legal status of TRTD in the African Charter on
Human and Peoples' Rights. It then introduces the notion of customary
international law and the elements required to prove that a norm has
been accepted as customary international law. The chapter deals with
the legal status of General Assembly Resolutions and international
conferences and then finally analyses the TRTD as customary
international law.

\section{International Treaties}

The fundamental character of international law-making treaties is
consent. It is the mutual consent of each of the parties to the treaty
which creates its binding force. International human rights treaties,
also fundamentally based on consent, have one vital difference. As a
legislature or constitution does not exist within the international
community, many states perform their legal obligations in 'good
faith'. In most treaties, the reciprocal obligations and duties of
each state party makes it easier to confront a state who is not
performing their legal obligations. For example, Sieghart compares
most international treaties as ``commercial contracts, involving some
exchange from which both (or all) the state parties will
benefit''. Sieghart explains that when formulating international human
rights treaties ``there are no such commercial incentives or
sanctions... Although it is the governments of states which enter
these treaties, the trouble is that the beneficiaries are not those
governments but their subjects, who are not themselves parties to the
treaties''.

Form this light, it seems that many states may enter a human rights
treaty without necessarily having to fulfill their obligations in
'good faith'. Although they may be legally obliged to do so, other
parties may not have the incentive to sanction a 'rogue'
state. However, to assume that other incentives are not present will
be to deny the value of human rights treaties.

For the purpose of this section, the provisions in international
treaties which directly refer to TRTD will be discussed here.

\subsection{The Right to Development and The African Charter on Human and Peoples' Rights}

The African Charter on Human and Peoples' Rights, also known as the
Banjul Charter, was adopted by the Organisation of African Unity (OAU)
in 1981. It entered into force in 1986, and is, to date, the only
treaty which recognises TRTD. As a regional treaty, it only binds
members of the OAU who have ratified it. To understand the substance
and the content of TRTD enshrined in the Banjul Charter, it would be
beneficial to provide a brief account on the influential regional
factors which were present at the time of its adoption. It can be
noted that although many discussions on TRTD acknowledge the legal
recognition given to the right in the Banjul Charter, many
commentators fail to provide any analysis on the concept~--- except to
mention that TRTD is a very controversial subject at the international
level. This is very unfortunate as an analysis on the very concept of
TRTD contained in the Banjul Charter could, by proxy, clarify the
substance of the concept at the international level.

The OAU was created in 1963 as a regional agreement between African
States. Although the primary focus of the treaty was not on human
rights, it reaffirmed the principles contained in the UN Charter and
the UDHR. Significant attention was placed on the right to
self-determination and the eradication of colonialism. It also
referred to the welfare and the well being of African people. The OAU
actively supported liberation movements in Mozambique and Angola, and
was committed to the achievement of self-determination and human
rights in South Africa and Namibia. However, the fundamental principle
which underlined the OAU was that of state sovereignty and
non-interference, due to the effects of colonialism and the
establishment of new independent states. Principles such as respect
for human rights, paramount to the international law of human rights,
were, to a large extent, deemed as domestic concerns of a state. To
this end, Umozurikestates that:

\begin{quote}
[T]he massacres of thousands of Hutu in Burundi in 1972 and 1973
were neither discussed nor condemned by the OAU, which regarded them
as matters of internal affairs. The notorious regimes of Idi Amin of
Uganda (1971-1979), Marcias Nguema of Equatorial Guinea (1969-1979),
and Jean-Bedel Bokassa of the Central African Republic (1966-1979)
escaped the criticism of the OAU and most of its members, Tanzania,
Zambia, and Mozambique being the exceptions.
\end{quote}

According to Umozurike, the OAU regarded matters relating to human
rights as the domestic concern of states fundamentally because the
principle of state sovereignty enshrined in the OAU charter was
``unduly emphasised'' and as a consequence the state parties were able
to maintain an ``indifferent attitude to the suppression of human
rights''. Thus although on the international plane the protection of
human rights was accepted as a concern for the whole international
community, alongside the adoption of other regional agreements,
African states seemed a long distance away from achieving respect for
human rights in their territories when their primary concern was to
establish themselves as independent nation states.

The Banjul Charter was adopted with the principle of respect for human
rights as a focal point. It included several provisions on CPR's as
well as ESCR's. It recognised the indivisibility of all
rights. Furthermore, it emphasised certain rights as belonging to the
'peoples'. However, the principle of state sovereignty was still as
paramount and Slimane suggests that:

\begin{quote}
The African Charter on Human and Peoples' Rights could not but reflect
the fundamental objectives of state sovereignty and territorial
integrity of each state and integrity of borders declared in the 1963
Organisation of African Unity (OAU) Charter, which was basically
concerned with relations between states. As such, it is safe to say
that it was not the intent of the drafters of the 1981 Charter to
equate the term 'peoples' with the notion of minorities or ethnic
groups. Rather, the concept of 'peoples' was identified with the
African nation-state.
\end{quote}

Thus once again the notion of a state right to development
emerges. Noting that the principle of non-intervention in the domestic
affairs of a member state ``had been the bane of the continent'',
Umozurike argues that the adoption of the Banjul Charter ``was
arguably the single most important development in the field of human
rights in Africa''. He implies that the principle of non-intervention
had to some extent, been superseded by that of the principle of
respect for human rights, with the adoption of the Banjul
Charter. Thus, although critics may argue that the Banjul Charter was
only ``basically concerned with relations between states'', the
fundamental nature of the Banjul Charter, with all its faults, still
attempts to build a bridge between the legitimate principle of state
sovereignty~--- the bedrock of international law~--- and the principle
of respect for human rights, needed to protect African people from
their own state leaders.

The Banjul Charter, unlike its regional counterparts~--- the European
Convention and the American Convention~--- did not create a regional
Court. Instead, the African Commission on Human and Peoples' Rights
was established, with a mandate to promote and protect the rights
enshrined in the Banjul Charter and also to interpret its provisions.

The Banjul Charter recognises TRTD twice, once in its preamble and
then in Article 22. The preamble states:

\begin{quote}
that it is henceforth essential to pay a particular attention to the
right to development and that civil and political rights cannot be
dissociated from Economic, social and cultural rights in their
conception as well as universality and that the satisfaction of
Economic, social and cultural rights ia (\textit{sic}) a guarantee for
the enjoyment of civil and political rights;
\end{quote}

This paragraph of the preamble unequivocally reinforces the
indivisibility and universality of all human rights. Moreover, it
expressly identifies that the enjoyment of ESCR's guarantees the
enjoyment of CPR's. The meaning of this sentiment has often been
criticised as it could suggest that ESC rights are a priority and thus
CPR could be curtailed in the achievement of them. However, this point
has been taken out of context from the whole paragraph. It is clear
that this paragraph, read as a whole, denotes that all human rights
are connected and are associated with each other.

Nevertheless, the substance of TRTD is unclear from the preamble; the
only clear aspect, is that the Banjul Charter recognises the existence
of a 'right to development'. To elaborate on the substance of TRTD,
Article 22, states that:

\begin{quote}
1. All peoples shall have the right to their economic, social and
cultural development with due regard to their freedom and identity and
in the equal enjoyment of the common heritage of mankind. 2. States
shall have the duty, individually or collectively, to ensure the
exercise of the right to development.
\end{quote}

The formulation of TRTD in Article 22(2) identifies that states have
the duty to realise TRTD both individually and collectively. This is
in line with TDRTD which also expresses that states have this duty. It
is unclear what the term ``all peoples'' denotes. It could once again
be construed in number of different ways. The identification of what
the term `peoples' means in the Banjul Charter is further complicated
by the effects of colonialism and the subsequent need of States to
establish themselves as an independent nation. This is illustrated
clearly by Morel, who describes that:

\begin{quote}
African States have traditionally viewed the recognition of the
distinct identities of minorities as an element that poses a threat to
national unity and undermines the objective of nation building.
\end{quote}

Thus, taking into account the history of the continent and of its
people, it is evident that the term `peoples' contained in the charter
could display elements which can be identified with the objective of
nation building. This supports the notion that TRTD enshrined in the
Banjul Charter is a right of states.

In order to clarify the term `peoples' and recognise that the term
also included groups of people within state, the African Commission
stated that:

\begin{quote}
Although the terms \textit{peoples rights} have not been defined in
the Charter, these rights generally refer to the rights of a community
(be it ethnic or national) to determine how they should be governed,
how their economies and cultures should develop; they include other
rights such as the right to national and international peace and
security, the right to a clean and satisfactory environment.
\end{quote}

Furthermore, the African Commission took the opportunity in the case
of \textit{Social and Economic Rights Action Center and the Center for
  Economic and Social Rights v Nigeria}, to reaffirm the term
'peoples' in the charter included ethnic minorities and it referred to
'peoples' as groups of peoples within a state. This case reinforced
that 'peoples' had the collective right, under article 24, ``to a
general satisfactory environment favorable to their development''.
This case also highlights the importance of the principle of
participation. The African Commission stated ``In all their dealings
with the oil consortiums, the Government did not involve the Ogoni
communities in the decisions that affected the development of
Ogoniland''. This is in accordance to the principle of participation
contained in TDRTD.

The addition of the text ``be it ethnic or national'' stated by the
African Commission above, suggests that the term peoples used in the
context of the charter can be also be identified with the right of the
national community, which is equated with the state. Accordingly, TRTD
contained in the Banjul Charter is also a right of states as well as a
right of peoples as groups within the state. Thus, there are dangers
of states curtailing the rights of people within their
jurisdiction. Furthermore, the absence of 'individuals' in Article 22,
suggests that TRTD is only a rights of 'peoples' and not also of
individuals. This element is in clear contravention with TDRTD which
states unequivocally that TRTD is a right of individuals and
peoples. Furthermore, TDRTD clearly stipulates that the individual is
central to development, whereas such a notion is notably absent in the
Banjul Charter. Thus, the concept of TRTD enshrined in the Banjul
Charter contradicts the concept of TRTD in TDRTD. The right which has
been postivised in the Banjul Charter seems, in essence, to be a right
of peoples and states to development, excluding the individual, which
can have serious consequences for the achievement of all human rights,
whether individual or collective rights.  However, if article 22 is
read alongside article 2 of the charter, the individual is entitled to
enjoy all the rights guaranteed by the Banjul Charter. Article 2
reads:

\begin{quote}
Every individual shall be entitled to the enjoyment of the rights and
freedoms recognised and guaranteed in the present Charter without
distinction of any kind such as race, ethnic group, colour, sex,
language, religion, political or any other opinion, national and
social origin, fortune, birth or any status.
\end{quote}

Thus the principle of non-discrimination contained in article 2, used
in tandem with article 22, indicates that TRTD in the Banjul charter
is a right of individuals as well as a right of collectives and of
states~--- however controversial that concept may be. It is evident in
the jurisprudence of the African Commission that the Articles in the
Charter can be used together. For example, Article 2 was used in
conjunction with other articles to protect minority rights in the case
of \textit{Malawi African Association etc v Mauritania}. Also, the
case concerning the Ogoni Community identifies that the rights
contained in the Banjul Charter are individual rights as well, it then
becomes clear that the `peoples' rights contained in the Charter can
also be individual rights.

The African Commission has not yet come to hear a case concerning
Article 22, but one of the key NGO's working in the field of minority
rights, Minority Rights Group International (MRG) have initiated a
communication to the African Commission alleging a violation of
article 22 of the Banjul Charter~--- alongside other articles~--- of
the Endorois Community in Kenya. The Endorois community will be the
first to claim their TRTD. As Morel notes:

\begin{quote}
While the right to development has not been formally adjudicated as
such by any human rights judicial or quasi-judicial body\ldots [T]he
African Commission made clear in the Ogoni communication that there
was no right in under the African Charter that was not able to be made
effective. The conclusion can therefore be drawn that the right to
development can be given practical and effective meaning in Africa.
\end{quote}

Thus, the African Commission will have the opportunity to give
substance to the concept of TRTD contained in the Banjul Charter,
which could also serve to clarify the concept at the international
level. It can therefore be noted that although the legal force of TRTD
and the duty to cooperate in the Banjul Charter concerns only the
African State parties who have ratified it, its clarification within
this area could serve as a significant step to uncover the substance
of the right at the international level. From the analysis above, it
can be noted that the concept of TRTD enshrined in the Banjul Charter
differs from that contained in TDRTD. The Banjul Charter recognises
the right primarily as a right of peoples and states, which is
fundamentally different from the concept at the international level,
as the latter underscores the individual. This point is crucial in
understanding why there is a political deadlock concerning TRTD at the
international level. However, most principles in both concepts are the
same, including the importance given to the principle of
participation, the principle of non-discrimination, among others. From
this we can extrapolate that the concept of TRTD in the Banjul Charter
and the one at the international level can be reconciled to some
extent.

\subsection{Customary International Law}

Despite customary law being identified as one of the principal sources
of international law, one is faced with serious difficulties when
attempting to establish a rule of customary law for a number of
reasons. By virtue of its very nature, customary law is created
through a very informal law-making process, which in turn reveals that
the precision and clarity required through other 'formal' law-making
methods will not necessarily be evident when attempting to discover
the existence of a customary rule.

International customary law comprises two elements, state practice and
the belief by states that such practice is carried out because it is
the law~--- otherwise known as the \textit{opinio juris sive
  necessitates (opinio juris)}. State practice can be evidenced by
both actions and abstentions from certain practices. A non-exhaustive
list of material which could be used as evidence of state practice
includes, \textit{inter alia}, actions of governments, statements made
by governmental persons to their respective Parliaments, or to foreign
governments, or in intergovernmental conferences. Custom was described
as ``a constant and uniform usage practiced by states''in the
\textit{Asylum Case: Columbia v Peru (1950) ICJ Rep 266}. Thus in
order to establish a customary law, state practice must be uniform and
consistent. The extent of uniformity of state practice required and
the duration of time that would be sufficient to be classified as
`\textit{constant and uniform}' was not considered in this case, thus
subsequent cases determined that there was no requirement for
excessive repetition of the state practice concerned nor was there a
requirement for absolute rigorous conformity of a certain practice for
it to be deemed as a customary rule; it would be sufficient for the
practice to be generally consistent. The duration of state practice
was discussed in the \textit{North Sea Continental Shelf Cases}, where
the ICJ indicated that:

\begin{quote}
Although the passage of only a short period of time is not
necessarily, or of itself, a bar to the formation of a new rule of
customary international law on the basis of what was originally a
purely conventional rule, an indispensable requirement would be that
within the period in question, short though it may be, state practice,
including that of states whose interests are specially affected,
should have been both extensive and virtually uniform in the sense of
the provision invoked~--- and should moreover have occurred in such a
way as to show a general recognition that a rule of law or legal
obligation is involved.
\end{quote}

Furthermore, where there are significant inconsistencies between
different State practices, the rule will not be deemed as customary
international law. The ICJ in the \textit{Fisheries Case} noted that
although a ten-mile limit for bays had

\begin{quote}
been adopted by certain States both in their national law and in
their treaties and conventions , and although certain arbitral
decisions have applied it as between these states, other States have
adopted a different limit. Consequently, the ten-mile rule has not
acquired the authority of a general rule of international law.
\end{quote}

The need for consistency in state practice was further underscored in
the Asylum Case, which although concerned a regional customary rule,
serves to clarify the rules on consistent state practice. The ICJ held
in this case;

\begin{quote}
The facts brought to the knowledge of the Court disclose so much
uncertainty and contradiction, so much fluctuation and discrepancy in
the exercise of diplomatic asylum and in official views expressed on
various occasions, there has been so much inconsistency in the rapid
succession of conventions on asylum, ratified by some States and
rejected by others, and the practice has been much influenced by
considerations of political expediency in the various cases, that it
is not possible to discern in all this any constant and uniform usage,
accepted as law\ldots
\end{quote}

State practice alone will not suffice to establish a customary law as
in order to distinguish binding legal norms with those of mere comity,
states must also demonstrate that they were obliged by law to behave
in such a manner, in other words, states have to accept such practice
as a legal obligation and believe they are obliged, by law to conform
with its requirements. The ICJ has indicated that the \textit{opinio
  juris} of states must be strictly proven, however in the
\textit{Nicaragua Case} the court changed its stance and accepted that
\textit{opinio juris} may be inferred.

\subsection{Generality of the Practice}

The recognition of a particular customary rule by the majority of
states infers that the custom is generally recognised as a binding
international norm. It will be legally binding on states generally as
long as the state practice is extensive and representative, although
an individual state may not be bound if they have persistently
objected to the rule. State practice does not need to be universal. A
customary international rule can be formed if the state practice has
only been followed by a small number of states, providing that those
states are significantly representative in the field where the norm is
to be created and that there is no significant dissent. Such a
principle was highlighted in the \textit{North Sea Continental Shelf
  Cases} where the ICJ held that providing the state practice included
participation by ``states whose interests are specially affected'',
the requirement of representative state practice would be
fulfilled. The International Law Association (ILA) further elaborates
upon this principle and suggests that;

\begin{quote}
The Criterion of representativeness has in fact a dual aspect~---
negative and positive. The positive aspect is that, if all major
interests (``specially affected States'') are represented, it is not
essential for a majority of States to have participated (still less a
great majority, or all of them). The negative aspect is that if
important actors do not accept the practice, it cannot mature into a
rule of general customary law.
\end{quote}

From this, it can be established that if a state or a group of
states~--- who are important in the field in which the norm is to be
created~--- significantly dissent and do not accept the rule, they can
stop the rule from becoming customary international law as it would
not conform to adequate representation required in the state
practice. According to the ILA, only one important state in the field
can by its opposition, stop the rule from being formed in customary
international law, as the rule would not conform to the requirement of
sufficiently representative participation of the practice.

\subsection{The Persistent Objector}

The principle that an individual state may not be bound if they have
persistently objected to a rule has been long established. It was
confirmed in the \textit{Anglo-Norwegian Fisheries case}. The ICJ
stated that even if the rule had acquired the status of customary
international law ``[i]n any event the \ldots rule would appear to be
inapplicable as against Norway, in as much as she has always opposed
any attempt to apply it''.  The State must object to the rule in
public and do so throughout the emergence of the rule in order to be a
persistent objector.

\subsection{General Assembly Resolutions and International Conferences}

There has often been confusion surrounding the nature of GA
resolutions and their legal effects. Some writers have suggested that
GA resolutions are a direct source of international law, either
categorising them into international treaties or attempting to
establish that they are customary international law. The United
Nations General Assembly (GA) is not an established law-making body
and do not have the legal authority to bind states. However, the
declarations and resolutions of the GA may have some legal effect to
the extent that they have been used as evidence of state practice
towards establishing emerging customary law or the content of an
existing customary rule.

In the \textit{Nicaragua case}, the International Court of Justice
(ICJ) accepted that Article 2(4) of the United Nations Charter,
regarding the prohibition of the use of force, with other instruments
such as the Declaration on Principles of International Law Concerning
Friendly Relations and Co-operation among States (Resolution 2625) had
generated a rule of customary law, which is similar to the content of
Article 2(4), but which exists alongside treaty provisions stating the
same principle. The \textit{Nicaragua case} highlighted that GA
resolutions may be used as evidence of state practice . This can be
seen by the use of the GA's Friendly Relations Resolution 2625 by the
ICJ in the \textit{Nicaragua case} as support for its ruling in
customary international law. The ICJ has used this resolution
alongside others, to establish the existence of a customary rule and
the \textit{opinio} \textit{juris} of the states. By looking at the
circumstances, conditions and attitude of the States at the time the
resolution was adopted, the content of the declaration was enough to
satisfy the court that this was a significant statement which was
reflective of the development of customary law alongside treaty
law. The ICJ held that the participating states acceptance of
resolution 2625, had confirmed the existence of an \textit{opinio
  juris} in customary law of the prohibition of the use of force. The
effect of consent to such GA resolutions ``may be understood as
acceptance of the validity of the rule or set of rules declared by the
resolutions themselves''.

As the GA consists of representatives throughout the world, its
resolutions and declarations are of particular value when attempting
to establish state practice which could possibly lead to binding
customary international law. The ICJ in the \textit{Legality of the
  Threat or Use of Nuclear Weapons advisory} \textit{opinion} clearly
articulated this point:

\begin{quote}
The Court notes that General Assembly resolutions, even if they are
not binding, may sometimes have normative value. They can, in certain
circumstances, provide evidence important for establishing the
existence of a rule or the emergence of an \textit{opinio juris}. To
establish whether this is true of a General Assembly resolution, it is
necessary to look at its content and the conditions of its adoption;
it is also necessary to see whether an \textit{opinio juris} exists as
to its normative character. Or a series of resolutions may show the
gradual evolution of the \textit{opinio juris} required for the
establishment of a new rule.
\end{quote}

Moreover, when a GA resolution or declaration refers to subjects which
are contained in the UN Charter, they can be seen as an authoritative
interpretation of the Charter itself.

Resolutions and declarations arising from international conferences of
a universal character, can also in the same manner as GA resolutions
and declarations, contribute to the evidence of state practice of a
new customary rule or can be used as evidence of the content of
existing customary law.

\subsection{The Right to Development in Customary International Law}

As has been established above, in order for TRTD to gain the status of
customary international law, state practice must be demonstrated. Such
practice can be evidenced in the form of resolutions, declarations and
international conferences. I will first analyse declarations,
resolutions and conferences, which explicitly recognise TRTD to
establish if it has gained the status of an international customary
norm, which would legally bind the international community to realise
the right. To analyse these instruments, McDougal recognises that a
number of factors must be considered. He suggests:

\begin{quote}
In order to decide whether a UN statement reflected an accurate
description of what peoples' expectations were concerning the law, one
needed to know several facts: Who voted for the statement? Who voted
against it? What was the relative and effective power of these votes?
How compatible is the asserted policy with past expectations? What
followed from the resolution? What were the expectations coming from
other sources? And so on.
\end{quote}

The implicit recognition of TRTD in instruments such as the UN
charter, UDHR, ICCPR and ICESR are surrounded by controversy. Some
writers have suggested that the elements which are contained in these
instruments and in TDRTD~--- for example the elements concerning
international cooperation~--- legally obligate the international
community, although others suggest that this is not so. Such
controversies will be discussed toward the end of this chapter.

TRTD was first explicitly recognised in the Commission of Human Rights
Resolution 4 (XXXIII) 1977. This resolution recognised TRTD as a human
right and asked the UN Secretary General to conduct a study on the
``international dimensions of the right to development as a human
right in relation with other human rights based on international
co-operation, including the right to peace, taking into account the
requirements of the NIEO and the fundamental human needs''. What is
notable in this resolution is the fact that it immediately expresses
TRTD as a human right without any questions. Thus it would imply that
such a concept has been discussed to see whether or not TRTD is a
human right. However, it can be noted that although other statements
have contained some of the principles of TRTD, the concept itself was
not discussed within the international community. Thus the explicit
recognition by the CHR of TRTD as a human right raises questions as to
how the right was elevated to this status. Evidence of state practice
toward the recognition of the rule could include the following
resolutions, although each must be read in light of the circumstances
in which they arose.

\textbf{General Assembly Resolution 1161 (XII) 1957} contains some of
the principles which are inherent to TRTD but it does not mention TRTD
explicitly. This would imply that TRTD has not yet come into existence
although the foundations for the emergence of the right are being
made. As this resolution does not explicitly mention TRTD, it can not
be used as evidence of state practice toward the emergence of TRTD,
\textit{per se}. However, as noted, it can contribute as evidence of
the emergence of certain principles contained in TRTD.

The \textbf{GA Resolution on Permanent Sovereignty over Natural
  Resources, 1962}, asserts that states have the right to control
their natural resources, including their development. In the
\textbf{UN Conference on Trade and Development (UNCTAD I) 1964}, more
equitable distribution of the world's resources was demanded by
developing countries.  Neither of these directly recognise TRTD so it
cannot be assumed that they are evidence of state practice although
the principles which resulted from these are relevant to the emergence
of TRTD and can be taken as evidence of that. Both the resolution and
the conference underline the principles of State control over natural
resources and self-determination. Such principles are evident in
TDRTD, for example in Article 1(2). Thus both the resolution and the
conference can be used as evidence that certain principles which make
up TRTD were emerging, although it cannot be said that TRTD itself had
been agreed.

The \textbf{Tehran International Conference on Human Rights 1968}
inextricably linked human rights with the development discourse,
without specifically mentioning TRTD. The Conference emphasised that,
``[t]he widening gap between the economically developed and developing
countries impedes the realisation of human rights in the international
community''. It further stated: ``The achievement of lasting progress
in the implementation of human rights is dependent upon sound and
effective national and international policies of economic and social
development''. Such notions were reinforced by the \textbf{Declaration
  on Social Progress and Development 1969}, which recognised that
certain conditions for social progress and development were necessary.

Both the Tehran Conference and the Declaration on Social Progress and
Development reflect the essence of TRTD without explicitly mentioning
it and can thus be used as evidence of the emergence of the concept
rather than as evidence of the right itself as only specific elements
are mentioned and not all.  This can also be said for the
\textbf{Charter of Economic Rights and Duties of States 1974}, which
provided a framework for the establishment of the NIEO, envisaging
more equality between nations and expressed several elements of
TRTD. In 1972 at the \textbf{UN Conference for Trade and Development
  (UNCTAD III)}, in Chile, TRTD was claimed by many of the developing
nations. Although developing countries explicitly recognised TRTD,
such recognition could be said as not representative of the whole
international community and thus it could not constitute toward
evidence of state practice. The\textbf{ Declaration on Race and Racial
  Prejudice} adopted in 1978 by UNESCO referred to ``the right of
every human being and group to full development''. The GA
\textbf{Declaration on the Preparation of Societies for Life in Peace
  1978} also stated that everyone has the right ``to determine the
road of their development''. All of these reveal that the concept of
TRTD was emerging but still did not exist and had not been endorsed by
the international community.

All the above resolutions, declarations and the outcomes of the
different international conferences serve to highlight specific
elements of TRTD and provide evidence of consensus for those
particular aspects. The fact that TRTD had not been expressly
recognised reveals that TRTD, as a concept in itself, had not emerged,
thus casting doubts over the validity of the above mentioned
instruments as evidence of state practice for the whole TRTD, although
they can be accepted as evidence of certain principles. Furthermore,
as TRTD had not been specifically addressed, it could not possibly be
argued that states had the necessary \textit{opinio juris} to be
legally bound by the right. One can also criticise the CHR on its
formulation of TRTD in \textbf{Resolution 4 (XXXIII) 1977}, as they
expressly state that TRTD is a human right without conducting any
prior investigation as to this aspect. \textbf{Resolution 4 (XXXIII)
  1977} assumes that this aspect is certain and then asks the SG to
conduct a study on the obstacles which the right may come across. It
was evident that the study was not to see whether the right existed;
this was implicitly recognised in the resolution. Such a presumption
has ultimately contributed to the controversies surrounding TRTD as a
human right, and the CHR should have undertaken a study on this aspect
before raising TRTD to the status of a human right.

By 1979, having considered the suggestions made by the Secretary
General, the CHR issued a further resolution recognising TRTD. Whilst
a working group was established to prepare a draft declaration on
TRTD, \textbf{The Banjul Charter}gave TRTD legal recognition for the
first time. Such activities suggest that TRTD had already been
established rather than considering it as an emerging concept, without
consistent, uniform or extensive state practice to reflect the views
of the international community at large.

In 1986 TDRTD was adopted by the GA with 146 votes for the
declaration, 8 abstentions and 1 negative vote. At first sight it
would seem that such consensus on the declaration would contribute to
evidence of state practice in favour of recognising TRTD as an
emerging rule, as the majority of states voted for the
declaration. The fact that TRTD at this point was still a relatively
new concept, would not itself bar the right from becoming an emerging
international customary norm. However, the key to understanding
whether or not TDRTD can be used as evidence of state practice comes
from analysing the negative vote and the abstentions.

The USA cast the only negative vote. State practice must include
participation of those states who are likely to be the most affected
by the new rule. It can be taken for granted that the USA is a state
whose ``interests are specially affected'' and thus its negative vote
reveals that participation in the state practice is not representative
of the whole international community, concluding that, at this stage,
TRTD could not evolve into a rule of general international customary
law. Without even analysing the abstentions, TDRTD received a
significant blow with the negative vote cast by the USA. For this
reason, it can not be used as evidence of state practice as it
effectively reveals the contrary~--- that of discontent.

The Working Group of governmental experts on the right to development
continued to clarify the content of TRTD and although consensus had
not been reached regarding certain aspects of the right, it was
endorsed at the \textbf{Rio Conference on Environment and Development
  1992} in the \textbf{Rio Declaration on Environment and Development}
(Rio Declaration). This was the first occasion in which TRTD was fully
affirmed by the international community. At first sight, it can be
used as evidence of state practice, which is uniform and
representative of the whole international community as it was adopted
unanimously. The Rio Declaration is a significant statement of the
general rights and obligations of states which affect the environment
and constitutes a statement of existing customary rules and newly
formulated principles. It has, however, been described as somewhat of
a `\textit{package deal}', balancing the interests of developing and
developed states. This can be seen throughout the declaration as some
provisions reflect the interests of the developed states, while others
reflect those of the developing countries. TRTD was strongly supported
by the developing states, and Principle 3 stipulated that; ``The right
to development must be fulfilled so as to equitably meet developmental
and environmental needs of present and future generations''. Principle
4 was supported by developed states and what is striking about the two
principles is their relationship to the principle of sustainable
development. As Birnie and Boyle noted:

\begin{quote}
One illustration of the declarations package-deal character is the
conjunction of Principles 3 and 4, which together form the core of the
principle of sustainable development.
\end{quote}

The Rio Declaration not only linked the development discourse with
human rights but also linked these with environmental protection. This
is of considerable importance as such links had not been previously
acknowledged by the international community.

Birnie and Boyle criticise TRTD on the grounds that it does not employ
any language appropriate to environmental protection or sustainability
and state that ``[t]his is a classical illustration of the UN's
inability to do joined up thinking''. This is a valid comment as TRTD,
had up until Rio, concentrated on integrating the development
discourse and the human rights discourse and did not specifically
relate to environmental protection and sustainable development.
Nevertheless, the affirmation of TRTD in the Rio Declaration is of
extreme significance as it was adopted unanimously and expresses the
will of the international community. A shadow of doubt can be cast
over such unanimity as evidence of state practice as the USA made
several reservations to some of the provisions in the declaration,
which included a reservation on Principle 3. However, Birnie and Boyle
state that despite these reservations, the principles and rules
contained in the Rio Declaration ``have a universal significance and
cannot be dismissed as the work of one segment of international
society'' and can thus be used as evidence of state practice toward
the establishment of TRTD as an emerging customary rule, whereby the
international community would be obliged to realise the right.
However as state practice must be consistent, more evidence is
required to conclude this. Further evidence of state practice could
include GA resolution 47/123 which reaffirmed TRTD as a human right
and recalled the principles of the Rio Declaration.

The \textbf{Vienna World Conference on Human Rights 1993} resulted in
the adoption of the Vienna Declaration and Programme of Action (Vienna
Declaration). This dealt with TRTD extensively, reaffirming it as a
``universal and inalienable human right'' as well as recognising it as
``an integral part of fundamental human rights''. Sengupta argues that
the affirmation of TRTD by consensus in the Vienna Declaration settles
the debate surrounding whether or not TRTD is a human right, to the
affirmative. While Sengupta may believe this to be true, I would argue
that the debate is far from settled.  The Vienna declaration
recognised the universality, indivisibility and interdependency of all
human rights which was also significant in TDRTD. It emphasised the
human person is the central subject of development and underscored
that:

\begin{quote}
While development facilitates the enjoyment of all human rights, the
lack of development may not be invoked to justify the abridgement of
internationally recognised human rights.
\end{quote}

This element was added to the Declaration on behalf of states who
feared that development would be used as an excuse by repressive
governments to limit other human rights. Boven has suggested that the
above statement questions the notion of indivisibility and
interdependency of all human rights, which the Vienna Declaration
notably promotes as it ``conveys a dual notion of human rights''. He
elaborates stating that:

\begin{quote}
`With ``all human rights'' the whole range of human rights is
  obviously included but it appears from the context of the statement
  that the term `internationally recognised human rights' basically
  relates to civil and political rights only.
\end{quote}

Such an analysis of the text does question what the Vienna Declaration
meant by using the phrase `\textit{internationally recognised human
  rights'} rather than just using `\textit{human rights'} which would
undeniably denote indivisibility and interdependency of all human
rights. Nevertheless, the Vienna Declaration also recognised that
democracy, development and respect for human rights and fundamental
freedoms are not only interdependent but also mutually reinforcing. It
also recognised, ``[r]espect for human rights and for fundamental
freedoms without distinction of any kind is a fundamental rule of
international human rights law''. This reaffirms the principle of
non-discrimination contained in previous human rights instruments and
TDRTD as fundamental and arguably legally binding in itself.

The Vienna Declaration clearly stipulates that extreme poverty affects
the capabilities of people to enjoy their human rights. Paragraph 14
unequivocally states that:

\begin{quote}
The existence of widespread extreme poverty inhibits the full and
effective enjoyment of human rights; its immediate alleviation and
eventual elimination must remain a high priority for the international
community.
\end{quote}

Thus although TDRTD did not expressly mention extreme poverty,
references made in the Vienna Declaration expressly link poverty and
human rights together and in turn as it affirms TRTD as a human right,
exteme poverty affects the realisation of TRTD.

The declaration linked environmental protection with the TRTD by
including; ``[t]he right to development should be fulfilled so as to
meet equitably the developmental and environmental needs of present
and future generations''. Thus once again the notion that the
environment, development and human rights are interrelated emerged, as
it had done so in the Rio Declaration.  The Vienna Declaration~---
similar to TDRTD~--- stipulates that effective international
cooperation should be promoted by the international community and it
emphasises that states should cooperate with each other to ensure
development and eliminate the obstacles to development. However, the
Vienna Declaration went beyond TDRTD, linking the external debt
problems of states to the effective realisation of ESCR's. The Vienna
Declaration

\begin{quote}
[C]alls upon the international community to make all efforts to help
alleviate the external debt burden of developing countries, in order
to supplement the efforts of the Governments of such countries to
attain the full realisation of the Economic, social and cultural
rights of their people.
\end{quote}

This paragraph can be interpreted to imply that states external debt
problems are hindering the realisation of all ESCR's, which in turn
would affect the realisation of TRTD. Thus, if it is found that the
international community is legally obligated to realise TRTD, the
issue of alleviating external debt burdens, could fall under this
obligation.

As consensus was reached by the international community on the Vienna
Declaration, it could be used as evidence of state practice toward
establishing TRTD as a customary rule. Furthermore, the obligation to
realise the right by the international community, through effective
policies, elimination of obstacles and an appropriate international
environment which is conducive to the realisation of TRTD seems to be
strongly supported by the Vienna Declaration. This could be used to
suggest that the obligation for international cooperation for the
realisation of TRTD was emerging as a customary norm, although the
Vienna Declaration is not legally binding in itself. However, the
provisions of the declaration relating to cooperation only infer that
states `should' cooperate, implying that it is a goal rather than a
legal obligation. This casts doubts over the relevance of the Vienna
Declaration as evidence of state practice on the emergence of TRTD,
including the element of international cooperation, as legally binding
customary rules. Moreover, what is notable about the Vienna
Declaration is that although it expressly reaffirms TRTD, it does not
attempt to define development or contribute to the clarification of
this aspect of TRTD.  The Vienna Declaration and its provisions on
TRTD have been readily endorsed in subsequent GA resolutions, offering
some evidence of state practice towards TRTD being accepted by the
international community as binding customary law - or at least reveals
the way in which international law is developing on TRTD.

The 1990's was a decade in which a number of international conferences
endorsed TRTD as a human right. These are as follows;

\begin{itemize}
\item International Conference on Population and Development, Cairo, 1994
\item World Summit for Social Development, Copenhagen, 1995
\item The Platform for Action 1995 Fourth World Conference on Women
\item The World Food Summit 1996
\item Second UN Conference on Human Settlements (Habitat II) 1996
\end{itemize}

Although they did not deal with TRTD as extensively as the Vienna
Declaration, some elements of each conference are relevant to
realising TRTD and can be used as evidence of state
practice. Furthermore, GA resolution 50/184affirmed this notion and
expressed that each of these conferences contained aspects which are
relevant to the realisation of TRTD. The Office of the High
Commissioner on Human Rights (OHCHR) concludes that:

\begin{quote}
The collective message of all the UN summits and conferences of the
1990's may be summed up as a call for greater recognition of human
rights in development.
\end{quote}

GA resolution 53/155is of considerable importance as it reveals the
discontent still present regarding the realisation of TRTD. The
resolution expresses concerns over the implementation of the right and
states that:

\begin{quote}
[O]bstacles to the realisation of the right to development still
persist at both national and international levels, that new obstacles
to the rights stated therein have emerged and that the progress made
in removing these obstacles remains precarious.
\end{quote}

It further expresses concern over the insufficient dissemination of
TDRTD and states that TDRTD ``should be taken into account, as
appropriate, in bilateral and multilateral cooperation programmes,
national development strategies and policies and activities of
international organisations''. Nevertheless, this resolution affirms
TRTD as stipulated in TDRTD.

By 1998, TRTD had been endorsed in numerous international conferences,
GA resolutions and in resolutions of the CHR. Taken from this alone,
TRTD would seem to be emerging as a customary international
norm. However, the key to understanding the legal status of TRTD, is
in an assessment of the concerns which states have expressed, these
demonstrate that the opinio juris of states to be legally bound to
realise the TRTD, is not present. Some of these will now be discussed.

In the view of the Government of Switzerland the implementation and
realisation of TRTD had been hindered because of the controversies
surrounding the very nature of TRTD and thus;

``In order to facilitate its implementation, certain controversial
points, such as the very concept of the right to development, need to
be clarified... Despite many years of discussion and debate, there is
still uncertainty over the Declaration's content.... The only
certainty is that ``the human person is the central subject of
development''.

Such a statement is a grave blow towards the recognition of TRTD as
emerging customary international law. It identifies that rather than
clarifying TRTD, the events which have taken place since TDRTD have
contributed very little to solving the problems inherent in the
right. At this point, it would seem very relevant that Switzerland
identifies that the very concept of TRTD is still very controversial
as although the right has been endorsed on many occasions, none of
these have elaborated on the actual concept. What seems to have
happened is that one declaration or conference relies upon another as
evidence. In turn, this falls back on to TDRTD, which is fundamentally
the source of the controversies because of it's ambiguity. These
notions are further expressed by Ghana. Underlying the need to
elaborate on the content of TRTD, the Government of Ghana notes that
the definition of development needs clarification as well as ``The
expressions ``individually'' and ``collectively'' should be better
defined so as to distinguish individual and collective rights, as well
as States' rights''. At first sight it would seem that this was a
straight forward criticism concerning the uncertainty surrounding TRTD
in general. However, read more closely, it is evident that even at
this stage, the fact that Ghana expressly mentions `\textit{State's
  rights}' , reveals that the controversies surrounding TRTD are far
from being solved. In fact it could be said that the efforts made
during the 1990's to recognise TRTD were made in vain and were merely
rhetoric as the Government of India noted$^{.}$ While taking this
account, it seems impossible to conclude~--- at this point~--- that
there is uniform and consistent state practice alongside
\textit{opinio juris}, which are necessary to establish international
customary law.

Although the resolutions and declarations on TRTD are not completely
invalid as the ``mere recognition of a rule and the conditions for its
execution in a resolution give it the beginning of legal force''.
From the above analysis it seems certain that TRTD has not yet
received the status of `hard law' or customary international
law. However, Gutto argues that as TDRTD directly links to the UN
Charter and can be seen as an authoritative interpretation of the
Charter itself and confidently asserts:

\begin{quote}
It can therefore be safely stated that General Assembly resolutions,
especially those that directly link to the letter and spirit of the
Charter of the United Nations, like resolution 41/28, have some
appreciable legal authority that ``bind'' States. The fact that even
countries that had abstained from voting for the resolution in 1986,
like Sweden, can today openly align themselves with the instrument and
attempt to integrate its core elements into the national development
policies, plans, and practices is very instructive.
\end{quote}

In principle, what Gutto asserts is correct and theoretically,
resolutions and declarations of the GA can elucidate and develop
customary law. However, his assertion is far from revealing the
truth. Although he notes the case of Sweden, he fails to address the
views of states like Switzerland, Ghana, the USA and India, which were
discussed above. Taking all these views into account and the
controversies surrounding the right, TRTD is not sufficiently clear or
precise to denote any form of a legal obligation.

Despite the relative discontent surrounding TRTD, it was subsequently
endorsed in the \textbf{Millennium Declaration 2000}, the
\textbf{World Conference against Racism, Racial Discrimination,
  Xenophobia and Related Intolerance 2001} and the \textbf{World
  Summit on Sustainable Development 2002} (WSSD) held in
Johannesburg. Furthermore, the GA issued an annual resolution on the
right. By 2003, some states regarded TRTD as an inalienable human
right but, as before, confusion surrounding the right was ever
present. Japan confirmed its view that the concept of TRTD was not
clarified and that it ``disagreed with the legal obligation of
developed countries to render assistance to developing
countries''. Furthermore, to Japan TRTD is an individual right and
could not be a collective right of any sort.

In opposing resolution 58/172 the US representative stated that the
resolution implies that ``that lack of development justifies the
denial of internationally recognised human rights''. This statement is
of crucial significance as the Vienna Declaration~--- which is
recalled in this resolution~--- explicitly states the
opposite. Furthermore, in line with the thoughts of Boven, it infers
that the USA do not recognise ESC's as human rights as they have not
ratified the ICESCR; and, to them, ``internationally recognised human
rights'' are only CPR's. This is just one more factor which undermines
TRTD as the USA do not recognise ESC rights as human rights at all.

Furthermore the USA disagree with the resolution's call to ``prepare a
concept document on a legally binding instrument on the Right to
Development'' and is of the opinion that ``such an effort is
unwarranted in view of the fact that any such legally binding
instrument is unlikely to ever garner significant support''. While
hesitating to agree with this, the controversies surrounding the right
reveal that it may not be possible to create a document which contains
legally binding obligations on TRTD.

In concluding their views, the USA representative states that;

``We cannot support the call to make progress on realising the Right
to Development. There is no internationally accepted definition of
such a right. Making such a call is premature and irrelevant.''

Consequently, there is significant evidence that states have not
agreed to be legally obligated to realising TRTD.

\section{The Right to Development in other Legal Instruments}

Many commentators have suggested that TRTD derived from several key
legal instruments~--- including the UN Charter and the International
Bill of Rights, amongst others ~--- and the legal obligation to
realise the right stems from these. While there is some merit in this
line of reasoning, such arguments have now lost their force.

At the time of its emergence, explicit recognition of the right, was
very minimal. It seems that, to address this problem, some
commentators have attempted to base the right in well established
documents. These efforts have, in most cases, augmented the
controversy and, after close scrutiny of TRTD, it would be redundant
to repeat the exercise. Whilst it is evident that some of the
principles of TRTD can be found in well endorsed legal instruments,
the whole right itself simply cannot. Nevertheless, it may be
worthwhile to briefly mention how this approach can be applied to the
UN Charter.

The preamble of the charter asserts that the Peoples of the UN are
determined ``to promote social progress and better standards of life
in larger freedom''. This has led to the claim that ``[t]he
relationship between the right to development and other human rights
had already been stated in the preamble of the United Nations
Charter''. M'Baye argued that TRTD ``is already enscribed
(\textit{sic}) in filigree in the Charter of the United Nation''. The
core aspects of TRTD echo one of the main purposes of the UN. From
Articles 1(3), Article 55 and 56, it can be implied that international
cooperation is a legal obligation to realise human rights, but
international cooperation as a legal obligation to realise TRTD cannot
be deduced from these. A number of provisions in the International
Bill of Rights have been used in these discussions as implicit
evidence of TRTD.

\chapter{Conclusion}

\lettrine{I}{t is} impossible to answer the question ``Is there a
Legal Obligation for the International Community to Realise the Right
to Development?''  in the affirmative. This is because there is no
agreement on the definition of TRTD. There is no agreement by the
international community on what kind of right TRTD is; whether it is a
human right, legal right, collective right, right of states, synthesis
of rights etc. This results in confusion on where the right is to be
placed~--- within the framework of human rights or within
international law as a whole. As TRTD is unclear and ambiguous, states
are able to interpret the right according to their own agenda. This
results in TRTD meaning different things, to different states, at
different times.

While the OEWG and the IE have concentrated their work on implementing
TRTD, the conceptual difficulties surrounding the right hinders its
effective realisation. It then becomes clear that the OEWG and IE have
steered the debate away from the conceptual difficulties towards that
of implementation. However, without clarifying the nature, scope and
definition of TRTD, it cannot be realised effectively. TRTD has been
built upon incomplete foundations and when the layers are stripped
off, it is evident that it is conceptually flawed, therefore the aim
of realising TRTD is a fallacy. I recommend that the international
community, the UN GA and the Commission on Human Rights, focus on the
conceptual problems surrounding TRTD once again, clarifying the
definition and content of the right. Furthermore, the relationship
between sustainable development and TRTD should be analysed. This
could serve to clarify some aspects of TRTD.

\begin{thebibliography}{1}

\bibitem{Abi- Saab G.} Abi- Saab G. {\em The Legal Formulation of the
  Right to Development} in R. Dupuy (ed), The Right to Development at
  the International Level, Netherlands: Sijthoff and Noordhoff, 1981,
  p163.

\bibitem{Aguda T.} Aguda T. {\em Human Rights and the Right to Development in
  Africa}, Lagos: Nigerian Institute of International Affairs, 1989.

\bibitem{Alston P. 1987} Alston P. and Quinn G., {\em The Nature and
  Scope of States Parties Obligations under the ICESCR}, 9 Hum Rts, Q
  122, 1987

\bibitem{Alston P. 1982} Alston P. {\em A third generation of
  solidarity rights: progressive development or obfuscation of
  international human rights law?} 29 Netherlands Int'l L Rev 307,
  1982

\bibitem{Alston P. 1988} Alston P. {\em Making Space for New Human
  Rights : the case of the right to development}, Harvard Human Rights
  Yearbook; V.1; pp.3-40, 1988

\bibitem {Alston P. 1991} Alston P.,{\em Revitalising United Nations
  Work on Human Rights and Development}, 18 MELB U. L. Rev. 216-219,
  1991

\bibitem {Alston P. 1981} Alston P., {\em The Right to Development at
  the International Level} in R. Dupuy (ed), The Right to Development
  at the International Level, Netherlands: Sijthoff and Noordhoff,
  1981

\bibitem {Alston P. 1985} Alston P., {\em The Shortcomings of a
  ``Garfield the Cat'' Approach to the Right to Development},
  California Western International Law Journal, Vol 15 (1985), 510.

\bibitem {Alston P. 1998} Alston P., {\em What is in a Name: Does it
  really matter if development refers to goals, ideals of human
  rights?} in H. Helmich (Ed.), Human Rights in Development
  Cooperation, SIM Special No 22 pp.95-106, Netherlands Institute for
  Human Rights, Utrecht, 1998.

\bibitem {Alston P. 1997} Alston P., {\em The Myopia of the
  Herdmaidens: International Lawyers and Globalisation}, 8 European
  Journal of International Law (1997) 435

\bibitem {Alston P. 2001} Alston P., (ed), {\em Peoples' Rights},
  Oxford: Oxford University Press, 2001.

\bibitem {Aristotle 1992} Aristotle., {\em The politics.} Rev ed
  (translated by TA Sinclair, revised and re-presented by TJ
  Saunders). Harmondsworth: Penguin, 1992.

\bibitem {Baehr P R. 2001} Baehr P R., {\em Human rights universality
  in practice }.Basingstoke: Palgrave, 2001.

\bibitem {Barsh R. 1991} Barsh R., {\em The Right to Development as a
  Human Right: Results of the Global Consultation}, Human Rights
  Quarterly, Vol 13 (1991), 321.

\bibitem {Bentham J. 1970} Bentham J.,{\em Of laws in general }(edited
  by HLA Hart). Athlone Press, 1970

\bibitem {Bentham J. 1988} Bentham J., {\em The principles of morals
  and legislation}. Prometheus, 1988

\bibitem Birnie P., and Boyle A., {\em International Law \& The
  Environment}, 2$^{nd}$ ed, Oxford: Oxford University Press, 2002.

\bibitem {Boven TV. } Boven TV.,{\em Human Rights and Rights of
  Peoples}, 6 EJIL (1995) 1-476,

\bibitem {Brownlie I. } Brownlie I., {\em Basic documents on human
  rights}. 4th ed, Oxford, Oxford UP, 2002.

\bibitem {Brownlie I. } Brownlie I., {\em Principles of public
  international law}, 5th ed, Oxford, OUP, 1998.

\bibitem {Bunn I D. } Bunn I D., '{\em The Right to Development:
  Implications for International Economic Law'}, 15 American
  University International Law Review 1425 (2000): 1425-67

\bibitem {Center for Development and Human Rights. } Center for
  Development and Human Rights., {\em The Right to Development: A
    Primer}, India: Sage Publishers, 2004

\bibitem {Chapman A.}, Chapman A. {\em A Violations Approach for
  Monitoring the International Covenant on Economic Social and
  Cultural rights}, 18 Hum Rts Q 23, 1996

\bibitem {Charlesworth H. } Charlesworth H., {\em 'The Public/Private
  Distinction and the Right to Development in International Law',
  Australian Year Book of International Law}, Vol 12 (1992), 190-204.

\bibitem {Chowdhury R. } Chowdhury R., Denters E. and de Waart
  P. (eds), {\em The Right to Development in International Law},
  Dordrecht: Martinus Nijhoff, 1992.

\bibitem {Craven M. } Craven M., {\em The International Covenant on
  Economic, Social and Cultural Rights: a perspective on its
  development}. Oxford, Clarendon Press, 1995

\bibitem {De Feyter K. } De Feyter K., {\em World development law :
  Sharing Responsibility for development}, Insentia, Belgium, 2001

\bibitem {Dixon M. } Dixon M., {\em Cases and materials on
  international law}, 3$^{rd}$ ed, London, Blackstone, 2000.

\bibitem {Donnelly J. } Donnelly J., {\em Human Rights as Natural
  Rights}, 4 Hum Rts Q 391, 1982

\bibitem {Donnelly J. } Donnelly J., {\em Human Rights, Democracy and
  Development}, Human Rights Quarterly; V.21(3); pp.608-633\\

\bibitem {Donnelly J. } Donnelly J., {\em In Search of the Unicorn :
  The Jurisprudence and Politics of the Right to Development}, 15 Cal
  W Int'l L J 473, 1985

\bibitem {Donnelly J. } Donnelly J., {\em Universal Human Rights in
  Theory and Practice}, Ithaca: Cornell University Press, 1989

\bibitem {Douzinas C. } Douzinas C., {\em The end of human rights
  critical legal thought at the turn of the century}. Oxford: Hart,
  2000

\bibitem {Dreze J. } Dreze J., {\em The political economy of hunger
  selected essays}, Oxford: Clarendon Press, 1994

\bibitem {Dupuy R. } Dupuy R., (ed), {\em The Right to Development at
  the International Level}, Netherlands: Sijthoff and Noordhoff, 1981

\bibitem {Eide A. } Eide A., Krause C., Rosas A., {\em 'Economic,
  Social and Cultural Rights: A Textbook'}, The Netherlands: Martinus
  Nijhoff Publishers, 1995.

\bibitem {Frankovits A. } Frankovits A., '{\em Rejoinder:The Rights
  Way to Development}', Food Policy 21 no 1, 123-28.

\bibitem {Galenkamp M. } Galenkamp M., {\em Collective Rights, }Report
  to the Advisory committee on Human Rights and Foreign Policy of the
  Netherlands, Rotterdam, Netherlands: 1998 available at
  http://www.uu.nl/content/16-3.pdf

\bibitem {Ghai Y. } Ghai Y., '{\em Human Rights and Governance: The
  Asia Debate'}, 15 Australian Year Book of International Law (1994)
  1.

\bibitem {Gibney M. } Gibney M., Skogly S., {\em Transnational Human
  Rights Obligations}, Human Rights Quarterly; V.24; pp.781

\bibitem {Goodwin-Gill G S. } Goodwin-Gill G S., {\em The reality of
  international law Essays in honour of Ian Brownlie}, Oxford : Oxford
  UP, 1999.

\bibitem {Hamm B. } Hamm B., {\em A human rights approach to
  development}, Human Rights Quarterly; V.23(4); pp.1005-10\\

\bibitem {Harris DJ. } Harris DJ., {\em Cases and materials on
  international law.} 5th ed, London: Sweet and Maxwell, 1998

\bibitem {Hausserman J. } Hausserman J., {\em The Realization and
  Implementation of Economic Social and Cultural Rights}, London:
  Macmillan, 1992

\bibitem {Haussermann J. } Haussermann J., {\em Rights and humanity: a
  human rights approach to development. }Department for International
  Development, London. 1998

\bibitem {Hobbes T. } Hobbes T., {\em Leviathan} (edited by JCA
  Gaskin).Oxford: OUP, 1996.

\bibitem {Howse R. } Howse R., {\em Maintreaming the Right to
  Development into International Trade Law and Policy at the World
  Trade Organization}, UN Doc. E/CN.4/Sub.2/2004/17.\\

\bibitem {Hurrell A } Hurrell A, {\em Inequality, globalization, and
  world politic}, Oxford : OUP, 1999.

\bibitem International Law Association, London Conference (2000), {\em
  Final Report of the Committee on Formation of Customary (General)
  International Law}, Available at www.ila-hq.org.

\bibitem {Kant I. } Kant I., {\em Groundwork of the metaphysics of
  morals} (translated and edited by M Gregor) Cambridge, CUP, 1998.

\bibitem {Locke J. } Locke J., {\em Two treatises of government}
  (edited by P Laslett. 3rd ed). Cambridge, CUP, 1988.

\bibitem {Marks S. } Marks S., {\em `The Human Right to Development:
  Between Rhetoric and Reality'}, 17 Harvard Human Rights Journal 137,
  2004

\bibitem {Marks S. } Marks S., {\em Emerging human rights : a new
  generation fo the 1980's?} 33 Rutgers L Rev 435, 451, 1981

\bibitem {Marks S. } Marks S., {\em The Human Rights Approach to
  Development: 5 Approaches}, FXB Centre for Health and Human Rights,
  Working Paper 5
  (2000). \textcolor{blue}{{\uline{http://www.hsph.harvard.edu/fxbcenter/FXBC\_WP6--}}}Marks.pdf.
  Last accessed 11.10.04

\bibitem {Marx K. } Marx K., {\em Capital a critique of political
  economy}. Volume 1, Book 1 - the process of production of capital. -
  Lawrence and Wishart, 1954

\bibitem {Marx K. } Marx K., {\em The Communist manifesto }(by K Marx
  and F Engels with introduction and notes by AJP Taylor). - London :
  Penguin, 1967

\bibitem {Middleton N. } Middleton N., {\em Rio plus ten, poverty and
  environment}, London: Pluto Press, 2003

\bibitem {Mill JS. } Mill JS., {\em Utilitarianism} (edited by R
  Crisp). - Oxford : OUP, 1998.

\bibitem {Mill JS. } Mill JS., {\em Utilitarianism; On liberty; Essay
  on Bentham; together with selected writings of Jeremy Bentham and
  John Austin} (edited with an introduction by M Warnock).Fontana,
  1962

\bibitem {Morel C. } Morel C., {\em Defending Human Rights in Africa:
  The Case for Minority and Indigenous Rights}, Essex Human Rights
  Review. Vol 1. No.1 available at
  http://projects.essex.ac.uk/EHRR/archive/pdf/49.pdf

\bibitem {Mowbray A. } Mowbray A., {\em Cases and Materials on the
  European Convention on Human Rights}, London: Butterworths, 2001.

\bibitem {Nanda V. } Nanda V., Shepherd G. and McCarthy-Arnolds
  E. (eds), {\em World Debt and the Human Condition: Structural
    Adjustment and the Right to Development}, Westport, CT: Greenwood
  Press, 1993.

\bibitem {Nuscheler. } Nuscheler., {\em The international debate on
  human rights and the right to development,} 1998, {\em The right to
  development: advance or greek gift in the development of human
  rights?}
  \textcolor{blue}{{\uline{http://www.inwent.org/E+Z/content/archive-eng/01-2004/tribune\_art1.htm}}}. Last
  accessed 15.8.04

\bibitem {Obiora Amede L. } Obiora Amede L.,{\em Beyond the Rhetoric
  of a Right to Development}, Law and Policy; V.18(3/4); pp.355-418

\bibitem {O'Brien J. } O'Brien J., {\em International law}, London:
  Cavendish, 2001

\bibitem {OHCHR. } OHCHR., {\em Human Rights and Poverty Reduction: A
  Conceptual Framework}, United Nations,
  2004. \textcolor{blue}{{\uline{http://www.ohchr.org/english/issues/poverty/}}}. Last
  accessed 25.10.04

\bibitem {Orford A. } Orford A., '{\em Globalisation and the Right to
  Development}', in Alston P. (ed), Peoples' Rights, Oxford: Oxford
  University Press, 2001., 127-185

\bibitem {Overseas Development Institute. } Overseas Development
  Institute., {\em What Can We Do with a Rights-Based Approach to
    Development?} Overseas Development Institute Briefing Paper 1999
  \textcolor{blue}{{\uline{http://www.odi.org.uk/publications/briefing/3\_99.html}}}
  Last accessed 25.10.04

\bibitem {Paine T. } Paine T., {\em Rights of man, Common sense, and
  other political writings} (edited with an introduction by Mark
  Philip). Oxford: OUP, 1995

\bibitem {Paul. } Paul., {\em The Human Right to Development its
  Meaning and Importance}, The John Marshall Law Review; V.25; pp.235

\bibitem {Piron L-H. } Piron L-H., {\em The Right to Development: A
  Review of the Current State of the Debate}, (2002). Prepared for the
  UK Department for International
  Development. \textcolor{blue}{{\uline{http://www.odi.org.uk/pppg/publications/papers\_reports/dfid/issues/rights01/index.html}}}. Last
  accessed 26.10.04

\bibitem {Piron L-H. } Piron L-H., {\em The Right to Development:
  Study on Existing Bilateral and Multilateral Programmes and Policies
  for Development Partnership}, UN Doc. E/CN.4/Sub.2/2004/15.

\bibitem {Plato. } Plato., {\em The republic}. 2nd ed (translated by D
  Lee). London: Penguin, 1970.

\bibitem {Rehman J. } Rehman J., {\em International human rights law a
  practical approach}, Harlow: Pearson Education, 2003

\bibitem {Rich R. } Rich R., '{\em The Right to Development: A Right
  of Peoples?}', in J. Crawford (ed), The Rights of Peoples, Oxford:
  Clarendon Press, 1988, 39-54.

\bibitem {Rousseau J-J. } Rousseau J-J., {\em The discourses and other
  early political writings} (edited by V Gourevitch).Cambridge:CUP,
  1997.

\bibitem {Rousseau J-J. } Rousseau J-J., {\em The social contract
}(translated with introduction and notes by Christopher Betts) Oxford:
  OUP, 1994

\bibitem {Salomon M E. } Salomon M E., and Sengupta A., {\em The Right
  to Development: Obligations of States and the Rights of Minorities
  and Indigenous Peoples}, Minority Rights Group International, 2003.

\bibitem {Santos Pais M. } Santos Pais M., {\em A Human Rights
  Conceptual Framework for UNICEF}, UNICEF, Innocenti Essay No. 9,
  Florence, 1999.Avaliable at
  \textcolor{blue}{{\uline{http://www.unicef-icdc.org/publications/pdf/essay9.pdf}}}. Last
  accessed 17.10.04

\bibitem {Sen A. } Sen A., {\em Democracy as a Universal Value},
  Journal of Democracy; V.10(3); pp.3-17 \\

\bibitem {Sen A. } Sen A., '{\em The Right not to be Hungry'}, in
  Alston P., and Tomasevski K., (eds), The Right to Food, Netherlands:
  SIM, 1984.

\bibitem {Sen A. } Sen A., {\em Development as Freedom}, Oxford
  University Press: Oxford, 1999

\bibitem {Sengupta A. } Sengupta A., {\em Theory and Practice on the
  Right to Development}, Human Rights Quarterly; V.24(4); pp.837-889

\bibitem {Sengupta A. } Sengupta A., {\em Towards Realizing the Right
  to Development}, Development and Change; June; V.31(3) \\

\bibitem {Sengupta A. } Sengupta A., {\em The Right to Development as
  Human Right}, 2000, Harvard School of Public Health, available at
  www.hsph.harvard.edu/fxbcenter/FXBC\_WP7--Sengupta.pdf

\bibitem {Sieghart P. } Sieghart P., {\em The Lawful Rights of
  Mankind}, Oxford: Oxford University Press,1985.

\bibitem {Sieghart P. } Sieghart P., {\em The International Law of
  Human Rights}, Oxford: Oxford University Press, 1985

\bibitem {Shaw M N. } Shaw M N., {\em International law}. 4$^{th}$
  ed. Cambridge: Cambridge UP, 1997

\bibitem {Shaw M N. } Shaw M N., {\em International law}. 5$^{th}$
  ed. Cambridge: Cambridge UP, 2003

\bibitem {Shelton D. } Shelton D., 'Environmental Rights', in Alston
  P. (ed), Peoples' Rights, Oxford: Oxford University Press,
  2001.189-256.

\bibitem {Slimane S. } Slimane S., '{\em Recognising Minorities in
  Africa}', Minority Rights Group International 2003, May
  2003. available at www.minorityrights.org

\bibitem {Sohn. } Sohn., {\em The international law of human rights: a
  reply to recent criticisms} 9 Hofstra L Rev 347, 1981

\bibitem {Steiner H J. } Steiner H J., and Alston P., {\em
  International human rights in context: law, politics, morals},
  Oxford: Oxford UP, 1996

\bibitem {Steiner H J. } Steiner H J., and Alston P., {\em
  International human rights in context: law, politics, morals},
  2$^{nd}$ ed, Oxford :Oxford UP, 2000

\bibitem {Symonides J (ed). } Symonides J (ed)., {\em Human rights
  concepts and standards}, Aldershot, Ashgate, Paris: UNESCO, 2000.

\bibitem {Symposium} {\em Symposium: Development as an Emerging Human
  Right}', California Western International Law Journal, (1985), 429.

\bibitem {Hammurabi} {\em The Code of Hummurabi} (1792-1750 BC); A
  translation of the Code of Hammurabi can be found at:
  http://www.wsu.edu:8080/~dee/MESO/CODE.HTM (last accessed 24.09.04)

\bibitem The Human Rights Council of Australia., {\em The Rights Way
  to Development: A Human Rights Approach to Development Assistance},
  The Human Rights Council of Australia, Australia, 2001,
  http://www.hrca.org.au/symposium.htm and
  http://www.hrca.org.au/The\_Rights\_Way\_to\_Development\_Manual.htm.
  Last accessed 26.10.04

\bibitem {{\em The Limburg Principles on the Implementation of the
    International Covenant on Economic Social and Cultural Rights} }
  {\em The Limburg Principles on the Implementation of the
    International Covenant on Economic Social and Cultural Rights}, 9
  Hum Rts Q 122, 1987

\bibitem {{\em The Maastrict Guidelines on Violations of Economic
    Social and Cultural Rights } {\em The Maastrict Guidelines on
    Violations of Economic Social and Cultural Rights,} 20 Hum Rts Q
  691, (1998).

\bibitem {Udombana N. } Udombana N., '{\em The Third World and the
  Right to Development: Agenda for the Next Millennium'}, Human Rights
  Quarterly, Vol 22(3) (2000), 753-87.

\bibitem {Umozurike U O. } Umozurike U O., {\em The African Charter on
  Human and Peoples' Rights}, 77 Am. J. Int'l L. 902, (1983).

\bibitem {United Nations. } }United Nations., {\em The realization of
  the right to development: Global consultation on the right to
  development as a human right : report prepared by the
  Secretary-General pursuant to Commission on Human Rights resolution
  1989/45,} United Nations, 1991.

\bibitem Uvin P., {\em On High Moral Ground: The Incorporation of
  Human Rights by the Development Enterprise}, Praxis- The Fletcher
  Journal of Development Studies, Volume XVII 2002, available at
  http://fletcher.tufts.edu/praxis/xvii/Uvin.pdf

\bibitem {Uvin P. } Uvin P., Human Rights and Development, USA:
  Kumarian Press Inc; 2004

\bibitem {Vasak K. } Vasak K., and Alston P (eds)., {\em The
  International Dimensions of Human Rights}, Westport, CT: Greenwood
  Press, Volume 1, 1982.

\bibitem {Watson. } Watson.,{\em The limited utility of international
  law in the protection of human rights,} 74 Am Soc'y Int'l L.Proc,
  1981

\bibitem Official Documents of the United Nations Available at:
  http://ap.ohchr.org/documents/dpage\_e.aspx?s=103

\bibitem {UN CHR } UN CHR , {\em Draft report of the Commission}, 60th
  session,13/04/2004, UN Doc.E/CN.4/2004/L.10/Add.7

\bibitem {UN CHR } UN CHR , {\em Summary Record of the 18th Meeting}
  60$^{th}$ session, 30/03/2004, UN Doc. E/CN.4/2004/SR.18

\bibitem {UN CHR } UN CHR, {\em Report of the Working Group on the
  Right to Development on its fifth session}, 60$^{th}$ session,
  22/04/2004 UN Doc. E/CN.4/2004/23/Corr.1

\bibitem {UN CHR } UN CHR, {\em Report of the Working Group on the
  Right to Development on its fifth session}, 60th session 16/03/2004,
  UN Doc.E/CN.4/2004/23

\bibitem {UN CHR } UN CHR, {\em Written statement submitted by the
  Hariri Foundation - The Islamic Foundation for Culture and Higher
  Education}, 60$^{th}$ session, 11/03/2004, UN
  Doc.E/CN.4/2004/NGO/226

\bibitem {UN CHR } UN CHR, {\em Country studies on the right to
  development - Argentina, Chile and Brazil}, 60$^{th}$ session,
  23/01/2004 UN Doc. E/CN.4/2004/WG.18/3

\bibitem {UN CHR } UN CHR, {\em Draft report of the Commission},
  59$^{th}$ session, 25/04/2003 UN Doc. E/CN.4/2003/L.10/Add.7

\bibitem {UN CHR } UN CHR, {\em Draft report of the Commission,}
  59$^{th}$ session, 25/04/2003 UN Doc. E/CN.4/2003/L.11/Add.8

\bibitem {UN CHR } UN CHR, {\em Exposicin escrita presentada por el
  Movimiento Cubano por la Paz y la Soberana de los Pueblos},
  60$^{th}$ session,11/03/2004, UN Doc.E/CN.4/2004/NGO/222

\bibitem {UN CHR } UN CHR, {\em Exposicin escrita presentada por el
  Movimiento Cubano por la Paz y la Soberana de los Pueblos},
  60$^{th}$ session, 11/03/2004, UN Doc.E/CN.4/2004/NGO/221

\bibitem {UN CHR } UN CHR, {\em Exposicin escrita presentada por el
  Movimiento Indio "Tupaj Amaru"}, 60$^{th}$ session, 11/03/2004, UN
  Doc.E/CN.4/2004/NGO/199

\bibitem {UN CHR } UN CHR, {\em Fifth report of the independent expert
  on the right to development}, 59$^{th}$ session, 16/09/2002 UN
  Doc. E/CN.4/2002/WG.18/6

\bibitem {UN CHR } UN CHR, {\em Fifth report of the independent expert
  on the right to development}, 59$^{th}$ session, 30/12/2002 UN
  Doc. E/CN.4/2002/WG.18/6/Add.1

\bibitem {UN CHR } UN CHR, {\em Forum on economic, social and cultural
  rights: the Social Forum}, 55$^{th}$ session, 27/04/1999 UN
  Doc. E/CN.4/RES/1999/53

\bibitem {UN CHR } UN CHR, {\em Fourth report of the independent
  expert on the right to development~--- Mission}, 58$^{th}$ session,
  05/03/2002 UN Doc. E/CN.4/2002/WG.18/2/Add.1

\bibitem {UN CHR } UN CHR, {\em Fourth report of the independent
  expert on the right to development}, 58$^{th}$ session, 20/12/2001
  UN Doc. E/CN.4/2002/WG.18/2

\bibitem {UN CHR } UN CHR, {\em Globalization and its impact on the
  full enjoyment of all human rights}, 57$^{th}$ session, 20/04/2001
  UN Doc. E/CN.4/RES/2001/32

\bibitem {UN CHR } UN CHR, {\em Globalization and its impact on the
  full enjoyment of human rights}, 59$^{th}$ session, 11/04/2003 UN
  Doc. E/CN.4/2003/L.25

\bibitem {UN CHR } UN CHR, {\em High-level seminar on the right to
  development: Global partnership for development}, 60$^{th}$ session,
  23/03/2004 UN Doc.E/CN.4/2004/23/Add.1

\bibitem {UN CHR } UN CHR, {\em Human rights and unilateral coercive
  measures}, 55$^{th}$ session, 23/04/1999 UN Doc. E/CN.4/RES/1999/21

\bibitem {UN CHR } UN CHR, {\em Interim report of the High
  Commissioner for Human Rights}, 55$^{th}$ session, 20/08/1999 UN
  Doc. E/CN.4/1999/WG.18/3

\bibitem {UN CHR } UN CHR, {\em Note by the secretariat: open-ended
  working group on the right to development}, 59$^{th}$ session,
  17/12/2002 UN Doc. E/CN.4/2003/WG.18/3

\bibitem {UN CHR } UN CHR, {\em Preliminary study of the independent
  expert on the right to development on the impact of international
  economic and financial issues on the enjoyment of human rights},
  59$^{th}$ session, 27/01/2003 UN Doc. E/CN.4/2003/WG.18/2

\bibitem {UN CHR } UN CHR, {\em Proceedings of the working group on
  the right to development}: Note by the secretariat, 56$^{th}$
  session, 08/03/2000 UN Doc. E/CN.4/2000/21

\bibitem {UN CHR } UN CHR, {\em Provisional Agenda - Open-ended
  working group on the right to development,} 57$^{th}$ session,
  18/12/2000 UN Doc. E/CN.4/2001/WG.18/1

\bibitem {UN CHR } UN CHR,{\em Provisional agenda of the Working Group
  on the Right to Development}, 56$^{th}$ session, 15/02/2000 UN
  Doc. E/CN.4/2000/WG.18/1

\bibitem {UN CHR } UN CHR,{\em Provisional agenda}, 60$^{th}$ session,
  08/12/2003 UN Doc. E/CN.4/2004/WG.18/1

\bibitem {UN CHR } UN CHR, {\em Report of the High Commissioner for
  Human Rights: the Right to Development}, 59$^{th}$ session,
  12/09/2002 UN Doc. E/CN.4/2003/7

\bibitem {UN CHR } UN CHR, {\em Report of the Open-ended working group
  on the right to development on its fourth session}, 59$^{th}$
  session, 24/03/2003 UN Doc. E/CN.4/2003/26

\bibitem {UN CHR } UN CHR, {\em Report of the Open-ended working group
  on the right to development on its fourth session}, 59$^{th}$
  session, 24/03/2003 UN Doc. E/CN.4/2003/26

\bibitem {UN CHR } UN CHR, {\em Report of the open-ended Working Group
  on the Right to Development on its third session}, 58$^{th}$
  session, 11/04/2002 UN Doc. E/CN.4/2002/28/Rev.1

\bibitem {UN CHR } UN CHR, {\em Report of the Working Group on the
  Right to Development on its fifth session }60$^{th}$ session{\em ,}
  16/03/2004 UN Doc. E/CN.4/2004/23

\bibitem {UN CHR } UN CHR, {\em Report of the Working Group on the
  Right to Development on its fifth session,} 60$^{th}$ session,
  22/04/2004 UN Doc. E/CN.4/2004/23/Corr.1

\bibitem {UN CHR } UN CHR,{\em Review of progress and obstacles in the
  promotion, implementation, operationalization, and enjoyment of the
  right to development}, 60$^{th}$ session, 17/02/2004, UN
  Doc. E/CN.4/2004/WG.18/2

\bibitem {UN CHR } UN CHR, {\em Right to development - Aide-mmoire
  prepared by the Secretariat}, 57$^{th}$ session, 18/12/2000 UN
  Doc. E/CN.4/2000/WG.18/CRP.5/Rev.1

\bibitem {UN CHR } UN CHR, {\em Right to Development - Report of the
  Open-Ended Working Group}, 57$^{th}$ session, 20/03/2001 UN
  Doc. E/CN.4/2001/26

\bibitem {UN CHR } UN CHR, {\em Study on the current state of progress
  in the implementation of the right to development}, 55$^{th}$
  session, 27/07/1999 UN Doc. E/CN.4/1999/WG.18/2

\bibitem {UN CHR } UN CHR, T{\em he importance and application of the
  principle of equity, at both the national and international levels:
  Report submitted by the High Commissioner}, 59$^{th}$ session,
  30/12/2002, UN Doc. E/CN.4/2003/25

\bibitem {UN CHR } UN CHR, {\em The Right to Development - Report of
  the High Commissioner for Human Rights}, 57$^{th}$ session,
  12/09/2000 UN Doc. E/CN.4/2000/WG.18/CRP.2

\bibitem {UN CHR } UN CHR, {\em The right to development - Report of
  the High Commissioner}, 60$^{th}$ session, 08/01/2004 UN
  Doc. E/CN.4/2004/22

\bibitem {UN CHR } UN CHR, {\em The right to development - Third
  report of the independent exper}t, 57$^{th}$ session, 02/01/2001 UN
  Doc. E/CN.4/2001/WG.18/2

\bibitem {UN CHR } UN CHR, {\em The right to development,} 54th
  session, 22/04/1998 UN Doc. E/CN.4/RES/1998/72

\bibitem {UN CHR } UN CHR, {\em The right to development}, 55$^{th}$
  session, 28/04/1999 UN Doc. E/CN.4/RES/1999/79

\bibitem {UN CHR } UN CHR, {\em The right to development}, 56$^{th}$
  session, 13/04/2000 UN Doc. E/CN.4/RES/2000/5

\bibitem {UN CHR } UN CHR, {\em The right to development,} 57$^{th}$
  session, 18/04/2001 UN Doc. E/CN.4/RES/2001/9

\bibitem {UN CHR } UN CHR, {\em The right to development}, 58$^{th}$
  session, 25/04/2002 UN Doc. E/CN.4/RES/2002/69

\bibitem {UN CHR } UN CHR, {\em The right to development}, 59$^{th}$
  session, 08/04/2003 UN Doc. E/CN.4/2003/L.14

\bibitem {UN CHR } UN CHR, {\em The right to development,} 59$^{th}$
  session, 08/04/2003 UN Doc. E/CN.4/2003/L.14

\bibitem {UN CHR } UN CHR,{\em The right to development,} 59$^{th}$
  session, 25/04/2003 UN Doc. E/CN.4/RES/2003/83

\bibitem {UN CHR } UN CHR, {\em The right to development}, 60$^{th}$
  session, 13/04/2004, UN Doc. E/CN.4/RES/2004/7

\bibitem {UN CHR } UN CHR, {\em The right to development}, 60$^{th}$
  session, 21/01/2004 UN Doc. E/CN.4/2004/116

\bibitem {UN CHR } UN CHR, {\em Working group on the right to
  development: Information submitted by the Centre Europe-}
\bibitem {{\em Tiers Monde} } {\em Tiers Monde}, 56$^{th}$ session,
  23/11/1999 UN Doc. E/CN.4/2000/WG.18/CRP.3

\bibitem {UN CHR } UN CHR, {\em Written statement submitted by
  International League for Human Rights}, 59$^{th}$ session 15/03/2003
  UN Doc. E/CN.4/2003/NGO/163

\bibitem {UN CHR } UN CHR, {\em Written statement submitted by the
  International Federation of Rural Adult Catholic }{\em Movements
  (FIMARC)}, 60$^{th}$ session, 18/03/2004, UN Doc.E/CN.4/2004/NGO/257

\bibitem {UN CHR } UN CHR, {\em Written submission by the World Health
  Organization (WHO)}, 60$^{th}$ session,12/03/2004, UN
  Doc.E/CN.4/2004/120

\bibitem {UN ECOSOC } UN ECOSOC, {\em The right to developmen}t, 1995,
  25/07/1995 UN Doc. E/DEC/1995/258

\bibitem {UN ECOSOC } UN ECOSOC,{\em The right to development,} 1998,
  30/07/1998 UN Doc. E/DEC/1998/269

\bibitem {UN ECOSOC } UN ECOSOC, {\em The right to development}, 2000,
  28/07/2000 UN Doc. E/DEC/2000/246

\bibitem {UN ECOSOC } UN ECOSOC, {\em The right to development}, 2002,
  25/07/2002 UN Doc. E/DEC/2002/271

\bibitem {UN GA } UN GA, {\em Globalization and its impact on the full
  enjoyment of all human rights}, 58$^{th}$ session, 07/08/2003 UN
  Doc. A/58/257

\bibitem {UN GA } UN GA, {\em The right to development}, 56$^{th}$
  session, 08/02/2002 UN Doc. A/RES/56/150

\bibitem {UN GA } UN GA, {\em United Nations Millennium Declaration},
  55$^{th}$ session, 23/09/2002 UN Doc. A/RES/55/2

\bibitem {UN SUBCOM } UN SUBCOM, {\em Promoting the right to
  development in the context of the United Nations Decade for the
  Elimination of Poverty (1997-2006)}, 56$^{th}$ session, 01/06/2004,
  UN Doc.E/CN.4/Sub.2/2004/13

\bibitem {UN SUBCOM } UN SUBCOM, {\em Mainstreaming the right to
  development into international trade law and policy at the World
  Trade Organization}, 56$^{th}$ session, 09/06/2004, UN
  Doc.E/CN.4/Sub.2/2004/17

\bibitem {UN SUBCOM } UN SUBCOM, {\em Promotion of the realization of
  the right to drinking water and sanitation}, 55$^{th}$ session,
  07/08/2003 UN doc. E/CN.4/Sub.2/2003/L.17

\bibitem {UN SUBCOM } UN SUBCOM, {\em The legal nature of the right to
  development and enhancement of its binding nature}, 56$^{th}$
  session, 01/06/2004, UN Doc.E/CN.4/Sub.2/2004/16

\bibitem {UN SUBCOM } UN SUBCOM, {\em The right to development,}
  55$^{th}$ session, 06/08/2003 UN Doc. E/CN.4/Sub.2/2003/L.7

\bibitem {UN SUBCOM } UN SUBCOM, {\em The right to development},
  56$^{th}$ session, 09/08/2004, UN Doc.  E/CN.4/SUB.2/DEC/2004/104

\bibitem {UN SUBCOM } UN SUBCOM, {\em The right to development},
  56$^{th}$ session, 26/07/2004, UN Doc.E/CN.4/Sub.2/2004/14

\bibitem {UN SUBCOM } UN SUBCOM, {\em The right to development: study
  on existing bilateral and multilateral programmes and }{\em policies
  for development partnership}, 03/08/2004, 56$^{th}$ session, UN
  Doc. E/CN.4/Sub.2/2004/15/Corr.1

\bibitem {UN SUBCOM } UN SUBCOM, {\em The right to development: study
  on existing bilateral and multilateral programmes and policies for
  development partnership}, 56$^{th}$ session, 03/08/2004, UN
  Doc. E/CN.4/Sub.2/2004/15

\bibitem {UNDP } UNDP, {\em Human Rights and Human Development,} Human
  Development Report 2000, United Nations, 2000
  http://hdr.undp.org/reports/global/2000/en/. Last accessed 5.7.04

\bibitem UNDP,{\em Integrating Human Rights with Sustainable Human
  Development:} A UNDP Policy Document, United Nations, January
  1998. http://magnet.undp.org/Docs/policy5.html. Last accessed 5.7.04

\bibitem {UNDP } UNDP, {\em Report of Symposium on Human Development
  and Human Rights,} Oslo, 2-3 October 1998

\bibitem UNDP, United Nations,
  1998. http://www.undp.org/hdro/events/oslo/Oslorep1.html. Last
  accessed 10.8.04

\begin{appendices}

\chapter{Declaration on the Right to Development}

Adopted by General Assembly resolution 41/128 of 4 December 1986
\\

The General Assembly,
\\

{\em Bearing in mind} the purposes and principles of the Charter of
the United Nations relating to the achievement of international
co-operation in solving international problems of an economic, social,
cultural or humanitarian nature, and in promoting and encouraging
respect for human rights and fundamental freedoms for all without
distinction as to race, sex, language or religion,

{\em Recognizing} that development is a comprehensive economic,
social, cultural and political process, which aims at the constant
improvement of the well-being of the entire population and of all
individuals on the basis of their active, free and meaningful
participation in development and in the fair distribution of benefits
resulting therefrom,

{\em Considering} that under the provisions of the Universal
Declaration of Human Rights everyone is entitled to a social and
international order in which the rights and freedoms set forth in that
Declaration can be fully realized,

{\em Recalling} the provisions of the International Covenant on
Economic, Social and Cultural Rights and of the International Covenant
on Civil and Political Rights,

{\em Recalling further} the relevant agreements, conventions,
resolutions, recommendations and other instruments of the United
Nations and its specialized agencies concerning the integral
development of the human being, economic and social progress and
development of all peoples, including those instruments concerning
decolonization, the prevention of discrimination, respect for and
observance of, human rights and fundamental freedoms, the maintenance
of international peace and security and the further promotion of
friendly relations and co-operation among States in accordance with
the Charter,

{\em Recalling} the right of peoples to self-determination, by virtue
of which they have the right freely to determine their political
status and to pursue their economic, social and cultural development,

{\em Recalling also} the right of peoples to exercise, subject to the
relevant provisions of both International Covenants on Human Rights,
full and complete sovereignty over all their natural wealth and
resources,

{\em Mindful }of the obligation of States under the Charter to promote
universal respect for and observance of human rights and fundamental
freedoms for all without distinction of any kind such as race, colour,
sex, language, religion, political or other opinion, national or
social origin, property, birth or other status,

{\em Considering }that the elimination of the massive and flagrant
violations of the human rights of the peoples and individuals affected
by situations such as those resulting from colonialism,
neo-colonialism, apartheid, all forms of racism and racial
discrimination, foreign domination and occupation, aggression and
threats against national sovereignty, national unity and territorial
integrity and threats of war would contribute to the establishment of
circumstances propitious to the development of a great part of
mankind,

{\em Concerned }at the existence of serious obstacles to development,
as well as to the complete fulfilment of human beings and of peoples,
constituted, inter alia, by the denial of civil, political, economic,
social and cultural rights, and considering that all human rights and
fundamental freedoms are indivisible and interdependent and that, in
order to promote development, equal attention and urgent consideration
should be given to the implementation, promotion and protection of
civil, political, economic, social and cultural rights and that,
accordingly, the promotion of, respect for and enjoyment of certain
human rights and fundamental freedoms cannot justify the denial of
other human rights and fundamental freedoms,

{\em Considering} that international peace and security are essential
elements for the realization of the right to development,

{\em Reaffirming} that there is a close relationship between
disarmament and development and that progress in the field of
disarmament would considerably promote progress in the field of
development and that resources released through disarmament measures
should be devoted to the economic and social development and
well-being of all peoples and, in particular, those of the developing
countries,

{\em Recognizing} that the human person is the central subject of the
development process and that development policy should therefore make
the human being the main participant and beneficiary of development,

{\em Recognizing} that the creation of conditions favourable to the
development of peoples and individuals is the primary responsibility
of their States,

{\em Aware} that efforts at the international level to promote and
protect human rights should be accompanied by efforts to establish a
new international economic order,

{\em Confirming }that the right to development is an inalienable human
right and that equality of opportunity for development is a
prerogative both of nations and of individuals who make up nations,

{\em Proclaims} the following Declaration on the Right to Development:

{\em Article 1}

\begin{enumerate}
\item The right to development is an inalienable human right by virtue
  of which every human person and all peoples are entitled to
  participate in, contribute to, and enjoy economic, social, cultural
  and political development, in which all human rights and fundamental
  freedoms can be fully realized.

\item The human right to development also implies the full realization
  of the right of peoples to self-determination, which includes,
  subject to the relevant provisions of both International Covenants
  on Human Rights, the exercise of their inalienable right to full
  sovereignty over all their natural wealth and resources.
\end{enumerate}

{\em Article 2}

\begin{enumerate}
\item The human person is the central subject of development and
  should be the active participant and beneficiary of the right to
  development.
\item All human beings have a responsibility for development,
  individually and collectively, taking into account the need for full
  respect for their human rights and fundamental freedoms as well as
  their duties to the community, which alone can ensure the free and
  complete fulfilment of the human being, and they should therefore
  promote and protect an appropriate political, social and economic
  order for development.
\item States have the right and the duty to formulate appropriate
  national development policies that aim at the constant improvement
  of the well-being of the entire population and of all individuals,
  on the basis of their active, free and meaningful participation in
  development and in the fair distribution of the benefits resulting
  therefrom.
\end{enumerate}

{\em Article 3}

\begin{enumerate}
\item States have the primary responsibility for the creation of
  national and international conditions favourable to the realization
  of the right to development.
\item The realization of the right to development requires full
  respect for the principles of international law concerning friendly
  relations and co-operation among States in accordance with the
  Charter of the United Nations.
\item States have the duty to co-operate with each other in ensuring
  development and eliminating obstacles to development. States should
  realize their rights and fulfil their duties in such a manner as to
  promote a new international economic order based on sovereign
  equality, interdependence, mutual interest and co-operation among
  all States, as well as to encourage the observance and realization
  of human rights.
\end{enumerate}

{\em Article 4}

\begin{enumerate}
\item States have the duty to take steps, individually and
  collectively, to formulate international development policies with a
  view to facilitating the full realization of the right to
  development.
\item Sustained action is required to promote more rapid development
  of developing countries. As a complement to the efforts of
  developing countries, effective international co-operation is
  essential in providing these countries with appropriate means and
  facilities to foster their comprehensive development.
\end{enumerate}

{\em Article 5}

States shall take resolute steps to eliminate the massive and flagrant
violations of the human rights of peoples and human beings affected by
situations such as those resulting from apartheid, all forms of racism
and racial discrimination, colonialism, foreign domination and
occupation, aggression, foreign interference and threats against
national sovereignty, national unity and territorial integrity,
threats of war and refusal to recognize the fundamental right of
peoples to self-determination.

{\em Article 6}

\begin{enumerate}
\item All States should co-operate with a view to promoting,
  encouraging and strengthening universal respect for and observance
  of all human rights and fundamental freedoms for all without any
  distinction as to race, sex, language or religion.
\item All human rights and fundamental freedoms are indivisible and
  interdependent; equal attention and urgent consideration should be
  given to the implementation, promotion and protection of civil,
  political, economic, social and cultural rights.
\item States should take steps to eliminate obstacles to development
  resulting from failure to observe civil and political rights, as
  well as economic social and cultural rights.
\end{enumerate}

{\em Article 7}

All States should promote the establishment, maintenance and
strengthening of international peace and security and, to that end,
should do their utmost to achieve general and complete disarmament
under effective international control, as well as to ensure that the
resources released by effective disarmament measures are used for
comprehensive development, in particular that of the developing
countries.

{\em Article 8}

\begin{enumerate}
\item States should undertake, at the national level, all necessary
  measures for the realization of the right to development and shall
  ensure, inter alia, equality of opportunity for all in their access
  to basic resources, education, health services, food, housing,
  employment and the fair distribution of income. Effective measures
  should be undertaken to ensure that women have an active role in the
  development process. Appropriate economic and social reforms should
  be carried out with a view to eradicating all social injustices.
\item States should encourage popular participation in all spheres as
  an important factor in development and in the full realization of
  all human rights.
\end{enumerate}

{\em Article 9}

\begin{enumerate}
\item All the aspects of the right to development set forth in the
  present Declaration are indivisible and interdependent and each of
  them should be considered in the context of the whole.
\item Nothing in the present Declaration shall be construed as being
  contrary to the purposes and principles of the United Nations, or as
  implying that any State, group or person has a right to engage in
  any activity or to perform any act aimed at the violation of the
  rights set forth in the Universal Declaration of Human Rights and in
  the International Covenants on Human Rights.
\end{enumerate}

{\em Article 10}

Steps should be taken to ensure the full exercise and progressive
enhancement of the right to development, including the formulation,
adoption and implementation of policy, legislative and other measures
at the national and international levels.

GA Resolution 4/128, December 4 1986, available at
http://www.unhchr.ch/html/menu3/b/74.htm

The Charter of the United Nations 1945, Article 55 states: 'With a
view to the creation of conditions of stability and well-being which
are necessary for peaceful and friendly relations among nations based
on respect for the principle of equal rights and self-determination of
peoples, the United Nations shall promote: a. higher standards of
living, full employment, and conditions of economic and social
progress and development; b. solutions of international economic,
social, health, and related problems; and international cultural and
educational co-operation; and c. universal respect for, and observance
of, human rights and fundamental freedoms for all without distinction
as to race, sex, language, or religion'. Article 56: 'All Members
pledge themselves to take joint and separate action in cooperation
with the Organisation for the achievement of the purposes set forth in
Article 55'.

Universal Declaration of Human Rights 1948, Adopted by GA Resolution 217A III, December 10, 1949
Ibid Articles 1-21
Ibid Articles 22-28
Eide A, Krause C, Rosas A, '{\em Economic, social and cultural rights: A Textbook', }1995, Martinus Nijhoff Publishers, The Netherlands.
International Labor Organisation Philadelphia Declaration, Article 2(a)
The Charter of the United Nations 1945, Preamble
Ibid Article 55
Ibid Article 56
Ibid Article 68
Cited in O'Brien J, {\em International law}, London: Cavendish, 2001, p478
Article 1 of the Universal Declaration of Human Rights
Article 2 ibid

CPR's in the Universal Declaration of Human Rights included: Article
3, the right to life liberty and security; Article 4, freedom from
slavery; Article 5, freedom from torture, or cruel, inhuman or
degrading treatment; Article 6 and Article 7 recognition and equality
before the law; Article 8, the right to an effective remedy; Article
9, freedom from arbitrary arrest, detention and exile; Articles 10 and
11, due process of law; Article 12, the right to privacy; Article 13,
freedom of movement; Article 14, right to seek asylum; Article 15,
right to a nationality; Article 16, right to marry; Article 17, to own
property; Article18, 19 and 20, freedoms of thought, expression and
peaceful assembly; Article 21, the right to participate in the
government of ones country.

ESCR's in the Universal Declaration of Human Rights included: Article
23, the right to work; Article 24, the right to leisure and rest;
Article 25, the right to an adequate standard of living including
food, clothing, housing and medical care and necessary social
services; Article 26,  the right to education; Article 27, the right
to participate in the community.

See  O'Brien J, {\em International law}, London: Cavendish, 2001;
Steiner H J and Alston P, {\em International human rights in context:
  law, politics, morals}, 2$^{nd}$ ed, Oxford :Oxford UP, 2000 ; Shaw
M N, {\em International law}. 5$^{th}$ ed. Cambridge: Cambridge UP,
2003.

GA Res 1161 (XII) 26$^{th}$ November 1957

UN Doc E/3347Rev. 1 1960

The International Covenant on Civil and Political Rights 1966,
includes from Article 1 which is a general acknowledgement of the
right of peoples to self-determination to the right to life, Art 6;
freedom from torture and inhuman treatment, Art 7; freedom from
slavery and forced labour, Art 8; the right to liberty and security,
Art 9; the right of detained persons to be treated with humanity, Art
10; freedom from imprisonment for debt, Art 11; freedom of movement
and choice of residence, Art 12; freedom of aliens from arbitrary
expulsion, Art 13; the right to a fair trial, Art 14;  prohibition of
conviction under retroactive criminal laws, Art 15; the right to
recognition as a person before the law, Art 16; the right to privacy,
Art 17; freedom of thought, conscience and religion, Art 18; freedom
of opinion and expression, Art 19; the prohibition of propaganda for
war and of incitement to national, racial or religious hatred, Art 20;
freedom of assembly, Art 21; freedom of association, Art 22; the right
to marry and have a family, Art 23; the rights of a child, Art 24; the
right to participation in political life, Art 25; the right to
equality before the law, Art 26 and the rights of minorities, Art 27.

The International Covenant on Civil and Political Rights 1966,
Optional Protocol 1 of 1966, in effect 1976, Article 1 'A State Party
to the Covenant that becomes a Party to the present Protocol
recognises the competence of the Committee to receive and consider
communications from individuals subject to its jurisdiction who claim
to be victims of a violation by that State Party of any of the rights
set forth in the Covenant. No communication shall be received by the
Committee if it concerns a State Party to the Covenant which is not a
Party to the present Protocol'

International Covenant on Economic, Social and Cultural Rights 1966:
Art 6, the right to employment; Art 7, the right to just and
favourable conditions of work, including fair wages, equal pay, and
paid holidays; Art 8, the right to form and join a trade union and the
right to strike; Art 9, the right to social security, Art 10,
protection of the family including mothers and children; Art 11, The
right to an adequate standard of living including food, clothing and
housing; Art 12, the right to health, physical and mental; Art 13, the
right to education, Art 14 permits states to progressively implement
Art 13; Art 15 the right to participate in the cultural life of the
community.

Ibid Article 18

Ibid Article 19

See Baehr P R., {\em Human rights universality in practice
}. Basingstoke: Palgrave, 2001; O'Brien J, {\em International law},
London: Cavendish, 2001;  Shaw M N, {\em International law}. 5$^{th}$
ed. Cambridge: Cambridge UP, 2003

Regional protections include the European Social Charter 1961.

See Vasak K and Alston P (eds)., {\em The International Dimensions of
  Human Rights}, Westport, CT: Greenwood Press, Volume 1, 1982.

Ibid

See Alston P.,{\em  Revitalising United Nations Work on Human Rights
  and Development}, 18 MELB U. L. Rev. 216-219

Ibid

Ibid

The Proclamation of Tehran, International Conference on Human Rights,
Tehran 13 May 1968, GA Res 2442 (XXIII) 19 December 1968

Declaration on Social Progress and Development, adopted 11 December
1969, GA Res 2542 (XXIV)

Cited in Alston P., {\em 'The Right to Development at the
  International Level'} in R. Dupuy (ed), The Right to Development at
the International Level,  Netherlands: Sijthoff and Noordhoff, 1981 p
101

Article 3 of the Declaration on Race and Racial Prejudice 1978
(UNESCO)

Declaration on the Preparation of Societies for Life in Peace 1978,
Resolution 33/73 (1978)

Commission on Human Rights, Resolution 4, (XXXV) 1979

African Charter of Human and Peoples' Rights 1981

The Federal Republic of Germany

Japan

Australia, although they did not abstain from TDRTD, they still
expressed this concern.

USA

UK

See Bunn I D., '{\em The Right to Development: Implications for
  International Economic Law'}, 15 American University International
Law Review 1425 (2000): 1425-67.

The Working Group of Governmental Experts was established pursuant to
UN Commission on Human Rights  Resolution 36 (XXXVII) of 11$^{th}$
March 1981.

Upon the recommendation of the Working Group of Governmental Experts

UN Commission on Human Rights  Resolution 1989/45 of 6$^{th}$ March
1989

Established pursuant to UN Commission on Human Rights  Resolution
1993/22 (1993). The second working group consisted of 15 experts and
had a mandate for three years, to identify the obstacles of
implementation and realisation of TRTD and to make recommendations
towards its realisation and implementation by all states . The third
working group was established pursuant to UN Commission on Human
Rights  Resolution 1996/15. This group consisted of 10 experts, who
had a mandate for two years, to elaborate on a strategy for the
implementation and promotion of TRTD, concerning its multidimensional
and integrated aspects.

General Assembly Resolution 52/156, 12$^{th}$ December 1997 UN
Doc. A/RES/52/136 adopted by a vote with 129 States in favour, 12
against and 32 abstentions.

The Independent Expert on the Right to Development, Arjun Sengupta,
has written extensively on TRTD and has produced a number of reports
for the Commission on Human Rights and also several journal articles.

Commission on Human Rights Res. 72 UN Doc. E/CN.4/1998/177 (1998)

See preamble and Article 6 of TDRTD

See Article 6 of TDRTD

This is beyond the scope of this paper, for further reading see Baehr
P R., {\em Human rights universality in practice}. Basingstoke:
Palgrave, 2001; Donnelly J., {\em Universal Human Rights in Theory and
  Practice}, Ithaca: Cornell University Press, 1989.

Arjun Sengupta, Fourth Report of the independent expert to the
Commission on Human Rights , E/CN.4/2002/WG.18/4, 20 December 2001,
para 25.

Arjun Sengupta, Fourth Report of the independent expert to the
Commission on Human Rights , E/CN.4/2002/WG.18/4, 20 December 2001,
para 25.

The IE defines non-discrimination as; ``Non-discrimination means that
there will be no discrimination on the grounds of race, colour, sex,
language, political or other opinion, religion, national or social
origin, property, birth or other status, not only between the
beneficiaries but also between agents''. See Arjun Sengupta, Fourth
Report of the independent expert to the Commission on Human Rights ,
E/CN.4/2002/WG.18/4, 20 December 2001 paras 26-32. Most of what the IE
states is fairly straightforward but the concept of non-discrimination
``between agents'' was not specifically mentioned in TDRTD.  This
element is not clarified satisfactorily in the IE's reports nor was it
found to be used outside of the before-mentioned documents. Thus an
in-depth discussion on this matter this is considered to be outside of
the scope of the paper.

It has been noted above that the word `should' carries less force and
thus perhaps states could interpret the provision as carrying less
force than other documents have given it. For example the ICESCR non
discrimination clause in Article 2(2) states; ``The State Parties to
the present Covenant undertake to guarantee that the rights enunciated
in the present Covenant will be exercised without discrimination of
any kind as to race, colour, sex, language, religion, political or
other opinion, national or social origin, property, birth or other
status''. Article 2(1) of the ICCPR states; ``Each State Party to the
present Covenant undertakes to respect and to ensure to all
individuals within its territory and subject to its jurisdiction the
rights recognised in the present Covenant, without distinction of any
kind, such as to race, colour, sex, language, religion, political or
other opinion, national or social origin, property, birth or other
status''.

Cited in Bunn I D., '{\em The Right to Development: Implications for
  International Economic Law'}, 15 American University International
Law Review 1425 (2000): 1425-67. p1447 Footnote 81.

United Nations, {\em The realization of the right to development:
  Global consultation on the right to development as a human right :
  report prepared by the Secretary-General pursuant to Commission on
  Human Rights resolution 1989/45,} United Nations, 1991, at 48.

Bunn I D., '{\em The Right to Development: Implications for
  International Economic Law'}, 15 American University International
Law Review 1425 (2000): 1425-67. p1445

 Human Rights Council of Australia Inc. {\em The Rights Way to
   Development: A Human Rights Approach to Development
   Assistance}. (1995) 118-21

Arjun Sengupta, Fourth Report of the independent expert to the
Commission on Human Rights , E/CN.4/2002/WG.18/4, 20 December 2001
paras 26-32

See Bunn I D., '{\em The Right to Development: Implications for
  International Economic Law'}, 15 American University International
Law Review 1425 (2000): 1425-67 p1451 (Report of the Secretary-General
on the Regional and International Dimensions of the Right to
Development, 1980).

Arjun Sengupta, Fourth Report of the independent expert to the
Commission on Human Rights , E/CN.4/2002/WG.18/4, 20 December 2001
paras 26-32

Article 2(3) TDRTD

Article 8 TDRTD

Arjun Sengupta, Fourth Report of the independent expert to the
Commission on Human Rights , E/CN.4/2002/WG.18/4, 20 December 2001
para 26.

Bunn I D., '{\em The Right to Development: Implications for
  International Economic Law'}, 15 American University International
Law Review 1425 (2000): 1425-67 p1447

Article 1(1) TDRTD

Article 1(1) TDRTD

Article 2(2) TDRTD

Article 5 TDRTD

Articles 6(1), 6(2) TDRTD

Article 6(3) TDRTD could imply this is less obligatory because of the
term ``should'' and thus is open to interpretation.

Sengupta A., {\em Theory and Practice on the Right to Development},
Human Rights Quarterly; V.24(4); pp.837-889, p 843

Sen A., '{\em The Right not to be Hungry'}, in Alston P., and
Tomasevski K., (eds), The Right to Food, Netherlands: SIM, 1984.

See Donnelly J, {\em In Search of the Unicorn : The Jurisprudence and
  Politics of the Right to Development}, 15 Cal W Int'l L J 473, 1985

Sengupta A, {\em Theory and Practice on the Right to Development},
Human Rights Quarterly; V.24(4); pp.837-889.

Donnelly J, {\em In Search of the Unicorn : The Jurisprudence and
  Politics of the Right to Development}, 15 Cal W Int'l L J 473, 1985

Alston P., {\em Making Space for New Human Rights : the case of the
  right to development,} Harvard Human Rights Yearbook; V.1; pp.3-40 p
32

Ibid p 29

Ibid p 29 and also Footnote 94

Article 1(2) TDRTD recognises the right of peoples to self
determination. This has been endorsed in numerous human rights
instruments and has been the subject of intense debate. For this
reason, this aspect of TRTD will not be discussed in this paper. For
further reading see; Alston P. (ed), {\em Peoples' Rights}, Oxford:
Oxford University Press, 2001; Abi- Saab G., {\em 'The Legal
  Formulation of the Right to Development}', in R. Dupuy (ed), The
Right to Development at the International Level,  Netherlands:
Sijthoff and Noordhoff, 1981. O'Brien J, {\em International law},
London: Cavendish, 2001; Steiner H J and Alston P, {\em International
  human rights in context: law, politics, morals}, 2$^{nd}$ ed, Oxford
:Oxford UP, 2000 ; Shaw M N, {\em International law}. 5$^{th}$
ed. Cambridge: Cambridge UP, 2003;

Rich R., '{\em The Right to Development: A Right of Peoples?}', in
J. Crawford (ed), The Rights of Peoples, Oxford: Clarendon Press,
1988, 39-54.

See Donnelly J., {\em In Search of the Unicorn : The Jurisprudence and
  Politics of the Right to Development}, 15 Cal W Int'l L J 473, 1985;
Donnelly J., {\em Human Rights as Natural Rights}, 4 Hum Rts Q 391,
1982; Donnelly J., {\em Universal Human Rights in Theory and
  Practice}, Ithaca: Cornell University Press, 1989; Ghai Y., '{\em
  Human Rights and Governance: The Asia Debate'}, 15 Australian Year
Book of International Law (1994) 1.

See Alston P., {\em Making Space for New Human Rights : the case of
  the right to development,} Harvard Human Rights Yearbook; V.1;
pp.3-40; Alston P., '{\em The Shortcomings of a ``Garfield the Cat''
  Approach to the Right to Development'}, California Western
International Law Journal, Vol 15 (1985), 510; Alston P., {\em A third
  generation of solidarity rights,: progressive development or
  obfuscation of international human rights law?} 29 Netherlands Int'l
L Rev 307, 1982; Abi- Saab G., {\em 'The Legal Formulation of the
  Right to Development}', in R. Dupuy (ed), The Right to Development
at the International Level,  Netherlands: Sijthoff and Noordhoff,
1981, p163; Marks S., {\em Emerging human rights : a new generation fo
  the 1980's?} 33 Rutgers L Rev 435, 451, 1981; Obiora Amede L,{\em
  Beyond the Rhetoric of a Right to Development}, Law and Policy;
V.18(3/4); pp.355-418For further reading, see Hausserman J, {\em The
  Realisation and Implementation of Economic Social and Cultural
  Rights}, London: Macmillan, 1992; Eide A, Krause C, Rosas A, {\em
  'Economic, Social and Cultural Rights: A Textbook'},  The
Netherlands: Martinus Nijhoff Publishers, 1995.

See Rich R., '{\em The Right to Development: A Right of Peoples?}', in
J. Crawford (ed), The Rights of Peoples, Oxford: Clarendon Press,
1988, 39-54.

Prott L, {\em Cultural Rights as Peoples' Rights}, in Crawford (ed)
The Rights of Peoples', 1992, Oxford University Press, New York.

Prott L, {\em Cultural Rights as Peoples' Rights}, in Crawford (ed)
The Rights of Peoples', 1992, Oxford University Press, New York. P
102-103

Sieghart's articulation of the incompatible nature of collective
rights demonstrates how dangerous the concept can actually be. See
Sieghart P., {\em The International Law of Human Rights}, Oxford:
Oxford University Press, 1985,  p163-164

Ghai Y., '{\em Human Rights and Governance: The Asia Debate'}, 15
Australian Year Book of International Law (1994) 1, at 10

Ibid

Donnelly J., {\em In Search of the Unicorn : The Jurisprudence and
  Politics of the Right to Development}, 15 Cal W Int'l L J 473, 1985,
p 499

Article 9 DRTD

Article 2 DRTD

To be discussed below.

Orford A., '{\em Globalisation and the Right to Development}', in
Alston P. (ed), Peoples' Rights, Oxford: Oxford University Press,
2001., 127-185, p137

Alston P., '{\em The Shortcomings of a ``Garfield the Cat'' Approach
  to the Right to Development'}, California Western International Law
Journal, Vol 15 (1985), 510, p 512

To be discussed below

Galenkamp M., {\em Collective Rights}, Report to the Advisory
committee on Human Rights and Foreign Policy of the Netherlands,
Rotterdam, Netherlands: 1998 available at
http://www.uu.nl/content/16-3.pdf p22

Alston P., {\em 'The Right to Development at the International Level'}
in R. Dupuy (ed), The Right to Development at the International Level,
Netherlands: Sijthoff and Noordhoff, 1981; Alston P., '{\em The
  Shortcomings of a ``Garfield the Cat'' Approach to the Right to
  Development'}, California Western International Law Journal, Vol 15
(1985), 510.

Working Group of Government Experts on the Right to Development, UN
Doc E/1981/25 at para 17

Working Group of Government Experts on the Right to Development UN Doc
ST/HR/SER.A.A/A at para 72

{\em The International Dimensions of the Right to Development as a
  Human Right in relation to Other Human Rights based on International
  Cooperation, including the Right to Peace, taking into account the
  Requirements of the New International Economic Order and the
  Fundamental Human Needs}, Report of the Secretary-General, UN Doc
E/CN.4/1334 (1979),  Cited in Donnelly J., {\em In Search of the
  Unicorn : The Jurisprudence and Politics of the Right to
  Development}, 15 Cal W Int'l L J 473, 1985, p 475

Ibid

Ghai Y., '{\em Human Rights and Governance: The Asia Debate'}, 15
Australian Year Book of International Law (1994) 1, p10

Donnelly J., {\em In Search of the Unicorn : The Jurisprudence and
  Politics of the Right to Development}, 15 Cal W Int'l L J 473, 1985,
p 502

Orford A., '{\em Globalisation and the Right to Development}', in
Alston P. (ed), Peoples' Rights, Oxford: Oxford University Press,
2001., 127-185 at p140

Donnelly J., {\em In Search of the Unicorn : The Jurisprudence and
  Politics of the Right to Development}, 15 Cal W Int'l L J 473, 1985,
p 502

As described in Donnelly J., {\em In Search of the Unicorn : The
  Jurisprudence and Politics of the Right to Development}, 15 Cal W
Int'l L J 473, 1985,  p 502; See Arjun Sengupta, Fifth Report of the
Independent Expert on the Right to Development, UN Doc
E/CN.4/2002/WG.18/6 (2002)

Sengupta A., {\em Theory and Practice on the Right to Development},
Human Rights Quarterly; V.24(4); pp.837-889, p 868

See Orford A., '{\em Globalisation and the Right to Development}', in
Alston P. (ed), Peoples' Rights, Oxford: Oxford University Press,
2001., 127-185,  p144

See Udombana N., '{\em The Third World and the Right to Development:
  Agenda for the Next Millennium'}, Human Rights Quarterly, Vol 22(3)
(2000), 753-87.  p755, footnote 9.

Rodney W., {\em How Europe Underdeveloped Africa}, Howard University
Press, (revised ed) 1982, London and Tanzanian Publishing House: Dar
es Salaam.

Ibid, Rodney analyses what economic development is and how people
developed themselves throughout history by having the capacity,
jointly, to deal with nature. For him, this means, people being able
to exploit the natural resources available to them and also
understanding the laws of science and technology in order to
effectively do this. See p 1-31

Uvin P., {\em On High Moral Ground: The Incorporation of Human Rights
  by the Development Enterprise}, Praxis- The Fletcher Journal of
Development Studies, Volume XVII 2002,   available at
\uline{http://fletcher.tufts.edu/praxis/xvii/Uvin.pdf }p2

Frankovits stated that; ``With an increasing demand for economic and
social rights to be a major factor in development assistance, donors
have tended to reformulate their terminology. Beginning with the World
Bank's statement at the 1993 Conference on Human Rights in Vienna,
followed by frequently heard assertions by individual donor agencies,
the claim is made that all development assistance contributes to
economic and social rights. Thus agricultural projects~--- whatever
their nature~--- are claimed to contribute directly to the fulfilment
of the right to food.'' Frankovits A., '{\em Rejoinder:The Rights Way
  to Development}', Food Policy 21 no 1, 123-28, p 26

Udombana N., '{\em The Third World and the Right to Development:
  Agenda for the Next Millennium'}, Human Rights Quarterly, Vol 22(3)
(2000), 753-87. p 757

See UNDP Human Development Reports available at http://hdr.undp.org/

Charlesworth stated that; ``While the formulation of the right to
development does not rest on a simple economic model of development,
and includes within it a synthesis of all recognised human rights,
redress of economic inequality is at its heart. An assumption of the
international law of development is that underdevelopment is caused by
a failure to meet the model of a capitalist economy. Development means
industrialisation and westernisation.'' Charlesworth H, `The
Public/Private Distinction and the Right to Development in
International Law, 12 Australian Year Book of International Law (1992)
at 196-197

Sieghart P., {\em The International Law of Human Rights}, Oxford:
Oxford University Press, 1985,  p164-165

Sen A., {\em Development as Freedom}, Oxford University Press: Oxford,
1999

Sen A., '{\em The Right not to be Hungry'}, in Alston P., and
Tomasevski K., (eds), The Right to Food, Netherlands: SIM, 1984.

Piron L-H, {\em The Right to Development: A Review of the Current
  State of the Debate}, (2002). Prepared for the UK Department for
International
Development. http://www.odi.org.uk/pppg/publications/papers\_reports/dfid/issues/rights01/index.html. Last
accessed 26.10.04

Article 2(2) of TDRTD

See Arjun Sengupta, Fourth Report of the independent expert to the
Commission on Human Rights , E/CN.4/2002/WG.18/4, 20 December 2001.

Cited in Steiner and Alston, International Human Rights in Context:
Law, Politics and Morals, (1996) at 488. Also see Orford A., '{\em
  Globalisation and the Right to Development}', in Alston P. (ed),
Peoples' Rights, Oxford: Oxford University Press, 2001., 127-185  at
143

Article 3(1) of TDRTD

Report of the Working Group on the Right to Development, UN Doc
E/CN.4/1995/27

Udombana N., '{\em The Third World and the Right to Development:
  Agenda for the Next Millennium'}, Human Rights Quarterly, Vol 22(3)
(2000), 753-87.

See Odumbana for an interesting notion of how Third World countries
need to define and implement development for themselves. He states
Third World countries ``must devise internal strategies for economic
growth. They must develop their own resources and technology. Their
future will remain bleak as long as they continue to copy foreign
patterns of development. They should search for means of development
within their own resources. They must change their attitude of
depending on the goodwill of others \ldots They must, consequently,
begin to pay more attention to the traditional values and attitudes of
their societies.'' Udombana N., '{\em The Third World and the Right to
  Development: Agenda for the Next Millennium'}, Human Rights
Quarterly, Vol 22(3) (2000), 753-87, at 763

See Reports of the Independent Expert on the Right to Development,
Arjun Sengupta, UN Docs E/CN.4/2002/WG.18/6 ;
E/CN.4/2002/WG.18/6/Add.1; E/CN.4/2002/WG.18/2;

Steiner and Alston enumerate five different obligations which include;
to respect the rights of others, create institutional machinery
essential to the realisation of rights, protect rights and prevent
violations, provide goods and services to satisfy rights and promote
rights. See Steiner H J., and Alston P., {\em International human
  rights in context: law, politics, morals}, 2$^{nd}$ ed, Oxford
:Oxford UP, 2000, pp 182-184

Marks S, {\em The Human Rights Approach to Development: 5 Approaches},
FXB Centre for Health and Human Rights, Working Paper 5
(2000). http://www.hsph.harvard.edu/fxbcenter/FXBC\_WP6-- Marks.pdf.
At 14

Ibid

Ibid

Ibid

Article 3(3) of TDRTD

For further reading see Udombana N., '{\em The Third World and the
  Right to Development: Agenda for the Next Millennium'}, Human Rights
Quarterly, Vol 22(3) (2000), 753-87.

Article 4 of TDRTD

The specific paragraph of the preamble of TDRTD states that ``Bearing
in mind} the purposes and principles of the Charter of the  United
  Nations relating to the achievement of international cooperation in
  solving international problems of an economic, social, cultural and
  humanitarian nature, and in promoting and encouraging respect for
  human rights and fundamental freedoms.''

Article 1 UN Charter

Article 55 UN Charter

Article 56 UN Charter

See Sengupta A., {\em Theory and Practice on the Right to
  Development}, Human Rights Quarterly; V.24(4); pp.837-889, p854

Ibid

Office of the High Commissioner on Human Rights, The Right to
Development, www.unhchr.ch/development/right. Also see Sengupta A,
{\em Towards Realizing the Right to Development}, Development and
Change; June; V.31(3)

Donnelly J., {\em In Search of the Unicorn : The Jurisprudence and
  Politics of the Right to Development}, 15 Cal W Int'l L J 473, 1985
at 505

Ibid at 506

Ibid

Ibid

Ghai Y., '{\em Human Rights and Governance: The Asia Debate'}, 15
Australian Year Book of International Law (1994) 1.  at 10

Ibid

Orford A., '{\em Globalisation and the Right to Development}', in
Alston P. (ed), Peoples' Rights, Oxford: Oxford University Press,
2001., 127-185 at p 144

Ibid

See Alston P., '{\em The Myopia of the Herdmaidens: International
  Lawyers and Globalisation'}, 8 European Journal of International Law
(1997) 435

Orford A., '{\em Globalisation and the Right to Development}', in
Alston P. (ed), Peoples' Rights, Oxford: Oxford University Press,
2001., 127-185 at p 144

Ibid

The link between globalisation and TRTD and the activities of the WB,
IMF and WTO have been extensively discussed in  Orford A., '{\em
  Globalisation and the Right to Development}', in Alston P. (ed),
Peoples' Rights, Oxford: Oxford University Press, 2001., 127-185. It
is deemed as beyond the scope of this paper to discuss this issue in
any depth.

Article 7 TDRTD

However, due to the nature of this paper and the strict restrictions
and as further research is necessary to understand the links between
security, peace and development, this area is deemed as beyond the
scope of this paper.

Article 38(1) of the Statute of the Court states; ``The Court, whose
function is to decide in accordance with international law such
disputes as are submitted to it, shall apply: (a) international
conventions, whether general or particular, establishing rules
expressly recognized by the contesting states; (b) international
custom, as evidence of a general practice accepted as law; (c) the
general principles of law recognized by civilized nations; (d) subject
to the provisions of Article 59, judicial decisions and the teachings
of the most highly qualified publicists of the various nations, as
subsidiary means for the determination of rules of law'', Article 38
(2) states: ``This provision shall not prejudice the power of the
Court to decide a case {\em ex aequo et bono}, if the parties agree
thereto''. Article 59: ``The decision of the Court has no binding
force except between the parties and in respect of that particular
case.''

See Sieghart P., {\em The International Law of Human Rights}, Oxford:
Oxford University Press, 1985

Ibid p92

There are other incentives to create human rights treaties, for
further reading see Sieghart P., {\em The International Law of Human
  Rights}, Oxford: Oxford University Press, 1985

Provisions contained in other international treaties which are linked
to TRTD indirectly will be discussed in a separate section.

Adopted June 27, 1981, OAU Doc. CAB/LEG/67/3 rev. 5, 21 I.L.M. 58
(1982), {\em entered into force} Oct. 21, 1986

Umozurike U O., {\em The African Charter on Human and Peoples'
  Rights}, 77 Am. J. Int'l L. 902, (1983). at 911

Ibid

Umozurike U O., {\em The African Charter on Human and Peoples'
  Rights}, 77 Am. J. Int'l L. 902, (1983). at 911

Slimane S, '{\em Recognising Minorities in Africa}', Minority Rights
Group International 2003, May 2003. available at
www.minorityrights.org

Ibid at p 3.

Umozurike U O., {\em The African Charter on Human and Peoples'
  Rights}, 77 Am. J. Int'l L. 902, (1983). at 911

Ibid

Slimane S, '{\em Recognising Minorities in Africa}', Minority Rights
Group International 2003, May 2003. available at
www.minorityrights.org p 4

The European Convention for the Protection of Human Rights and
Fundamental Freedoms, entered into force on 3$^{rd}$ September 1953,
as amended by Protocol II which entered into force on 1$^{st}$
November 1998.

The American Convention on Human Rights entered into force on
18$^{th}$ July 1978

However, an additional Protocol to establish an African Court on Human
Rights entered into force on 26$^{th}$ December 2003. See Protocol to
the African Charter on Human and Peoples' Rights Establishing an
African Court on Human and Peoples' Rights.

As noted above

Morel C., {\em Defending Human Rights in Africa: The Case for Minority
  and Indigenous Rights}, Essex Human Rights Review. Vol 1. No.1  p55

The African Commission on Human and People's Rights, Guidelines for
Submission of Communications, published by the Secretariat of the
African Commission on Human and People's Rights, available at:
http://www.achpr.org/english/\_info/guidelines\_communications\_en.html

{\em Social and Economic Rights Action Center and the Center for
  Economic and Social Rights v Nigeria }Communication 155/96, 27
October 2001

Article 24 of the Banjul Charter states; ``All peoples shall have the
right to a general satisfactory environment favourable to their
development.'' This principle will be discussed in the subsequent
chapter.

{\em Social and Economic Rights Action Center and the Center for
  Economic and Social Rights v Nigeria }Communication 155/96, 27
October 2001, para 55

{\em Malawi African Association etc v Mauritania} (2000) AHRLR 149
(ACHPR 2000).

{\em Social and Economic Rights Action Center and the Center for
  Economic and Social Rights v Nigeria }Communication 155/96, 27
October 2001

{\em Social and Economic Rights Action Center and the Center for
  Economic and Social Rights v Nigeria }Communication 155/96, 27
October 2001. Paragraph 67 states ``the survival of the Ogonis
depended on their land and farms \ldots These and similar brutalities
not only persecuted individuals on Ogoniland, but also the Ogoni
community as a whole.''

Morel C., {\em Defending Human Rights in Africa: The Case for Minority
  and Indigenous Rights}, Essex Human Rights Review. Vol 1. No.1
available at
\uline{http://projects.essex.ac.uk/EHRR/archive/pdf/49.pdf}  p 57

For further reading see; International Law Association, London
Conference (2000),  {\em Final Report of the Committee on Formation of
  Customary (General) International Law}, Available at
\uline{www.ila-hq.org}.

Asylum Case: Columbia v Peru (1950) ICJ Rep 266

See Anglo-Norwegian Fisheries Case (United Kingdom v Norway), Judgment
of 18$^{th}$ December 1951 (1951) ICJ Rep 116. North Sea Continental
Shelf Cases (Federal Republic of Germany v Denmark); Federal Republic
of Germany v Netherlands) Judgment of \nth{20} February 1969 (1969)
ICJ Rep 3.

Military and Paramilitary Activities in and Against Nicaragua
(Nicaragua v United States of America), (Merits), Judgment of
27$^{th}$ June 1986 ICJ Rep 14

North Sea Continental  Shelf Cases (Federal Republic of Germany v
Denmark); Federal Republic of Germany v Netherlands) Judgment of
\nth{20} February 1969 (1969) ICJ Rep 3.

Ibid

Anglo-Norwegian Fisheries Case (United Kingdom v Norway), Judgment of
18$^{th}$ December 1951 (1951) ICJ Rep 116

Ibid

Asylum Case: Columbia v Peru (1950) ICJ Rep 266

Asylum Case: Columbia v Peru (1950) ICJ Rep 266, at p 288

North Sea Continental  Shelf Cases (Federal Republic of Germany v
Denmark); Federal Republic of Germany v Netherlands) Judgment of
\nth{20} February 1969 (1969) ICJ Rep 3,

Military and Paramilitary Activities in and Against Nicaragua
(Nicaragua v United States of America), (Merits), Judgment of
27$^{th}$ June 1986 ICJ Rep 14 pp 97-98

Military and Paramilitary Activities in and Against Nicaragua
(Nicaragua v United States of America), (Merits), Judgment of
27$^{th}$ June 1986 ICJ Rep 14. This was decided in relation to where
a customary rule existed alongside a treaty. `

North Sea Continental  Shelf Cases (Federal Republic of Germany v
Denmark); Federal Republic of Germany v Netherlands) Judgment of
\nth{20} February 1969 (1969) ICJ Rep 3. at pp 43-44

Ibid

International Law Association, London Conference (2000),  {\em Final
  Report of the Committee on Formation of Customary }{\em (General)
  International Law}, Available at \uline{www.ila-hq.org}. at p 26

Anglo-Norwegian Fisheries Case (United Kingdom v Norway), Judgment of
18$^{th}$ December 1951 (1951) ICJ Rep 116

Ibid pp 116 at 136

For further reading see Shaw M N, {\em International law}. 5$^{th}$
ed. Cambridge: Cambridge UP, 2003

Declaration on Principles of International Law Concerning Friendly
Relations and Co-operation among States in Accordance with the Charter
of the United Nations 1970, Resolution 2625 (XXV).

Military and Paramilitary Activities in and Against Nicaragua
(Nicaragua v United States of America), (Merits), Judgment of
27$^{th}$ June 1986 ICJ Rep 14 pp 101

Legality of the Threat or Use of Nuclear Weapons (Request for Advisory
Opinion by the World Health Organisation) (1996) ICJ Rep 90

Ibid pp 213

See Brownlie I., {\em Principles of public international law}, 5th ed,
Oxford, OUP, 1998.

It has been argued that Development Assistance could be used as
evidence of state practice of TRTD. However, the {\em opinio juris} of
states who provide such assistance, is questionable. For an in depth
study on why states do not believe they are legally obligated to give
development assistance see; Piron L-H, {\em The Right to Development:
  A Review of the Current State of the Debate}, (2002). Prepared for
the UK Department for International
Development. http://www.odi.org.uk/pppg/publications/papers\_reports/dfid/issues/rights01/index.html. Last
accessed 26.10.04

McDougal, Panel Discussions, The Legal Status of General Assembly
Resolutions, Proceedings of the American Society of International
Law. 73$^{rd}$ Annual Meeting (1979), 324, 328-29. Cited in
International Law Association, London Conference (2000), {\em Final
  Report of the Committee on Formation of Customary (General)
  International Law}, Available at www.ila-hq.org. p59

Commission of Human Rights Resolution 4 (XXXIII) 1977

General Assembly Resolution 1161 (XII) 1957 states for example, ``a
balanced and integrated economic and social development would
contribute towards the promotion and maintenance of peace and
security, social progress and better standards of living, and the
observance of and respect for human rights and fundamental
freedoms''. Compare with TDRTD Articles, for example, Article 1, 8 and
the preamble of TDRTD confirm that economic, social and cultural
development would contribute to the well-being of the entire
population.  Articles 1(1), 2(2), 5 and 6, underline the need for
respect for human rights.  Article 7 linking development and
international peace and security.

International Conference on Human Rights, Tehran 1969, Proclamation of
Tehran paragraph 12

Ibid para 13

Declaration on Social Progress and Development 1969, Resolution 2542
(XXIV) stated ``[T]he right and responsibility of each State and, as
far as they are concerned, each nation and people to determine freely
its own objectives of social development, to set its own priorities
and to decide in conformity with the principles of the Charter of the
United Nations the means and methods of their achievement without any
external interference''

Charter of Economic Rights and Duties of States 1974, Resolution 3281
(XXIX).

Declaration on Race and Racial Prejudice adopted 1978 (UNESCO)
Article 3

Declaration on the Preparation of Societies for Life in Peace 1978,
Resolution 33/73 (1978)

Commission on Human Rights, Resolution 4, (XXXV) 1979

The Working Group of Governmental Experts was established pursuant to
UN Commission on Human Rights  Resolution 36 (XXXVII) of 11$^{th}$
March 1981. With a mandate to clarify the nature and content of the
right to development including its national and international aspects
and also to prepare a legal document of some kind on TRTD, (a
resolution, convention or declaration, see Commission on Human Rights,
thirty-eighth session, 1 February-12 March 1982, Report of the Working
Group of governmental experts on the right to development. UN
Doc. E/CN.4/1489. 11$^{th}$ February 1982, paragraph 45)

African Charter of Human and Peoples' Rights 1981 (also known as the
Banjul Charter)

If, on the other hand it was not taken for granted that the USA was a
state whose interests were specially affected by the rule, then TDRTD
could be used as evidence of uniform practice and the USA could be
regarded as a persistent objector who would not be bound by the rule
if it was finally established as customary law. However, this is
merely speculative and it would not be likely that anyone could argue
that the USA was not a state whose interests were specially affected.

The eight abstentions to the declaration were made by Denmark,
Finland, The Federal Republic of Germany, Iceland, Israel, Japan,
Sweden and the United Kingdom. Objections to TDRTD included; the
possibility that the declaration would lead to an erosion of
individual rights (The Federal Republic of Germany); TRTD could be
used to legitimise violations of the rights of citizens (Japan).
TDRTD did not make a distinction between individual rights and group
rights (Australia, although they did not abstain from TDRTD, they
still expressed this concern.); the human rights discourse tended to
become diluted and confused with the addition of TRTD (USA); the
declaration provided a mistaken link between human rights and a NIEO,
furthermore it did not take into account the complex relationship
between development, security and disarmament but rather provided a
very simplistic view of the relationship (UK).

Birnie P., and Boyle A., {\em International Law \& The Environment},
2$^{nd}$ ed, Oxford: Oxford University Press, 2002. p87

For example, Principle 4 (integration of environmental protection and
development), Principle 10 (public participation), Principle 15
(precautionary approach) and Principle 17 (environmental impact
assessment) were all strongly supported by developed states.

Principle 3 (right to development), Principles 6 \& 7 (special needs
of developing countries and principle of common but differentiated
responsibilities), Principles 5 \& 9 (poverty and capacity building),
were strongly supported by developing countries.

Principle 4 ``In order to achieve sustainable development,
environmental protection shall constitute an integral part of the
development process and cannot be considered in isolation from it''.

Birnie P., and Boyle A., {\em International Law \& The Environment},
2$^{nd}$ ed, Oxford: Oxford University Press, 2002. p87

Birnie P., and Boyle A., {\em International Law \& The Environment},
2$^{nd}$ ed, Oxford: Oxford University Press, 2002. p156 footnote 71

Report of the Rio Conference on Environment and Development UN Doc
A/CONF.151/26/Rev.1 (Vol II) (1993) Para 16

Birnie P., and Boyle A., {\em International Law \& The Environment},
2$^{nd}$ ed, Oxford: Oxford University Press, 2002. p87

GA Resoulution UN Doc A/RES/47/ 123 of 18 December 1992

See paragraph 10, ``The World Conference on Human Rights reaffirms the
right to development, as established in the Declaration on the Right
to Development, as a universal and inalienable right and an integral
part of fundamental human rights. As stated in the Declaration on the
Right to Development, the human person is the central subject of
development. While development facilitates the enjoyment of all human
rights, the lack of development may not be invoked to justify the
abridgement of internationally recognized human rights. States should
cooperate with each other in ensuring development and eliminating
obstacles to development. The international community should promote
an effective international cooperation for the realization of the
right to development and the elimination of obstacles to
development. Lasting progress towards the implementation of the right
to development requires effective development policies at the national
level, as well as equitable economic relations and a favourable
economic environment at the international level.''

Sengupta A., {\em Theory and Practice on the Right to Development},
Human Rights Quarterly; V.24(4); pp.837-889 p 841 states; ``There has
been considerable debate as to whether the right to development can be
regarded as a human right. This issue can now be taken as settled,
after the achievement of the consensus for the Vienna Declaration and
Programme of Action  in 1993''

Boven TV.,{\em  Human Rights and Rights of Peoples}, 6 EJIL (1995)
1-476

Boven TV.,{\em  Human Rights and Rights of Peoples}, 6 EJIL (1995)
1-476, p9

Boven TV.,{\em  Human Rights and Rights of Peoples}, 6 EJIL (1995)
1-476, p9

World Conference on Human Rights, Vienna, Vienna Declaration and
Programme of Action 1993, Paragraph 15

See Shaw M N, {\em International law}. 5$^{th}$ ed. Cambridge:
Cambridge UP, 2003.

World Conference on Human Rights, Vienna, Vienna Declaration and
Programme of Action 1993, paragraph 11 states ``11. The right to
development should be fulfilled so as to meet equitably the
developmental and environmental needs of present and future
generations. The World Conference on Human Rights recognizes that
illicit dumping of toxic and dangerous substances and waste
potentially constitutes a serious threat to the human rights to life
and health of everyone. Consequently, the World Conference on Human
Rights calls on all States to adopt and vigorously implement existing
conventions relating to the dumping of toxic and dangerous products
and waste and to cooperate in the prevention of illicit
dumping. Everyone has the right to enjoy the benefits of scientific
progress and its applications. The World Conference on Human Rights
notes that certain advances, notably in the biomedical and life
sciences as well as in information technology, may have potentially
adverse consequences for the integrity, dignity and human rights of
the individual, and calls for international cooperation to ensure that
human rights and dignity are fully respected in this area of universal
concern''

Ibid Paragraph 12

Ibid Paragraph 13; ``There is a need for States and international
organizations, in cooperation with non-governmental organizations, to
create favourable conditions at the national, regional and
international levels to ensure the full and effective enjoyment of
human rights. States should eliminate all violations of human rights
and their causes, as well as obstacles to the enjoyment of these
rights.''

See General Assembly Resolutions UN Docs: A/RES/50/184 of 22 December
1995; A/RES/49/183 of 23 Decemeber 1994; A/RES/ 48/130 of 20 December
1993

For example Principle 3 ``The right to development is a universal and
inalienable right and an integral part of fundamental human rights,
and the human person is the central subject of development.  While
development facilitates the enjoyment of all human rights, the lack of
development may not be invoked to justify the abridgement of
internationally recognized human rights.  The right to development
must be fulfilled so as to equitably meet the population, development
and environment needs of present and future generations''. Paragraph
9.2 states  ``The objectives are:

 (a)  To foster a more balanced spatial distribution of the population
by promoting in an integrated manner the equitable and ecologically
sustainable development of major sending and receiving areas, with
particular emphasis on the promotion of economic, social and gender
equity based on respect for human rights, especially the right to
development;'' Also see paragraphs 3.16 and 9.2

For example Commitment 1 states ``We commit ourselves to creating an
economic, political, social, cultural and legal environment that will
enable people to achieve social development.

To this end, at the national level, we will:

(f) Reaffirm, promote and strive to ensure the realization of the
rights set out in relevant international instruments and declarations,
such as the Universal Declaration of Human Rights, 6/ the Covenant on
Economic, social and cultural rights 7/ and the Declaration on the
Right to Development, 8/ including those relating to education, food,
shelter, employment, health and information, particularly in order to
assist people living in poverty;

At the international level, we will:

(n) Reaffirm and promote all human rights, which are universal,
indivisible, interdependent and interrelated, including the right to
development as a universal and inalienable right and an integral part
of fundamental human rights, and strive to ensure that they are
respected, protected and observed.'' Also see Programme of Action,
Chapter 1, Paragraphs 15, 17 which also directly refer to TRTD.

For example states ``The Platform for Action 1995 Fourth World
Conference on Women: ``reaffirm that all human rights~--- civil,
cultural, economic, political and social, including the right to
development~--- are universal, indivisible, interdependant and
interrelated, as expressed in the Vienna Declaration and Programme of
Action adopted by the World Conference on Human Rights.''

The World Food Summit 1996: ``emphasise the need for democracy and the
promotion and protection of all human rights and freedoms, the right
to development and the full and equal participation of men and women
as essential determinants of success in achieving sustainable food
security for all.''

For example states ``While the significance of national and regional
particularities and various historical, cultural and religious
backgrounds must be borne in mind, it is the duty of all States to
promote and protect all human rights and fundamental freedoms,
including the right to development.''

General Assembly Resolution UN Doc: A/RES/50/184 of 22 December 1995

OHCHR Human Rights in Development Latest news: www.unhchr.ch/development
General Assembly Resolution UN Doc: A/RES/53/155 of 9 December 1998

United Nations Commission on Human Rights, Fifty-fourth session
Question of the realisation of the right to development, Report of the
Secretary-General in accordance with Commission resolution 1997/72. UN
Doc. E/CN1998/28. 16$^{th}$ February 1998. Para 19

See United Nations General Assembly, Fifty-third session, Human rights
questions, Right to development, Report of the Secretary-General. UN
Doc. A/53/268. 17$^{th}$ August 1998. Para 11. Ghana states that ``In
its preliminary views and observations, the Government of Ghana, while
recognising that the Declaration on the Right to Development and the
International Bill of Human Rights constitute a milestone in the
construction of a universally accepted human rights architecture,
believes that these instruments need to be revisited for the following
reasons:

(a) The Definition of ``development'' is ambiguous and therefore needs
more clarity and precision;''

United Nations General Assembly, Fifty-third session, Human rights
questions, Right to development, Report of the Secretary-General. UN
Doc. A/53/268. 17$^{th}$ August 1998. Para 11

United Nations Commission on Human Rights, Fifty-fourth session,
Question of the realisation of the right to development, Report of the
Secretary-General in accordance with Commission resolution 1997/72. UN
Doc. E/CN1998/28. 16$^{th}$ February 1998. Para 8

Obiora Amede L,{\em  Beyond the Rhetoric of a Right to Development},
Law and Policy; V.18(3/4); pp.355-418  p 379

UN SUBCOM, {\em The legal nature of the right to development and
  enhancement of its binding nature}, 56$^{th}$ session, 01/06/2004,
UN Doc.E/CN.4/Sub.2/2004/16, para 40

United Nations Millennium Declaration, General Assembly Resolution
55/2 UN Doc. A/RES/55/2 of 8 September 2002 stated in paragraph 11
that  ``We will spare no effort to free our fellow men, women and
children from the abject and dehumanizing conditions of extreme
poverty, to which more than a billion of them are currently
subjected. We are committed to making the right to development a
reality for everyone and to freeing the entire human race from want.''
See further paragraphs 12-14 and 24 and 25 which link good governance
and TRTD, references are also made to international financial
institutions and cooperation.

World Conference against Racism, Racial Discrimination, Xenophobia and
Related Intolerance 2001, Report: ``the solemn commitment of all
States to promote universal respect for, and observance and protection
of, all human rights, economic, social, cultural, civil and political,
including the right to development, as a fundamental factor in the
prevention and elimination of racism, racial discrimination,
xenophobia and related intolerance.''

See General Assembly Resolutions on the Right to
development:Resolution 50/184 (1996) Right to development; Resolution
51/99 (1997) Right to development; Resolution 52/136 (1998) Right to
development; Resolution 53/155 (1999) Right to development; Resolution
54/175 (2000) The right to development; Resolution 55/108 (2001) The
right to development; Resolution 56/150 (2002) The right to
development; Resolution 58/172 (2003) The right to development

For example in supporting resolution 58/172,  The Syrian Arab Republic
``emphasised its support to the resolution entitled ``The Right to
Development'' because it believed that the right to development was as
inalienable human right, that the equality of opportunity for
development was a prerogative both of nations and of individuals and
the human person was the main benificiary and central subject of
development.'' United Nations General Assembly, Fifty-eighth session,
Human rights questions, The right to development, Report of the
Secretary-General. UN Doc. A/58/276. 12$^{th}$ August 2003. Paragraph
35

United Nations General Assembly, Fifty-eighth session, Human rights
questions, The right to development, Report of the
Secretary-General. UN Doc. A/58/276. 12$^{th}$ August 2003. Paragraph
11.

Japan stated ``The Government of Japan expressed its conviction that
the promotion and realisation of the right to development was relevant
in the context of development, but stressed its view that, at present,
the concept of the right to development was not yet clear or distinct
and further discussion was therefore necessary. It also expressed the
belief that the right to development should be ensured for individuals
in the territory of each country and that each Government should have
the primary responsibility for the protection and promotion of that
right. While international cooperation was important in the area of
development, the Government of Japan disagreed with the legal
obligation of developed countries to render assistance to developing
countries. United Nations General Assembly, Fifty-eighth session,
Human rights questions, The right to development, Report of the
Secretary-General. UN Doc. A/58/276. 12$^{th}$ August 2003. Paragraph
11.

Explanation of Vote (prior to vote) Item 7: Right to Development
Delivered by Joel Danies on April 25, 2003 available at
\uline{http://www.humanrights-usa.net/2003/statements/0425RtoD.htm}. Last
accessed 11.11.04

World Conference on Human Rights, Vienna, Vienna Declaration and
Programme of Action 1993

See Boven TV.,{\em  Human Rights and Rights of Peoples}, 6 EJIL (1995)
1-476

Explanation of Vote (prior to vote) Item 7: Right to Development
Delivered by Joel Danies on April 25, 2003 available at
\uline{http://www.humanrights-usa.net/2003/statements/0425RtoD.htm}.

Ibid

Ibid

This discussion in no way attempts to be exhaustive as, for the
purpose of this paper, it is deemed inappropriate to attempt to
discuss all the possible treaties and instruments which may contain
references to the principles contained in TRTD. Other instruments
which have been suggested to include indirect references of principles
of TRTD include; (International); Convention on the Elimination of All
Forms of Racial Discrimination 1965, See Articles 1, 2, 5(e);
Convention on the Elimination of All Forms of Discrimination Against
Women 1979, See Articles 1, 10-14; Convention on the Rights of the
Child 1989, See Articles 23, 24, 26-29, 31; International Convention
on the Protection of the Rights of All Migrant Workers and Members of
their Families 1990; (Regional) European Social Charter 1961; Revised
European Social Charter 1996; Additional Protocol to the European
Social Charter Providing for a System of Collective Complaints 1995;
American Declaration of Human Rights and Duties of Man 1948 Art VII,
IX, XI-XVI; American Convention on Human Rights 1969, See Article 26;
Additional Protocol to the American Convention on Human Rights in the
Area of Economic, social and cultural rights Protocol of San Salvador
1988; African Charter on the Rights and Welfare of the Child 1990, See
Articles 4, 11-16, 23; Draft Protocol to the African Charter on Human
and Peoples' Rights on the Rights of Women;  Arab Charter on Human
Rights 1994, See Articles 29-34, 36-39.

For an in depth discussion on the provisions which have been suggested
to relate to TRTD see Donnelly J., {\em In Search of the Unicorn : The
  Jurisprudence and Politics of the Right to Development}, 15 Cal W
Int'l L J 473, 1985; Sengupta A., {\em Theory and Practice on the
  Right to Development}, Human Rights Quarterly; V.24(4); pp.837-889

Cited in Donnelly J., {\em In Search of the Unicorn : The
  Jurisprudence and Politics of the Right to Development}, 15 Cal W
Int'l L J 473, 1985 p 483; UN Doc E/CN.4/1982/SR.3 at para 2 (China)

Ibid

The UDHR provisions which have been cited include Articles
22-28. Donnelly discusses the value of using Article 28 to establish
TRTD and his argument is impeccable. Furthermore, he explains how the
right to self-determination contained in both Covenants cannot be used
to imply TRTD. He also dismisses the use of the separate rights and
the right to an adequate standard of living as evidence of TRTD. See
Donnelly J., {\em In Search of the Unicorn : The Jurisprudence and
  Politics of the Right to Development}, 15 Cal W Int'l L J 473,
1985. However, for a contrasting view on the legal obligation for
international cooperation of ESC rights, see Hamm B, {\em A human
  rights approach to development}, Human Rights Quarterly; V.23(4);
pp.1005-10

\end{thebibliography}

\end{document}
